### 2020-06-30 Figures/article.tex
Overleaf version - no h1n1

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{color}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

% Not in template
\usepackage{amsmath}
\usepackage{lineno}
\linenumbers

\newcommand{\RAN}[1]{{\color{blue}Richard: #1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}

% \title{Seasonal influenza: the limits of predictability}
% \title{Limited predictability of seasonal influenza virus evolution}
\title{Limited predictability of clade frequencies of seasonal influenza viruses}

%\date{September 9, 1985}	% Here you can change the date presented in the paper title
\date{} 					% Or removing it

\author{Pierre Barrat-Charlaix\\
	Biozentrum\\
	Univeristät Basel\\
	Basel, Switzerland \\
	\texttt{pierre.barrat@unibas.ch} \\
	% %% examples of more authors
	\And
	\hspace{1mm}Richard Neher \\
	Biozentrum\\
	Univeristät Basel\\
	Basel, Switzerland \\
	\texttt{richard.neher@unibas.ch} \\
}

\begin{document}
\maketitle

\begin{abstract}
    \RAN{the abstract needs streamlining -- I can take a crack once the ms is stable}
	Seasonal influenza viruses causes a major epidemic every year in the human population, while mutating rapidly to evade host immune systems. Predicting which mutations will fix in its genome is of course critical to developing a vaccine, but also challenges our ability to understand the underlying evolutionary dynamics driving them. In this work, we use sequences of the influenza hemagglutinin (HA) from the last 20 years to assess our ability to predict the outcome of mutations in the virus\'s genome. Investigating the frequency trajectories of new amino acid mutations and their probability to fix in the flu population, we failed to find signatures of selection. On the contrary, our results differ little from a scenario of neutral evolution, implying strong limits to predictability. Interestingly, looking at known antigenic positions in HA or using proxies for fitness such as the \emph{Local Branching Index} (LBI) changes very little to these results, whereas the geographical spread or the age of a mutation are found to be informative of its probability to fix in the population. Finally, we try to find reasons for the predictive success of quantities such as the LBI. In doing so, we show the simple method of taking the consensus sequence of the current flu population fares slightly but consistently better than the LBI. Overall, our results indicate that the evolution of HA sequences happens in an unintuitive manner and does not agree with predictions of usual models such as travelling fitness waves.  
\end{abstract}

\section*{Introduction} % (fold)
\label{sec:introduction}

	Seasonal influenza A viruses (IAV) infect on the order of 10\% of the global population every year, resulting in hundreds of thousands of deaths~\cite{WHOfactsheet,petrova_evolution_2017}. Vaccination is primary measure to reduce influenza morbidity and the World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. However, the long time needed to develop and produce vaccines combined with the rapid antigenic evolution of influenza strains makes keeping the vaccine up to date a challenging task. Indeed, the surface proteins hemagglutinin (HA) and neuraminidase (NA) continuously accumulate mutations at a fast rate, leading to frequent antigenic changes \cite{petrova_evolution_2017, Shih6283, koelle2009understanding}. While a vaccine targeting a particular antigenic subtype may be efficient for some time, antigenic drift will sooner or later render it obsolete. This makes our ability to forecast the evolution of influenza of essential interest to public health. \\
	Over the last 20 years, sequencing efforts and better sequence sharing possibilities have resulted in an exponential increase in the availability of high quality HA and NA sequences~\cite{bogner2006global, shu2017gisaid}. This has made possible the analysis of sequence evolution at a fine temporal scale, resulting in a rich field of viral evolution prediction models~\cite{morris2018predictive}. These models aim at predicting the future population of influenza viruses based on estimations of the current fitness of strains. Fitness has been estimated by a variety different approaches, all showing a degree of success. In \cite{luksza_predictive_2014}, a sequence fitness model capturing antigenic drift and protein stability is trained using epitope and non-epitope mutations. Another approach in \cite{Steinbrueck12123} relies on hemagglutination inhibition (HI) assays to determine possible antigenic shifts for clades in the genealogy of the HA protein. Finally, \cite{neher_predicting_2014} uses the dense branching of HA genealogy as a growth measure. \\
	The underlying assumption of all these methods is that some mutations in surface proteins are under a strong positive selection, especially those responsible for antigenic drift, and thus rapidly fix in the population. In this work, we use sequences from HA and NA proteins in H3N2 influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. The question we ask is whether the selective pressure for new antigenic properties confers, on average, a typical shape to frequency trajectories of amino acid mutations, and whether this could be used to predict their future. To our surprise, we find that the predictability of these trajectories is very limited. Indeed, many of their properties seem compatible with selective neutrality, even in epitope positions. This observation is not attributable solely to clonal interference and genetic linkage, as simple simulations including those effects fail to reproduce it. In coherence with these observations, we show that a simple predictor uninformed about fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. This suggests that although LBI has some predictive power, the reason for its success may not be related to it approximating fitness of strains.

% section introduction (end)

\section*{Results} % (fold)
\label{sec:results}

	The main underlying question asked in this work is the following: given a mutation $X$ in the genome of H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
	There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$? %In figure \ref{fig:panel2} (bottom part), this amounts to predicting the shape of the frequency trajectory curves after the date indicated by a vertical line using all past information. 
	This type of prediction will be referred to as \emph{short term}.  \\
	Another way to formulate this question is to ask whether $X$ will fix in the population, will disappear, or will stay as a polymorphism. Here, fixation is meant as reaching a dominant frequency in the population for a long enough period of time (see Methods). Since many mutations define a clade in the phylogenetic tree of influenza, this amounts to asking which of these clades will be the ancestor of the future flu population, a question of critical importance for vaccine design. It will be referred to in the following as a \emph{long term prediction}.

	% \begin{figure}
	% 	\centering
	% 	\includegraphics[width=0.9\textwidth]{Figures/Panel1.png}
	% 	\caption{}
	% 	\label{fig:panel1}
	% \end{figure}

	In this work, we use all publicly available amino-acid sequences of the HA and NA genes of H3N2 influenza since the year 2000. This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represents a snapshot of the H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\ref{fig:nseq_per_year} and \ref{fig:nseq_per_month} in SM for details).\\
	The central quantities that are derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences. If an amino acid $X_i$ is found at position $i$ at a frequency between 5\% and 95\% in the population of a given time bin $t$, then the population is considered polymorphic at position $i$ and at time $t$. This polymorphism is characterized by the frequency of $X_i$, $f_{X_i}(t)$, and also by frequencies of other amino acids at $i$. The series of values $f_{X_i}(t)$ for contiguous time bins constitutes the frequency trajectory of $X_i$. A trajectory is terminated if the corresponding frequency is measured above 95\% (resp. below 5\%) for two time bins in a row, in which case amino acid $X_i$ is considered as \emph{fixed} (resp. \emph{absent}) in the population. Otherwise, the trajectory is considered \emph{active}. Examples of trajectories can be seen in figure~\ref{fig:sample_trajectories} of the Supplement.\\% [NOTE: I also explain this in Methods. Maybe shorten this part]\\
	Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. \\
	In the rest of this work, we will focus on frequency trajectories that are starting at a null frequency, \emph{i.e.} $f(t=0)=0$. These represent new amino acid mutations which were absent in the past and are currently rising in the population. Two reasons motivate this. The first is simply that this avoids the problem of non-independent trajectories. Indeed, each rising trajectory implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. By considering new mutations only, we restrict ourselves to independent trajectories. \\
	The second is that we expect to see a stronger signal of selection for this type of mutations. Current research on influenza seems to indicate that surface proteins HA and NA are under strong selective pressure as they have to escape hosts immune systems. This escape is driven by adaptive mutations, for instance on epitopes of the mentioned proteins. For this reason, one could expect new amino acid mutations that have reached a high enough frequency to be on average \emph{positively} selected, and their frequency trajectories should reflect this by showing a regular increase until they take over the population. We will call such a change in the dominant amino acid at a position a \emph{sweep}, examples of which can be seen in some trajectories represented in supplementary figure~\ref{fig:sample_trajectories}. By restricting ourselves to trajectories that start at low frequency, we should bias the mutations we analyse towards positively selected ones, thus making them easier to predict as well as recovering a clearer signal of selection. \\

	Sequences are also used to build a phylogenetic tree of HA and NA genes. For computational reasons, the number of sequences used to infer the tree is limited to 100 per 4 months (randomly chosen), totalling 4402 sequences over the 2000-2019 time period. The inference is done using the \emph{iqtree} software with default settings~\cite{10.1093/molbev/msaa015,10.1093/molbev/msu300}. \\

	\subsection*{Short term prediction}

	Having observed the frequency trajectory $f(t)$ of a mutation until a given date $t_0$, how much can we say about the future values of $f$ after $t_0$? To make the question more specific, we consider the idealized case sketched on panel \textbf{A} of figure~\ref{fig:panel2}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f)$ of observing it at a value $f$ at time $t_0 + \Delta t$? \\
	To answer this question retrospectively, we use all frequency trajectories extracted from HA H3N2 sequences that satisfy these conditions for a given $f_0$. Since actual frequency trajectories rely on a finite number of sequences, we cannot exactly measure them at a precise frequency $f_0$. For this reason, we allow some leeway around $f_0$, considering all trajectories measured in the interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$. \\
	For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:panel2}, where time is shifted such that $t_0 =0$. It can be noted that some of those trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. This again comes from the fact that frequencies are based on a finite number of sequences, resulting in some trajectories being measured below the selected frequency bin at a given time bin and above it at the next time bin. These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. Removing them does not change results significantly. \\
	Since influenza is often thought of as permanently evading the human immune system, one could expect that a significant fraction of new amino acid mutations that have reached a finite frequency represented by rising trajectories on figure~\ref{fig:panel2} are \emph{adaptive} mutations under positive selection. We could thus expect that most of these trajectories continue rising after reaching frequency $f_0$ in an "inertial" manner, at least for some short enough amount of time. A fraction of those would then sweep through the population and smoothly reach a dominant frequency or fix. \\
	Trajectories collected and shown in figure~\ref{fig:panel2} are used to verify if past trajectories did behave in this inertial manner, and to measure the corresponding "persistence time", that is the amount of time during which the trajectory will keep rising. In practice, we  estimate the probability distribution $P_{\Delta t}(f)$ of finding a trajectory at frequency $f$ after a time $\Delta t$. Results are shown on panel \textbf{C} of figure~\ref{fig:panel2}. Initially, \emph{i.e.} at time $t_0=0$, this distribution is by construction peaked around $f_0$. If a large fraction of the trajectories keep increasing after this time, we should see the "mass" of $P_{\Delta t}(f)$ move to the right towards higher frequencies as time progresses.\\
	 However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. Indeed, our results show that the frequency distribution seems to  flatten more or less symmetrically in the first 30 to 60 days. After longer times of a year to two years, if we exclude mutations that have disappeared or fixed, this distribution becomes almost flat (leftmost and rightmost bins of the distribution). This indicates that what we called the persistence time of frequency trajectories is rather small, of the order of a month at most. After a few months, knowledge that the trajectory was rising and arriving at frequency $f_0$ is mostly lost. This contradicts the view that influenza's evolution is dominated by adaptive sweeps which take over the population and suggests a strong limit to our ability to predict the future of frequency trajectories.

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
		\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we "catch" at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f)$? \textbf{B}: All frequency trajectories of amino acid mutations in the HA gene that were absent in the past, are seen around 30\% frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. The data is shown by the circles, the lines are a smooth fit to the distribution and should be seen as a guide for the eye. }
		\label{fig:panel2}
	\end{figure}

	\subsection*{Long term prediction}

	The future values of a frequency trajectory $f(t)$ do not seem to be easily predictable. However, another way to predict evolution is to focus not on dynamical details of the frequency of a mutation but rather on the probability that a mutation fixes in the population. For influenza, this question is of particular importance as the knowledge of the future H3N2 population and of the mutations it acquired is critical to the development of a vaccine. Here, we use the past frequency trajectories to assess our ability to predict fixation of mutations in the population. Since most mutations define a unique clade in the phylogenetic tree of HA or NA genes, this amounts to predicting which clade will be the ancestor of the future influenza population. \\
	We first estimate the number of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. For this, we consider the 106 new amino acid mutations in HA that are seen at least once above a frequency of 25\%. Panel \textbf{A} of figure~\ref{fig:panel3} shows the fraction of frequency trajectories corresponding to these mutations that either fix, are lost or remain active as a function of the time elapsed since they were first seen above 25\% frequency. It is clear from this plot that most mutations are either lost or become fixed in a few years, with no trajectory remaining active after 6 years. It is relevant to note that this value of 6 years is also the typical coalescence time observed in phylogenetic trees of H3N2 influenza. We also note that the fraction of lost trajectories increases sharply at small times, indicating that many mutations observed at a significant frequency are lost relatively quickly, while it takes longer to fix a mutation in the whole population. \\

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
		\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency. \textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population.}
		\label{fig:panel3}
	\end{figure}

	We then examine the probability of mutations to fix in the population as a function of the frequency at which they are seen. For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 5\%, f + 5\%]$ at any given time. The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold. \\
	Results are shown for proteins HA and NA in the panel \textbf{B} of figure~\ref{fig:panel3} by plotting $P_{fix}(f)$ as a function of $f$. We notice that for both these proteins, the probability of fixation of a new mutation seen at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$. This result is quite surprising, as this is exactly what is expected from a population evolving in the \emph{absence} of selection. Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{10.1080/10635150500354860}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. Panel \textbf{B} of figure~\ref{fig:panel3} indicates that mutations in the surface proteins of H3N2 influenza are in good agreement with the neutral expectation.\\
	This is surprising, since influenza is usually described as being under strong selective pressure to evade its host's immune systems. If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$. \\
	To investigate further, we try to find groups of mutations that deviate from the neutral expectation of figure~\ref{fig:panel3}. We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and is arguably a good proxy for the fitness of a leaf (\emph{i.e.} a strain) or an ancestral node in a tree. Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. This effectively represents the average density of the tree around nodes that carry the mutation. We then reproduce the plot of figure~\ref{fig:panel3}, but this time dividing trajectories into two groups: those corresponding to mutations with higher LBI, meaning that their LBI is higher than the median value recorded for mutations present in the same frequency bin, and those with lower LBI. Panel \textbf{A} of figure~\ref{fig:panel4} shows this plot for HA. Because both groups are again consistent with neutral evolution, we see that LBI carries very little information on the probability of fixation of a mutation as both subgroups are again similar to the neutral expectation. \\
	Next, we concentrate on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. These sites are considered responsible for the ability of H3N2 to evade immune systems, and mutations at these locations should therefore be under strong selection. Four lists of relevant epitope positions from different authors are used, comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. Panel \textbf{B} of figure ~\ref{fig:panel4} shows the $P_{fix}(f)$ as a function of $f$ curve for the four lists of epitopes. Results show that mutations on epitope sites as reported in~\cite{Koel976, luksza_predictive_2014} have slightly higher chances of fixation, while no clear difference is found in the case of~\cite{Shih6283, wolf_long_2006}. However, even in the former case, observations are different from what one would expect if the dynamics where dominated by selective sweeps. Several other ways of looking at subset of mutations or positions reach the same result: fixation probability hardly deviates from the neutral expectation (see figure \ref{fig:pfix_v_freq_misc} in SM). [I should maybe add something saying that those epitope positions might have been found \emph{using} the fact that sweeps often take place there? Is that the case? depends on authors most likely. EDIT: Wolf et al. seems to report all sites that are somehow important to the immune system. The others have a selection of those. For Koel and Luksza, it seems to be based on the fact there have been sweeps at these positions (in an indirect manner). Should read papers more carefully. Also, Koel et al. has 7 epitope positions, so error bars on the $y$-axis must be huge. And finally: taking RBS positions actually increases $P_{fix}(f)$ significantly (Koel et al. takes positions close to the RBS).]\\
	\pierre{Write something on panels \textbf{C} and \textbf{D}, \emph{i.e.} binary mutations and mutations found once or more in the tree.}

	Motivated by the fact that the number of influenza cases varies strongly with the seasons in non-tropical regions, we turn to ways of classifying mutations that are not based on genomes and phylogeny but rather on "epidemiological" observations. First, a score reflecting \emph{geographical spread} is assigned to each mutation. It is based on the entropy the regional distribution of each mutation (details in Methods). We also use the age of trajectories as a second score, that is, for each frequency $f$, the amount of time between the date at which the corresponding mutation is first seen, and the date at which it reaches frequency $f$. \\
	Fixation probabilities as a function of observed frequency with mutations classified by higher or lower values of these scores (\emph{i.e.} higher or lower than the median of observed values) are shown in panels \textbf{A} and \textbf{B} of supplementary figure~\ref{fig:geospread_and_time}. These two scores also allow a quantitatively moderate distinction between mutations: for a given frequency $f$, mutations found in many regions or that are older (in the sense that they have taken more time to reach frequency $f$) tend to fix more often than geographically localized mutations or more recent ones. These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\ref{fig:geospread_v_time} of SM. \\
	A natural way to explain these observations would be to refer to the seasonal characteristics of H3N2. In temperate regions, the number of H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. For this reason, one could imagine that some mutations rise in frequency in an "opportunistic" way during an epidemic season, but do not survive the following decline in population size. For the same reason, mutations that are localized in a particular region might mostly be due to the epidemic season specific to that region. \\
	However, it is important to note that verifying this kind of hypothesis remains challenging with the available data. Indeed, supplementary figures~\ref{fig:nseq_per_year} and ~\ref{fig:nseq_per_month} clearly shows that sequence are not uniformly sampled in either in time or space: some countries or regions sequence much more than others, but also do so at a varying intensity in time (see for instance Oceania samples in figure~\ref{fig:nseq_per_year}). Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non trivial to measure. It is however interesting to observe that quantities based on the population's geographical and temporal structure can apparently give as much or more information on fixation probability and predictability than genomic or phylogenetic ones.

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
		\caption{Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
		\label{fig:panel4}
	\end{figure}

	\subsection*{A simple model}
	\label{sub:a_simple_model}

	Quick summary. The results shown in figures~\ref{fig:panel3} and \ref{fig:panel4} are surprising. One could argue that the apparently neutral behaviour of mutations is due to strong genetic linkage inside each protein and high competition between different adaptive mutations. We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time. 
	Two different fitness landscapes are used, described below. \\
	First, we use an additive fitness function, with sequence $(x_1\ldots x_L)$ having a fitness $h_1 x_1 + \ldots + h_L x_L$. This implies that for a given genome position $i$, the trait $x_i = 1$ is favored if $h_i>0$ whereas  $x_i = -1$ is favored if $h_i<0$. All $h_i$'s have the same amplitude, and only their signs matter. Every $\Delta t$ generations, we randomly choose a position $i$ and flip the sign of $h_i$, effectively changing the fitness landscape. Individuals in the population now have the opportunity to make an adaptive mutation at site $i$ giving them a fitness advantage $2\vert h \vert$. However, because of the absence of recombination, a genome that has adaptively mutated at $i$ might also carry non-optimal mutations referred to as mutational load. Each "flip" in the fitness landscape will thus create a competition between genomes carrying the adaptive mutation as well as their own mutational load. \\
	To increase competition between genomes, we design a second model that includes epistasis. Once again, the baseline fitness of a genome is an additive function, this time with values of $h_i$ that do not change through time. Every $\Delta t$ generation, we now introduce "antibodies" that target a specific sub-sequence of length $l=5$, noted $(x^{ab}_{i_1}, \ldots, x^{ab}_{i_l})$. The positions $(i_1\ldots i_l)$ are chosen at random, while the value $x^{ab}_i$ is chosen to be the dominant trait at position $i$. Genomes that include the \emph{exact} sub-sequence targeted by the antibody suffer a strong fitness penalty. However, a single mutation away from that sub-sequence removes this penalty, resulting in a fitness landscape with very strong epistasis. This has the effect of triggering a strong competition between adaptive mutations: for a given antibody, $l=5$ possible mutations are now adaptive, but combinations of these mutations do not bring any fitness advantage. \\
	Having simulated populations in these two fitness landscapes, we perform the same analysis of frequency trajectories as for the real influenza data. Figure~\ref{fig:simulations} of the SM shows the $P_{fix}(f)$ as a function of $f$ curve for the two models and for different values of the inverse rate of change $\Delta t$ of the fitness landscape. For all models, this curve deviates significantly from the diagonal. This is most evident in the case without antibodies and with a very long rhythm of change $\Delta t=1000$. In this scenario, rising mutations almost always fix in the population, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent. This is corroborated by visual inspection of the trajectories, which shows that evolution in this regime is driven by regular selective sweeps that take a typical time of $\sim 400$ generations. In other regimes, with smaller $\Delta t$ or with the introduction of antibodies, $P_{fix}(f)$ typically becomes smaller and closer to the diagonal. However, even in an extremely fast changing fitness landscape with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ still significantly differs from $f$. \\
	These models do not claim to be in any way realistic, and even less to be a good representation of the evolution of influenza or viruses. What we learn from them and from figure~\ref{fig:simulations} is that it is not a trivial task to design a model which includes selection and yet has a behaviour that is in apparent good agreement with a neutral evolution scenario, such as what figure~\ref{fig:panel3} shows. This makes the results of the previous section all the more interesting, as influenza is generally regarded as being under strong selection in a changing fitness landscape. In other words, the results of figure~\ref{fig:panel3} may not be be regarded as obvious manifestations of genetic linkage and clonal interference.

	\subsection*{Why do predictions work}

	Results that we obtained on the frequency trajectories of mutations are in good agreement with what is expected of a neutrally evolving population. What is even more surprising is that the Local Branching Index, a quantity that has previously been thought of as a good approximation for the fitness of strains, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:panel4}. However, LBI has also been demonstrated to be a good predictor of the  future influenza population. This raises an obvious question: why does predicting the evolution of influenza work at all? \\
	We investigate this using a subsample of the HA sequences used in the previous sections. We decide to work with larger non-overlapping time bins of 4 months. For each such time bin from year 2003 to 2019, we randomly sample 100 HA sequences, reaching a total of 4402 (some 4 month intervals contain less than 100 sequences in total). A tree is also inferred using those sequences, and the LBI of each node in that tree is computed.\\
	Each time bin is considered as a snapshot of the H3N2 influenza population for a given date. Predicting the evolution of the influenza population consists of obtaining as much information as possible on the sequences contained in a future time bin $t+\Delta t$, using only sequences in time bin $t$ and before. Sequences in $t$ are referred to as the \emph{present} population, and sequences in $t+\Delta t$ as the future population. We attempt to measure the efficiency of two predictors consisting of a single sequence: the sequence of the strain with the highest LBI, and the amino acid consensus sequence of members of the present population (see Methods for a definition of the consensus sequence). The choice of the consensus sequence as a predictor is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population if Hamming distance is used as a measure (see SM). However, it is important to note that this sequence does \emph{not} necessarily exist in the population. Our procedure is then the following: for a given 4 month time bin $t$, that is for a given "present" population, we compute the \emph{average Hamming distance} between the sequence of the predictor strain and the sequences of future populations at times $t+\Delta t$, for $\Delta t$ comprised between $0$ and $32$ months. Results are then averaged over all  possible values of $t$, giving us an average efficiency of a predictor over 16 years of influenza evolution. \\
	The top panel in figure~\ref{fig:panel5} shows the results of this procedure, also comparing them to a predictor based on a randomly chosen strain. It is very clear that both predictors, the LBI-based one and the consensus, are consistently and significantly more accurate than a random pick in the present population. This can be taken as both a positive sign and a confirmation of previous results: it is to some extent possible to predict the evolution of influenza in the sense that one can do better than a purely random guess. However, what is striking is the high similarity between the predictions of the top-LBI strain and the consensus sequence. Moreover, the consensus sequence fares slightly but consistently better than the LBI, even if only slightly. This is surprising in that the consensus sequence is the best possible long-term possible predictor for a neutrally evolving population, whereas the LBI was designed as an approximate measure of fitness of strains. \\
	An explanation for this observation is that the LBI tends to be high for nodes in a tree that are close to the root of a dense and large clade. If we think of the influenza population as divided into a few different clades, and if the strains with the maximal LBI are close to the root of the largest of those clades, then it may also be that their sequence is close to the consensus of the whole population. To test that hypothesis, we measure the hamming distance from the sequence of the top LBI strain to the consensus sequence for populations of all time bins. Panel \textbf{B} of figure~\ref{fig:panel5} shows these distances, scaled with respect to an average strain (details in caption). It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only once is the sequence of the top-LBI strain farther away from the consensus than the average sequence is. Moreover, the sequence of the top-LBI strain \emph{exactly} matches the consensus in 23 cases. \\

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{./Figures/Panel5.png}
		\caption{\textbf{A}: Average Hamming distance of the sequences of different predictors to HA sequences of future influenza populations, themselves averaged over all "present" populations from year 2003 to 2019. Predictors are: a randomly picked sequence in the present population; the sequence of the strain with the highest LBI in the present population; the consensus sequence of the present population. \textbf{B}: Scaled Hamming distance between the sequence of the top LBI strain and the consensus sequence for populations at different dates. The scaling is such that for each date, the Hamming distance between a strain from the population and the consensus is on average $1$. The strain with the highest LBI is almost always closer to the consensus sequence than the average strain.}
		\label{fig:panel5}
	\end{figure}

% section results (end)

\section*{Discussion} % (fold)
\label{sec:discussion}

The idea of predicting the outcome of a mutation is based on two assumptions: significant fitness effects between genomes carrying different mutations and a stable enough environment. These two conditions are essential for the existence of a consistent selective pressure on mutations, which is expected to give inertia to frequency trajectories of mutations, thus making them predictible. We simulated a population having such a selective pressure in  section \ref{sec:a_simple_model} and found indeed that as long as the fitness landscape does not change too rapidly (\emph{i.e.} small $dt$ limit), frequency trajectories are predictible to some extent. \\
Methods for predicting the future evolution of influenza usually rely on modeling how this selection operates and on finding the values of fitness for strains or mutations. In this work, we tries to understand whether the selective pressure acting on influenza is strong and consistent enough to make frequency trajectories of mutations follow "inertial" patterns such as selective sweeps. By analyzing 19 years of evolution of the HA and NA genes in h3n2 influenza, we find that these trajectories do not behave in an easily predictible way. On the short time scale, not much can be said about the future frequency values or even the direction of a rising trajectory. On the longer time scale, we find that the probability of an amino acid mutation to fix is in very good agreement with what is expected of a neutrally evolving population. \\
Usual prediction tools add surprisingly little information to this. On the one hand, Mutations at some epitope positions are slightly more likely to fix than ones at other positions, but their probabilty of fixation remains very close to the neutral expectation. On the other, the Local Branching Index seems to carry no information at all on probability of fixation. It is interesting to note that other simple quantities also give an indication on probability of fixation, such as the number of times the mutation appears in the tree, the restriction to positions that never see three amino acids compete at the same time, or the age and geographical spread of the mutation (SM figure \ref{fig:geospread_and_time} for the latter). \\
These results naturally trigger the question of the reasons why prediction methods have been found to work. We focused on the surprising case of LBI. In the last section, we showed that the strain with the highest LBI in a given present population is a better predictor of the future population than a randomly picked one. However, we also find that the very simple method of taking the consensus sequence of all present strains performs slightly but consistently better than the LBI. Interestingly, the consensus sequence is the best possible predictor for a neutrally evolving population, and does not model fitness in any way. Thus, even though the LBI claims to approximate the fitness of strains based on their genealogy, it may be a good predictor for different reasons. \\
Overall, these results indicate that the potential predictibilty of the evolution of influenza is fairly low, challenging the views that this evolution operates under a strong and consistent selective pressure. On the contrary, we fail to find evident signs of a such a selective pressure and show that the LBI, designed to measure fitness, may not work for the "right" reasons.

Quick list of what should be said: 
\begin{itemize}
	\item Flu is usually seen as evolving under strong selection. Predicting its evolution should thus rely on finding how this selection operates, and finding which strains are the most fit according to it. 
	\item Analysing 19 years of flu evolution, we find that frequency trajectories of mutations in the HA and NA proteins do not behave in an obviously predictable way. More precisely, the "inertia" or "persistence time" of these trajectories is at best quite short, and seeing a mutation rising in frequency does not tell us much on its future. 
	\item We also find that the probability of a specific mutation to fix in the flu population is in very good agreement from what is expected of a neutrally evolving population. Usual predictive tools such as the LBI or the epitope positions on the HA protein do not add much information to the probability of fixation. However, we find evidence that population geographical and temporal structure does give a limited amount of information. 
	\item Using a simulation of a constant size population evolving under selection, we demonstrate that the previous findings are not easily reproduced by usual views of population dynamics. Simple but reasonable models do not behave in an apparently neutral manner
	\item Surprised by these observations, we wonder whether predictive tools such as the LBI work at all. By using hamming distance to future populations as a measure, we show that the strain with the highest LBI is indeed a good prediction of future sequences in the sense that it is better than a random pick. However, we also find that the consensus sequence fares consistently better than the LBI. This is in perfect agreement with previous results, since the consensus sequence is the best possible predictor for a neutrally evolving population. 
	\item Overall, this work leaves us with more questions than we started with. It shows that the usual views of evolution of influenza may be quite inaccurate and potentially misleading. Indeed, selection may be too small to be easily detected by the methods we used here, or operates in a way that we do not understand. It also suggests that population structure (geographical and temporal) should not be discarded as it may play an important role in this evolution. 
\end{itemize}

% section discussion (end)

\section*{Methods} % (fold)
\label{sec:methods}

\subsection*{Data} % (fold)
\label{sub:data}

% subsection data (end)

\subsection*{Frequency trajectories} % (fold)
\label{sub:frequency_trajectories}

	For a set of sequences in a given time bin, we compute frequencies of amino acids at each position by simple counting. We make the choice of not applying any smoothing method in an attempt to be as close to the data and "model-less" as possible. This is especially important for the short term prediction of frequency trajectories, as estimations of the "persistence time" of a trajectory might be biased by a smoothing method. \\
	We compute frequency trajectories based on the frequencies of amino acids. A trajectory begins at time $t$ if an amino acid is seen under the lower frequency threshold of 5\% (resp. above the higher threshold of 95\%) for the two time bins preceding $t$, and above this lower threshold (resp. below the higher threshold) for time bin $t$. It ends in the reciprocal situation, that is when the frequency is measured below the lower threshold (resp. above the higher threshold) for two time bins in a row. \\
	In order to avoid estimates of frequencies that are too noisy, we only keep trajectories that are based on a population of at least 10 sequences for \emph{each} time bin. As said in the Results section, we also restrict the analysis to trajectories that begin at a $0.$ frequency, in part to avoid double counting. We find a total of 460 such trajectories. However, only 106 reach a frequency of 20\%, on which figure~\ref{fig:panel3} is based for instance.

% subsection frequency_trajectories (end)

\subsection*{Local Branching Index} % (fold)
\label{sub:local_branching_index}

	LBI was introduced in \cite{neher_predicting_2014} as an approximation of fitness in populations evolving under persistent selective pressure that is fully based on a phylogenetic tree. It relies on the intuition that the tree below high-fitness individuals will show dense branching events, whereas absence of branching is a sign of low-fitness individuals. Quantitatively, the LBI $\lambda_i(\tau)$ of a node $i$ is the integral of all of the tree's branch length around $i$, with an exponentially decreasing weight $e^{-t/\tau}$ with $t$ being the branch length. $\tau$ is the time scale for which the tree is informative of the fitness of a particular node. Here, we use a value of $\tau$ equal to $T_C\simeq6$ years, the coalescence time for influenza H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA). We have observed that given our method to predict the future from present populations corresponding to time bins of 4 months, changing the value of $\tau$ has little effect on the pick of the top LBI strain. By retrospectively optimizing its value, it is possible to reduce the average distance to the population 2 years ahead by $\sim0.25$ amino acids on average, making the LBI method almost as good as the consensus on figure~\ref{fig:panel5}.

% subsection local_branching_index (end)

\subsection*{Measuring the geographical spread of a mutation} % (fold)
\label{sub:measuring_the_geographical_spread_of_a_mutation}
	For a mutation $X$ we define its regional distribution using the numbers $n_r(X)$ that represent the number of sequences sampled in region $r$ that carry $X$. Regional weights are then defined as 
	$$ w_r(X) = \frac{n_r(X)}{\sum_{r}n_r(X)}.$$
	We can then measure the geographical spread $G(X)$ of $X$ by using the Shannon entropy of the probability distribution $w_r(X)$: 
	$$G(X) = \sum_r w_r(X)\log(w_r(X)).$$
	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
	Region used are the ones defined in the \texttt{nextstrain} tool~\cite{10.1093/bioinformatics/bty407}. Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
% subsection measuring_the_geographical_spread_of_a_mutation (end)

\subsection*{Consensus sequence} % (fold)
\label{sub:consensus_sequence}
	Given a set of $N$ sequences $(\sigma^1,\ldots,\sigma^N)$ based on an alphabet $\mathcal{A}$ (\emph{e.g.} $\mathcal{A}$ has 20 elements for amino acids, 4 for nucleotides), we can define a \emph{profile} distribution $p_i(a)$ by the following expression:
	$$ p_i(a) = \sum_{n=1}^N \delta_{\sigma^n_i,a} $$ 
	where $i$ is a position in the sequence, $\sigma^n_i$ the character appearing at position $i$ in sequence $\sigma^n$, $a$ a character of the alphabet and $\delta$ the Kronecker delta. The profile $p_i(a)$ simply represents the fraction of sequences which have character $a$ at position $i$. \\
	We then simply define the consensus sequence $\sigma^{cons}$ such that 
	$$\sigma^{cons}_i = \text{argmax}_a \;p_i(a).$$
	In other words, the consensus sequence is the one that has the dominant character of the initial set of sequences at each position.   
% subsection consensus_sequence (end)

\subsection*{Earth Mover's Distance} % (fold)
\label{sub:earth_mover_s_distance}

	In order to measure the distance of several predictor sequences to the future population, we rely on the \emph{EarthMover's Distance} (EMD), a metric commonly applied in machinelearning to compare collections of pixels or words \cite{710701,10.5555/3045118.3045221}. Here, we apply it to compute the distance between the sequences of two populations, noted as $\mathcal{X} = \left\{(x^n, p^n)\right\}$ and $\mathcal{Y}=\left\{(y^m,q^m)\right\}$ with $n\in\{1\ldots N\}$ and $m\in\{1\ldots M\}$. In this notation, $x^n$ and $y^m$ are sequences, and $p^n$ and $q^m$ are the frequencies at which these sequences are found in their respective populations. For convenience, we also define $d_{mn} = H(x^n,y^m)$ as the Hamming distance between pairs of sequences in the two populations. \\
	We now introduce the following functional
	$$ F(\mathbf{w}) = \sum_{n,m} d_{nm}w_{nm}, $$
	with $\mathbf{w} = \{w_{nm}\}$ being a matrix of positive weights. The EMD between the two populations $\mathcal{X}$ and $\mathcal{Y}$ is now defined as the minimum value of function $F$ under the conditions 
	\begin{equation*}
		% \begin{split}
			\sum_{n=1}^N w_{nm} = q^m,\;\; \sum_{m=1}^M w_{nm} = p^n, \,\text{and}\;\; w_{nm}\geq 0
		% \end{split}
	\end{equation*}
	Intuitively, the weight $w_{nm}$ tells us how much of sequence $x^n$ is "moved" to sequence $y^m$. The functional $F$ sums all of these moves and attributes them a cost equal to the Hamming distance $d_{nm}$. The conditions on weights in $\mathbf{w}$ ensure that all the weight $p^n$ of $x^n$ is "moved" to elements in $\mathcal{Y}$ and vice versa. \\
	The minimization is easily performed by standard linear optimization libraries. Here, we use the Julia library {JuMP}~\cite{DunningHuchetteLubin2017}.
% subsection earth_mover_s_distance (end)

% section methods (end)

\bibliographystyle{unsrt}
\bibliography{references}

\newpage

\section*{Supplementary material} % (fold)
\label{sec:supplementary_material}

\subsection{Consensus sequence as a predictor for neutrally evolving populations} % (fold)
\label{sub:consensus_sequence_as_a_predictor}

	We consider the case of a neutrally evolving and structure-less population, such as the one in the Wright-Fisher model of evolution~\cite{10.1080/10635150500354860}. At an initial time $t=0$, the population consists of $N$ individuals with genomes $(\sigma^1\ldots\sigma^N)$ of length $L$ (not necessarily distinct). \\
	% We introduce the distribution $f(\sigma)$ representing the frequency at which a particular genome $\sigma$ appears in the population, that is 
	% $$ f(\sigma) = \frac{1}{N}\sum_{n=1}^N \delta_{\sigma, \sigma^n}$$
	% where $\delta_{x,y}$ is the Kronecker delta. \\
	We make two hypotheses about this population. We first suppose that \emph{no} mutations occur during the evolution of this population. This may seem surprising and is of course not true in the case of influenza. This assumption is however in line with the fact that the object of this work is to predict the outcome of \emph{already existing} mutations in the influenza population. The prediction of mutations that we have not yet seen is not in its scope. Thus, assuming that no new mutations take place can be seen as a simple way to model the fact that we have no information about such events. \\
	The second assumption is that the population evolves in a completely neutral way, meaning that the average number of descendants of each genome $\sigma^n$ is the same. Let us now consider the population after it has evolved for a long time $t\gg T$ where $T$ is the typical coalescence time (for the Wright-Fisher model, $T=2N$). At this point, all individuals in the future population will descend from a unique individual $n_0$ in the $t=0$ population. Our two hypotheses now allow us to make two statements. First, since no new mutations are allowed, the population at $t\gg T$ will be clonal, with all individuals having genome $\sigma^{n_0}$. Second, since the evolution is neutral and does not favour any genome in particular, the probability that $\sigma^{n_0}$ is equal to a given genome $\sigma$ is $1/N$. In other words, the probability that a genome at $t=0$ ultimately becomes the ancestor of all the future population is equal to its frequency in the $t=0$ population. \\

	We now try to find the genome $\sigma$ that best predicts the future population on the long run, that is for $t\gg T$. Here, we take best to mean that the predictor minimizes $H(\sigma,\sigma^{n_0})$ where $H$ is the Hamming distance defined by
	\begin{equation}
	 	H(\sigma^a, \sigma^b) = \sum_{i=1}^L (1 - \delta_{\sigma^a_i, \sigma^b_i}),
	 	\label{eq:hdist}
	\end{equation} 
	with $\sigma_i$ being the character appearing at position $i$ of genome $\sigma$ and $\delta$ the Kronecker delta. Since we do not know $n_0$, we have to average over all its possible values. $\sigma$ must thus minimize the following quantity: 
	\begin{equation}
		\begin{split}
			\langle H(\sigma, \sigma^{n_0})\rangle_{n_0} &= \sum_{n=1}^N H(\sigma, \sigma^n)\\ 
			&= \sum_{i=1}^L\sum_{n=1}^N (1 - \delta_{\sigma_i,\sigma^n_i})
		\end{split}
	\label{eq:2}
	\end{equation}
	by using the definition of the Hamming distance. We now assume that characters at each positions of the genomes can be indexed by an integer $a$ running from $1$ to $q$. For instance, if these were amino acid sequences, we could index the $20$ amino acids by $a$ running from $1$ to $q=20$. We rewrite the Kronecker delta in the previous expression using this indexation:
	$$ \delta_{\sigma_i,\sigma_i^n} = \sum_{a=1}^q \delta_{\sigma_i,a}\delta_{\sigma_i^n,a}.$$
	We also introduce the \emph{profile} frequencies $p_i(a)$ of the population at time $t=0$: 
	\begin{equation}
		p_i(a) = \sum_{n=1}^N \delta_{\sigma^n_i, a}.
		\label{eq:profile}	
	\end{equation}
	$p_i(a)$ represents the frequency at which character $a$ appears at position $i$ in genomes of the initial population. \\
	Equation~\ref{eq:2} now becomes
	\begin{equation}
		\begin{split}
			\langle H(\sigma, \sigma^{n_0})\rangle_{n_0} &= \sum_{i=1}^L\sum_{n=1}^N\left(1 - \sum_{a=1}^q \delta_{\sigma_i,a}\delta_{\sigma_i^n,a}\right)\\
			&= \sum_{i=1}^N \left(1- \sum_{a=1}^q\delta_{\sigma_i,a}p_i(a) \right) \\ 
			&= \sum_{i=1}^L \left(1-p_i(\sigma_i)\right)
		\end{split}
		\label{eq:3}
	\end{equation}
	This means that the genome $\sigma = (\sigma_1\ldots\sigma_L)$ which best predicts the future population according to our definition is the one that minimizes the quantity $(1-p_i(\sigma_i))$ for all positions $i$. This obviously implies that each $\sigma_i$ must be chosen as to maximize $p_i(a)$, that is $\sigma_i$ must be the character that appears the most frequently at position $i$. Thus, $\sigma$ must be the \emph{consensus} sequence of the initial population.

% subsection consensus_sequence_as_a_predictor (end)

\subsection{Predictor based on the local LBI maxima} % (fold)
\label{sub:predictor_based_on_the_local_lbi_maxima}
	In figure~\ref{fig:emd_to_future}, we use several sequences as a predictor of the future population. Distance between two sets of sequences, \emph{i.e.} the predictor sequences and the ones of the future population, is defined as the Earth Mover's Distance (EMD). Here, we show that for a population evolving under the same hypotheses as in section~\ref{sub:consensus_sequence_as_a_predictor}, the best \emph{multiple} sequence long term predictor is again the consensus sequence with weight $1$. \\
	Let the predictor be a set of weighted sequences $\{(s^\alpha, q_\alpha)\}$. We again use the fact that in the long term, a unique sequence $\sigma^{n_0}$ from the present will be the ancestor of the entire population. We want to compute the EMD from the predictor to $\sigma^{n_0}$, that is the EMD between the sets $\mathcal{X}=\{(s^\alpha, q_\alpha)\}$ and $\mathcal{Y} = \{\sigma^{n_0},1\}$. Applying the definition of the Methods section, it follows that the weights $\mathbf{w}$ are in this case equal to the $q_\alpha$s. By averaging over all values of $n_0$, we now obtain
	$$ \left\langle \,\text{EMD}\left(\{(s^\alpha, q_\alpha)\}\right) \right\rangle_{n_0} = \sum_{n=1}^N\sum_{\alpha} H(s^\alpha, \sigma^n)\cdot q_\alpha.$$
	By the same calculation procedure as in the previous section, this expression simplifies to 
	$$ \left\langle \,\text{EMD}\left(\{(s^\alpha, q_\alpha)\}\right)\right\rangle_{n_0} = \sum_{i=1}^L\left( 1 - \sum_{a=1}^q p_i(a)q_i(a) \right), $$
	where the profile of the present population $p_i(a)$ has already been defined, and $q_i(a)$ stands for the profile of the predictor, that is 
	$$ q_i(a) = \sum_{\alpha} \delta_{s^\alpha_i, a}q_\alpha.$$
	To minimize this distance, we find a profile $q_i(a)$ that maximizes the quantity $\sum_{\alpha} \delta_{s^\alpha_i, a}q_\alpha$ for each position $i$. It is clear that this is done by assigning a value $q_i(a)=1$ if $a$ maximizes $p_i(a)$, and $q_i(a)=0$ otherwise. Thus, the profile of the predictor must be that of the consensus sequence, which is only possible if the predictor becomes $\{\sigma^{cons},1\}$.
% subsection predictor_based_on_the_local_lbi_maxima (end)

\subsection{Supplementary figures} % (fold)
\label{sub:supplementary_figures}

	\begin{figure}
		\centering
		\includegraphics[scale=0.5]{SM_figures/Nseq_per_year.png}
		\caption{Supplementary figure 1 - Number of HA sequences per year from year 1990. }
		\label{fig:nseq_per_year}
	\end{figure}
	\begin{figure}
		\centering
		\includegraphics[scale=0.5]{Figures/Nseq_per_month.png}
		\caption{Supplementary figure 2 - Number of HA sequences per month from year 2000. }
		\label{fig:nseq_per_month}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[scale=0.25]{Figures/sample_trajectories.png}
		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the HA protein.}
		\label{fig:sample_trajectories}
	\end{figure}

	% \begin{figure}
	% 	\centering
	% 	\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_misc.png}
	% 	\caption{Supplementary figure 4 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for different types of mutations. \textbf{Top panel}: Number of times the mutation appears in the phylogenetic tree. \textbf{Middle panel}: Binary positions, \emph{i.e.} positions at which only 2 amino acids are ever observed at the same time. \textbf{Bottom panel}: Trajectories with a positive derivative when measured at frequency $f$.}
	% 	\label{fig:pfix_v_freq_misc}
	% \end{figure}
	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time.png}
		\caption{Supplementary figure 4 - \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods). \textbf{B}: Mutations whose trajectories are older or more recent, based on the median age of trajectories when reaching the considered frequency $f$.}
		\label{fig:geospread_and_time}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_syn.png}
		\caption{Supplementary figure 5 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for synonymous mutations.}
		\label{fig:pfix_v_freq_syn}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/GeoSpread_v_time_ha.png}
		\caption{Supplementary figure 6 - Geographic spread of mutations as a function of the time for which they have been present in the population above a frequency of 5\%. Points represent individual mutations and for a population in a given time bin. The line is the average of dots for a given value on the $x$-axis.}
		\label{fig:geospread_v_time}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/EMD_to_future.png}
		\caption{Supplementary figure 7 - Earthmover's distance to the future population for different predictors. A present population consists of all HA sequences sampled in a 4 months time window. Quantities are averaged over all possible "present" populations from the year 2002. Predictors are: \textbf{Global consensus}: Consensus sequence of the present population. Best long-term predictor for a structure-less neutrally evolving population. \textbf{All present population}: All sequences in the present population. Perfect predictor if the population does not change at all through time. \textbf{Cluster-wise consensus}: Consensus sequence for each cluster in the present population. Clusters are based on maxima of the LBI. Sequences are assigned to a given cluster based on their tree distance to the corresponding local maximum.}
		\label{fig:emd_to_future}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/simulation.png}
		\caption{Supplementary figure 8 - Simulations}
		\label{fig:simulations}
	\end{figure}

% subsection supplementary_figures (end)

% section supplementary_material (end)

\end{document}

### 2020-06-30 article.tex
Overleaf version - no h1n1

\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,pre]{revtex4-1}

% \usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage[numbers]{natbib}
\usepackage{color}      
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{enumitem}
\usepackage{csvsimple}
\linenumbers

\newcommand{\RAN}[1]{{\color{blue}Richard: #1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\sref}[1]{S\ref{#1}}

\begin{document}
% \title{Seasonal influenza: the limits of predictability}
% \title{Limited predictability of seasonal influenza virus evolution}
\title{Limited predictability of amino acid substitutions in seasonal influenza viruses}
\date{}

\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Switzerland}
%
\author{John Huddleston}
\affiliation{Molecular Cell Biology, University of Washington, Seattle, WA, USA}
%
\author{Trevor Bedford}
\affiliation{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
%
\author{Richard Neher}
\email{Correspondence to: Richard Neher, Biozentrum, Klingelbergstrasse 70, 4056, Basel, Switzerland. \bf{richard.neher@unibas.ch}}
\affiliation{Biozentrum, Universität Basel, Switzerland}

\begin{abstract}
	Seasonal influenza viruses repeatedly infect humans in part because they rapidly change their antigenic properties and evade host immune responses, necessitating frequent updates of the vaccine composition. 
	Accurate predictions of strains circulating in the future could therefore improve the vaccine match. 
	Here, we studied the predictability of frequency dynamics and fixation of amino acid substitutions.
	Current frequency was strongest predictor of eventual fixation, as expected in neutral evolution. 
    Other properties, such as for example falling in characterized epitopes or high/low \emph{Local Branching Index} (LBI) had little predictive power. 
	Only parallel evolution was found to be predictive of fixation. 
	While the LBI had little power to predict frequency dynamics, is was still successful at picking strains representative of future populations. 
	The latter is due to a tendency of the LBI to be high for consensus-like sequences that are closer to the future than the average sequence. 
	These results are in contrast to simulations of models of adapting populations, where we find the clear signals of predictability. 
	This indicates that the evolution of influenza HA and NA, while driven by strong selection pressure to change, is poorly described by common models of directional selection such as travelling fitness waves.  
\end{abstract}

\maketitle

\section*{Introduction} % (fold)
\label{sec:introduction}

	Seasonal influenza A viruses (IAV) infect about 10\% of the global population every year, resulting in hundreds of thousands of deaths~\cite{WHOfactsheet,petrova_evolution_2017}. Vaccination is the primary measure to reduce influenza morbidity.
	However, the surface proteins hemagglutinin (HA) and neuraminidase (NA) continuously accumulate mutations at a high rate, leading to frequent antigenic changes \cite{petrova_evolution_2017, Shih6283, bhatt_genomic_2011, Koel976}. 
	While a vaccine targeting a particular strain may be efficient for some time, antigenic drift will sooner or later render it obsolete.
	World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. 
	Since developing, manufacturing, and distributing the vaccine takes many months, forecasting the evolution of influenza of essential interest to public health \citep{morris2018predictive,klingen_silico_2018}.

	The number of available high quality HA and NA sequences has increased rapidly over the last 20 years~\cite{bogner2006global, shu2017gisaid} and virus evolution and dynamics can be now be tracked at high temporal and spatial resolution \cite{rambaut_genomic_2008}.
	This wealth of data has given rise to an active field of predicting influenza virus evolution ~\cite{morris2018predictive,klingen_silico_2018}. 
	These models predict the future population of influenza viruses by estimating strain fitness or use proxies of fitness.
	\citet{luksza_predictive_2014}, for example, train a fitness model to capture antigenic drift and protein stability on patterns of epitope and non-epitope mutations. 
	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} use  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein to predict. 
	Finally, \citet{neher_predicting_2014} use branching patterns of HA phylogenies as a proxy for fitness. 
	These branching patterns are summarized by the Local Branching Index (LBI), which was shown to be a proxy of relative fitness in mathematical models of rapidly adapting populations \citep{neher_predicting_2014}.

	The underlying assumption of all these methods is that (i) differences in growth rate between strains can be estimated from sequence or antigenic data and (ii) that these growth rate differences persist for long enough to be predictive of future success. 
	Specific positions in surface proteins are of particular interest in this context.
	The surface proteins are under a strong positive selection and change their amino acid sequence much more rapidly than other IAV proteins or than expected under neutral evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}. 
	Epitope positions, i.e., positions targeted by human antibodies, are expected to change particularly often since virus with altered epitopes can evade existing immune responses \citep{Koel976,wolf_long_2006,Shih6283}.
    It therefore seems plausible that mutations at these positions have a tendency to increase fitness and a higher probability of fixation \citep{strelkowa_clonal_2012}. 
    But one has to be careful to account for the fact that these positions are often ascertained post-hoc \citep{Shih6283} and human immune responses are diverse with substantial inter-individual variation
    \citep{lee_mapping_2019}.

	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

% section introduction (end)

\section*{Results} % (fold)
\label{sec:results}

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

The central quantities that are derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences. If an amino acid $X_i$ is found at position $i$ at a frequency between 5\% and 95\% in the population of a given time bin $t$, then the population is considered polymorphic at position $i$ and at time $t$. This polymorphism is characterized by the frequency of $X_i$, $f_{X_i}(t)$, and also by frequencies of other amino acids at $i$. The series of values $f_{X_i}(t)$ for contiguous time bins constitutes the frequency trajectory of $X_i$. A trajectory is terminated if the corresponding frequency is measured above 95\% (resp.~below 5\%) for two time bins in a row, in which case amino acid $X_i$ is considered as \emph{fixed} (resp.~\emph{absent}) in the population. Otherwise, the trajectory is considered \emph{active}. Examples of trajectories can be seen in figure~\sref{fig:sample_trajectories} of the Supplement.

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.

\subsection*{Predicting future frequencies}
Having observed the frequency trajectory $f(t)$ of a mutation until a given date $t_0$, how much can we say about the future values of $f$ after $t_0$? 
We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f)$ of observing it at a value $f$ at time $t_0 + \Delta t$?

To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. 
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling. 
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. 
Removing them does not change results significantly.

Since rapid sequence evolution of influenza HA and NA mediates immune evasion, one could expect that a significant fraction of new amino acid mutations on rising trajectories in figure~\ref{fig:trajectories} are \emph{adaptive}. 
We could thus expect that most of these trajectories continue to rise after reaching frequency $f_0$, at least for some time. 
A fraction of those would then sweep through the population and fix.

To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$. 
The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C. 
Initially, \emph{i.e.} at time $t_0=0$, this distribution is by construction peaked around $f_0$. 
If a large fraction of the trajectories keep increasing after this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.

However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$.
The fact that this average rose $t<0$ has hence no information for $t>0$ and is due to the conditions by which these trajectories were selected.

Consistent with the average, the frequency distribution of the selected trajectories broadens in time without a significant shift of the mean as time passes. 
After 60 days, the distribution is rather symmetrical around the initial $f_0=0.3$ value, suggesting that the knowledge that the trajectories were rising is lost after two months. 
On a timescale of 60 to 120 days, the only possible prediction is that trajectories are likely to be found in a broad interval around the initial frequency $f_0$. 
After one year the distribution becomes almost flat (excluding mutations that have disappeared or fixed), and the initial peak at $f_0$ is not visible anymore. 
The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with the notion of classic sweeps which take over the population.

While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is uninformative.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the HA gene that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panel \textbf{A} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency. \textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population. Colored dashed lines represent synonymous mutations. Colored solid lines represent amino acid mutations. Error bars represent a 95\% confidence interval.}
	\label{fig:fix_loss_active}
\end{figure*}

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panel \textbf{B} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{B} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.

This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

To investigate further, we try to find categories of mutations that deviate from the neutral expectation of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
%However, even in the former case of the positions from \citet{Koel976}, observations are different from what one would expect if the dynamics where dominated by selective sweeps. 
One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.

Two ways of categorising mutations, however, suggest some power to predict fixation. 
In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into binary trajectories where only two amino acid variants co-circulate and non-binary ones with more than two variants. 
Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation. 
In panel \textbf{D}, we separated mutations that appear more than once or only once in the reconstructed tree (see methods), and found that the former fix more often. 
Panels \textbf{C} and \textbf{D} show that it is possible to gain some information on the chance of fixation of a particular mutation, as was done in panel \textbf{B}. 
However, the predictive power remains small, with the ``top'' curves in panels \textbf{C\&D} being very close to the diagonal.

Since influenza is seasonal in temperate regions, geographic spread and persistence might be predictive of the success of mutations. 
We quantify geographic spread of a mutation by the entropy of its frequency distribution across regions (see methods) and its persistence by the age of the trajectory by the time it reaches frequency $f$. 
Figure~\sref{fig:geospread_and_time} shows the fixation probabilities as a function of observed frequency for mutations classified according to these scores. 
The two scores also allow a quantitatively moderate distinction between mutations: for a given frequency $f$, mutations found in many regions or those that are older (in the sense that they have taken more time to reach frequency $f$) tend to fix more often than geographically localized mutations or more recent ones, but the effect is small.
These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\sref{fig:geospread_v_time} of SM.
% A natural way to explain these observations would be to refer to the seasonal characteristics of A/H3N2. 
% In temperate regions, the number of A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
% For this reason, one could imagine that some mutations rise in frequency in an ``opportunistic'' way during an epidemic season, but do not survive the following decline in population size. 
% For the same reason, mutations that are localized in a particular region might mostly be due to the epidemic season specific to that region. 
However, it is important to note that sampling biases and heterogeneity across time and space (see supplementary figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month}) make answering such specific hypothesis challenging.
Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non-trivial to measure. 
%It is however interesting to observe that quantities based on the population's geographical and temporal structure can apparently give as much or more information on fixation probability and predictability than genomic or phylogenetic ones.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

\subsection*{Simulations of models of adaptation}
\label{sub:a_simple_model}

The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection. 
One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}. 
We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.

First, we use an additive fitness function, with sequence $(x_1\ldots x_L)$ having a fitness $\sum_i h_i x_i$. 
This implies that for a given genome position $i$, the trait $x_i = 1$ is favored if $h_i>0$ whereas  $x_i = -1$ is favored if $h_i<0$. 
All $h_i$'s have the same magnitude, and only their signs matter. 
Every $\Delta t$ generations, we randomly choose a position $i$ and flip the sign of $h_i$, effectively changing the fitness landscape. 
Individuals in the population now have the opportunity to make an adaptive mutation at site $i$ giving them a fitness advantage $2\vert h \vert$. 
A ``flip'' at position $i$ of the fitness landscape will decrease fitness of all individuals that carried the adapted variant at position $i$ and increases the fitness of those that happened to carry a deleterious variant.

To increase competition between genomes, we designed a second model that includes epistasis. 
Once again, the baseline fitness of a genome is an additive function, this time with values of $h_i$ that do not change through time. 
In addition, we added a component that mimics immune selection.
Every $\Delta t$ generation, we now introduce ``antibodies'' that target a specific sub-sequence of length $l=5$, noted $(x^{ab}_{i_1}, \ldots, x^{ab}_{i_l})$. 
The positions $(i_1\ldots i_l)$ are chosen at random, while the targeted sub-sequence is the dominant state at each position.
Genomes that include the \emph{exact} sub-sequence targeted by the antibody suffer a strong fitness penalty. However, a single mutation away from that sub-sequence removes this penalty completely, resulting in a fitness landscape with very strong epistasis. 
This has the effect of triggering a strong competition between adaptive mutations: for a given antibody, $l=5$ possible mutations are now adaptive, but combinations of these mutations do not bring any fitness advantage.

Having simulated populations in these two fitness landscapes, we perform the same analysis of frequency trajectories as for the real influenza data. 
Figure~\sref{fig:simulations} of the SM shows the $P_{fix}(f)$ as a function of $f$ for the two models and for different values of the inverse rate of change $\Delta t$ of the fitness landscape. 
For all models, this curve deviates significantly from the diagonal. 
This is most evident for the case of a simple additive fitness landscape that changes rarely $\Delta t=1000$: rising mutations almost always fix in the population, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent. 
This is corroborated by visual inspection of the trajectories, which shows that evolution in this regime is driven by regular selective sweeps that take a typical time of $\sim 400$ generations. 
In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal. 
However, even in an extremely fast changing fitness landscape with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ still significantly differs from $f$.

These models are not meant to be accurate models of influenza viruses evolution. 
But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.

\subsection*{Why do predictions work?}
The quasi-neutral statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
Likewise, the LBI, a quantity that correlates with fitness in mathematical models and is used to predict future influenza populations \citep{neher_predicting_2014}, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:pfix_stratified}. 
To resolve this conundrum, we first note that criterion by which predictive power for influenza was measured in \citep{neher_predicting_2014} was the distance between the strain with the highest LBI and the future population, not the ability of the LBI to predict dynamics.
The distance was compared to the average distance between the present and future population, as well as the post-hoc optimal representative and the future.

To quantify the ability of the LBI and other measures to pick good representatives of the future, we construct a large tree of HA sequences with 100 sequences in non-overlapping time bins of 4 months from year 2003 to 2019 (a total of 4402 as some 4 month intervals contain less than 100 sequences).
Each time bin is considered as a snapshot of the A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.

To assess the ability of the LBI to pick a close representative of the future, we compute the LBI of each node of one time bin in the tree using only the leaves that belong to that time bin. 
The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for randomly chosen strain. 
The figure shows the distance averaged over all possible values of $t$ for $\Delta t$ between $0$ and $32$ months, giving us an average efficiency of a predictor over 16 years of influenza evolution.

The strain with the highest LBI is consistently closer to the future than the average strain by about 1-2 amino acids, while the overall distance increases linearly due to the continuous evolution of the population.
We hence reproduce previous results showing that the LBI picks closer than average representatives \citep{neher_predicting_2014}.
To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different measure: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence). 
The choice is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population in terms of Hamming distance (see SM section \ref{sub:consensus_sequence_as_a_predictor}). 
Figure~\ref{fig:LBI_consensus} shows that the consensus sequence is in fact a equally good or even slightly better representative of the future than the sequence with highest LBI 
(note that this sequence does \emph{not} necessarily exist in the population).

This near equivalence of the consensus and the strain with highest LBI is curious can be explained as follows.
The LBI tends to be high for nodes in a tree that are close to the root of a dense and large clade. 
A typical sample of influenza HA sequences fall into a small number of recognizable clades, and the strains with maximal LBI will often be close to the root of the largest of those clades. 
This root of the largest clade, however, will therefore often be close to the consensus of the whole population, explaining the similar distance patterns.
To test that hypothesis, we measure the hamming distance from the sequence of the top LBI strain to the consensus sequence for populations of all time bins. 
Panel \textbf{B} of figure~\ref{fig:LBI_consensus} shows these distances, scaled with respect to an average strain (details in caption). It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only once is the sequence of the top-LBI strain farther away from the consensus than the average sequence is. Moreover, the sequence of the top-LBI strain \emph{exactly} matches the consensus in 19 cases.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel5.png}
	\caption{\textbf{A}: Average Hamming distance of the sequences of different predictors to HA sequences of future influenza populations, themselves averaged over all ``present'' populations from year 2003 to 2019. Predictors are: a randomly picked sequence in the present population; the sequence of the strain with the highest LBI in the present population; the consensus sequence of the present population. \textbf{B}: Scaled Hamming distance between the sequence of the top LBI strain and the consensus sequence for populations at different dates. The scaling is such that for each date, the Hamming distance between a strain from the population and the consensus is on average $1$. The strain with the highest LBI is almost always closer to the consensus sequence than the average strain.}
	\label{fig:LBI_consensus}
\end{figure*}

% section results (end)

\section*{Discussion} % (fold)
\label{sec:discussion}

The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time. 
Under such conditions, it is expected that frequency trajectories will show an persistent behavior which would make them predictable for some time. 
However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution.
This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.

The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency. 
In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}. 
Only mutations at the seven ``Koel''-positions fixed more often than expected by chance.
The only other stratification that had a measurable effect on fixation were was whether or not convergent evolution (at the level of amino acid or position) was observed. 
On average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.

This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}. 
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
In addition, every successful variant will produce population immunity limiting its future spread. 
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

Methods for predicting the future evolution of influenza either construct explicit fitness models \citep{luksza_predictive_2014}, use historical patterns of evolution \citep{luksza_predictive_2014,bush_predicting_1999}, phenotypic assays \citep{neher_prediction_2016,steinbruck_inference_2012}, or dynamic or phylogenetic patterns \citep{neher_predicting_2014,klingen_sweep_2018}. 
The goal of these methods is to pick strains that are good representatives of future populations and could serve as vaccine candidates \citep{morris2018predictive}.

The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work. 
Picking representatives of the future and predicting frequency dynamics are distinct objectives and that success at the former (as compared to random picks) is consistent with a lack of predictable dynamics.
The strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one -- consistent with the result in \citep{neher_predicting_2014} -- despite the fact future frequencies are not predicted by the LBI.
In fact, the very simple method of taking the consensus sequence of all present strains performs slightly but consistently better than picking the strain with the LBI. 
The consensus sequence is the best possible predictor for a neutrally evolving population, and does not attempt to model fitness in any way. 
While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, the reason it is predictive for influenza, does not seem to due to an ability to measure fitness of influenza viruses from genealogical structure.
Instead, we believe it picks closer than average strains simply because it has the tendency to be maximal at the base of large and dense clades. 
These basal genotypes are closer to the future populations than the current tips of the tree and hence a better predictor on average.

At the same time, influenza virus phylogenies show clear deviations from those expected from the neutral Kingman coalescent, similar to those expected under Bolthausen-Sznitman coalescent (BSC) processes that are generated by traveling wave models of rapid evolution \citep{neher_genealogies_2013,desai_genetic_2013}.
The correspondence between the BSC and traveling wave models comes from transient exponential amplification of fit strains before these fitness differences are wiped out by further mutation. 
This exponential amplification generates long-tailed effective offspring distributions which in turn can leads to genealogies described by the BSC \citep{schweinsberg_coalescent_2003,neher_genealogies_2013}. 
Many processes other than selection, including seasonality and spatio-temporal heterogeneity, can generate effective long tailed offspring distributions even in absence of bona-fide fitness differences, which might explain ladder-like non-Kingman phylogenetic trees.

% section discussion (end)

\section*{Methods} % (fold)
\label{sec:methods}

\subsection*{Data and code availability} % (fold)
\label{sub:data}
	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}. 
	Strain names and accession numbers are given as tables in two supplementary files. \\
	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}. 
% subsection data (end)

\subsection*{Frequency trajectories} % (fold)
\label{sub:frequency_trajectories}

	For a set of sequences in a given time bin, we compute frequencies of amino acids at each position by simple counting. We make the choice of not applying any smoothing method in an attempt to be as close to the data and ``model-less'' as possible. This is especially important for the short term prediction of frequency trajectories, as estimations of the ``persistence time'' of a trajectory might be biased by a smoothing method. \\
	We compute frequency trajectories based on the frequencies of amino acids. A trajectory begins at time $t$ if an amino acid is seen under the lower frequency threshold of 5\% (resp. above the higher threshold of 95\%) for the two time bins preceding $t$, and above this lower threshold (resp. below the higher threshold) for time bin $t$. It ends in the reciprocal situation, that is when the frequency is measured below the lower threshold (resp. above the higher threshold) for two time bins in a row. \\
	In order to avoid estimates of frequencies that are too noisy, we only keep trajectories that are based on a population of at least 10 sequences for \emph{each} time bin. As said in the Results section, we also restrict the analysis to trajectories that begin at a $0.$ frequency, in part to avoid double counting. We find a total of 460 such trajectories. However, only 106 reach a frequency of 20\%, on which figure~\ref{fig:fix_loss_active} is based for instance.  \\
	Note that the fact that we use samples of relatively small sizes -- at least for some time bins -- leads to biases in the estimation of frequencies. We show in Supplementary Material that these biases are generally small and do not induce any qualitative changes to results presented here.

% subsection frequency_trajectories (end)

\subsection*{Local Branching Index} % (fold)
\label{sub:local_branching_index}

	LBI was introduced in \cite{neher_predicting_2014} as an approximation of fitness in populations evolving under persistent selective pressure that is fully based on a phylogenetic tree. It relies on the intuition that the tree below high-fitness individuals will show dense branching events, whereas absence of branching is a sign of low-fitness individuals. Quantitatively, the LBI $\lambda_i(\tau)$ of a node $i$ is the integral of all of the tree's branch length around $i$, with an exponentially decreasing weight $e^{-t/\tau}$ with $t$ being the branch length. When considering a time binned population, the LBI is computed once for each time bin by considering only the leaves of the tree that belong to the time bin. This means that only branches that ultimately lead to a leaf that belongs to the time bin are considered in the integration.\\ 
	$\tau$ is the time scale for which the tree is informative of the fitness of a particular node. Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA). We have observed that given our method to predict the future from present populations corresponding to time bins of 4 months, changing the value of $\tau$ has little effect on the pick of the top LBI strain. By retrospectively optimizing its value, it is possible to reduce the average distance to the population 2 years ahead by $\sim0.25$ amino acids on average, making the LBI method almost as good as the consensus on figure~\ref{fig:LBI_consensus}.

% subsection local_branching_index (end)

\subsection*{Measuring the geographical spread of a mutation} % (fold)
\label{sub:measuring_the_geographical_spread_of_a_mutation}
	For a mutation $X$ we define its regional distribution using the numbers $n_r(X)$ that represent the number of sequences sampled in region $r$ that carry $X$. Regional weights are then defined as 
	$$ w_r(X) = \frac{n_r(X)}{\sum_{r}n_r(X)}.$$
	We can then measure the geographical spread $G(X)$ of $X$ by using the Shannon entropy of the probability distribution $w_r(X)$: 
	$$G(X) = \sum_r w_r(X)\log(w_r(X)).$$
	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
	Region used are the ones defined in the \texttt{nextstrain} tool~\cite{10.1093/bioinformatics/bty407}. Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
% subsection measuring_the_geographical_spread_of_a_mutation (end)

\subsection*{Consensus sequence} % (fold)
\label{sub:consensus_sequence}
	Given a set of $N$ sequences $(\sigma^1,\ldots,\sigma^N)$ based on an alphabet $\mathcal{A}$ (\emph{e.g.} $\mathcal{A}$ has 20 elements for amino acids, 4 for nucleotides), we can define a \emph{profile} distribution $p_i(a)$ by the following expression:
	$$ p_i(a) = \sum_{n=1}^N \delta_{\sigma^n_i,a} $$ 
	where $i$ is a position in the sequence, $\sigma^n_i$ the character appearing at position $i$ in sequence $\sigma^n$, $a$ a character of the alphabet and $\delta$ the Kronecker delta. The profile $p_i(a)$ simply represents the fraction of sequences which have character $a$ at position $i$. \\
	We then simply define the consensus sequence $\sigma^{cons}$ such that 
	$$\sigma^{cons}_i = \text{argmax}_a \;p_i(a).$$
	In other words, the consensus sequence is the one that has the dominant character of the initial set of sequences at each position.   
% subsection consensus_sequence (end)

\subsection*{Earth Mover's Distance} % (fold)
\label{sub:earth_mover_s_distance}

	In order to measure the distance of several predictor sequences to the future population, we rely on the \emph{EarthMover's Distance} (EMD), a metric commonly applied in machinelearning to compare collections of pixels or words \cite{710701,10.5555/3045118.3045221}. Here, we apply it to compute the distance between the sequences of two populations, noted as $\mathcal{X} = \left\{(x^n, p^n)\right\}$ and $\mathcal{Y}=\left\{(y^m,q^m)\right\}$ with $n\in\{1\ldots N\}$ and $m\in\{1\ldots M\}$. In this notation, $x^n$ and $y^m$ are sequences, and $p^n$ and $q^m$ are the frequencies at which these sequences are found in their respective populations. For convenience, we also define $d_{mn} = H(x^n,y^m)$ as the Hamming distance between pairs of sequences in the two populations. \\
	We now introduce the following functional
	$$ F(\mathbf{w}) = \sum_{n,m} d_{nm}w_{nm}, $$
	with $\mathbf{w} = \{w_{nm}\}$ being a matrix of positive weights. The EMD between the two populations $\mathcal{X}$ and $\mathcal{Y}$ is now defined as the minimum value of function $F$ under the conditions 
	\begin{equation*}
		% \begin{split}
			\sum_{n=1}^N w_{nm} = q^m,\;\; \sum_{m=1}^M w_{nm} = p^n, \,\text{and}\;\; w_{nm}\geq 0
		% \end{split}
	\end{equation*}
	Intuitively, the weight $w_{nm}$ tells us how much of sequence $x^n$ is ``moved'' to sequence $y^m$. The functional $F$ sums all of these moves and attributes them a cost equal to the Hamming distance $d_{nm}$. The conditions on weights in $\mathbf{w}$ ensure that all the weight $p^n$ of $x^n$ is ``moved'' to elements in $\mathcal{Y}$ and vice versa. \\
	The minimization is easily performed by standard linear optimization libraries. Here, we use the Julia library {JuMP}~\cite{DunningHuchetteLubin2017}.
% subsection earth_mover_s_distance (end)

% section methods (end)

\bibliographystyle{unsrtnat}
\bibliography{references}

\newpage 
\appendix
\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}
\setcounter{table}{0}
\renewcommand{\tablename}{Table S}
\onecolumngrid

\section*{Supplementary material} % (fold)
\label{sec:supplementary_material}

\subsection{Consensus sequence as a predictor for neutrally evolving populations} % (fold)
\label{sub:consensus_sequence_as_a_predictor}

	We consider the case of a neutrally evolving and structure-less population, such as the one in the Wright-Fisher model of evolution~\cite{10.1080/10635150500354860}. At an initial time $t=0$, the population consists of $N$ individuals with genomes $(\sigma^1\ldots\sigma^N)$ of length $L$ (not necessarily distinct). \\
	% We introduce the distribution $f(\sigma)$ representing the frequency at which a particular genome $\sigma$ appears in the population, that is 
	% $$ f(\sigma) = \frac{1}{N}\sum_{n=1}^N \delta_{\sigma, \sigma^n}$$
	% where $\delta_{x,y}$ is the Kronecker delta. \\
	We make two hypotheses about this population. We first suppose that \emph{no} mutations occur during the evolution of this population. This may seem surprising and is of course not true in the case of influenza. This assumption is however in line with the fact that the object of this work is to predict the outcome of \emph{already existing} mutations in the influenza population. The prediction of mutations that we have not yet seen is not in its scope. Thus, assuming that no new mutations take place can be seen as a simple way to model the fact that we have no information about such events. \\
	The second assumption is that the population evolves in a completely neutral way, meaning that the average number of descendants of each genome $\sigma^n$ is the same. Let us now consider the population after it has evolved for a long time $t\gg T$ where $T$ is the typical coalescence time (for the Wright-Fisher model, $T=2N$). At this point, all individuals in the future population will descend from a unique individual $n_0$ in the $t=0$ population. Our two hypotheses now allow us to make two statements. First, since no new mutations are allowed, the population at $t\gg T$ will be clonal, with all individuals having genome $\sigma^{n_0}$. Second, since the evolution is neutral and does not favour any genome in particular, the probability that $\sigma^{n_0}$ is equal to a given genome $\sigma$ is $1/N$. In other words, the probability that a genome at $t=0$ ultimately becomes the ancestor of all the future population is equal to its frequency in the $t=0$ population. \\

	We now try to find the genome $\sigma$ that best predicts the future population on the long run, that is for $t\gg T$. Here, we take best to mean that the predictor minimizes $H(\sigma,\sigma^{n_0})$ where $H$ is the Hamming distance defined by
	\begin{equation}
	 	H(\sigma^a, \sigma^b) = \sum_{i=1}^L (1 - \delta_{\sigma^a_i, \sigma^b_i}),
	 	\label{eq:hdist}
	\end{equation} 
	with $\sigma_i$ being the character appearing at position $i$ of genome $\sigma$ and $\delta$ the Kronecker delta. Since we do not know $n_0$, we have to average over all its possible values. $\sigma$ must thus minimize the following quantity: 
	\begin{equation}
		\begin{split}
			\langle H(\sigma, \sigma^{n_0})\rangle_{n_0} &= \sum_{n=1}^N H(\sigma, \sigma^n)\\ 
			&= \sum_{i=1}^L\sum_{n=1}^N (1 - \delta_{\sigma_i,\sigma^n_i})
		\end{split}
	\label{eq:2}
	\end{equation}
	by using the definition of the Hamming distance. We now assume that characters at each positions of the genomes can be indexed by an integer $a$ running from $1$ to $q$. For instance, if these were amino acid sequences, we could index the $20$ amino acids by $a$ running from $1$ to $q=20$. We rewrite the Kronecker delta in the previous expression using this indexation:
	$$ \delta_{\sigma_i,\sigma_i^n} = \sum_{a=1}^q \delta_{\sigma_i,a}\delta_{\sigma_i^n,a}.$$
	We also introduce the \emph{profile} frequencies $p_i(a)$ of the population at time $t=0$: 
	\begin{equation}
		p_i(a) = \sum_{n=1}^N \delta_{\sigma^n_i, a}.
		\label{eq:profile}	
	\end{equation}
	$p_i(a)$ represents the frequency at which character $a$ appears at position $i$ in genomes of the initial population. \\
	Equation~\ref{eq:2} now becomes
	\begin{equation}
		\begin{split}
			\langle H(\sigma, \sigma^{n_0})\rangle_{n_0} &= \sum_{i=1}^L\sum_{n=1}^N\left(1 - \sum_{a=1}^q \delta_{\sigma_i,a}\delta_{\sigma_i^n,a}\right)\\
			&= \sum_{i=1}^N \left(1- \sum_{a=1}^q\delta_{\sigma_i,a}p_i(a) \right) \\ 
			&= \sum_{i=1}^L \left(1-p_i(\sigma_i)\right)
		\end{split}
		\label{eq:3}
	\end{equation}
	This means that the genome $\sigma = (\sigma_1\ldots\sigma_L)$ which best predicts the future population according to our definition is the one that minimizes the quantity $(1-p_i(\sigma_i))$ for all positions $i$. This obviously implies that each $\sigma_i$ must be chosen as to maximize $p_i(a)$, that is $\sigma_i$ must be the character that appears the most frequently at position $i$. Thus, $\sigma$ must be the \emph{consensus} sequence of the initial population.

% subsection consensus_sequence_as_a_predictor (end)

\subsection{Predictor based on the local LBI maxima} % (fold)
\label{sub:predictor_based_on_the_local_lbi_maxima}
	In figure~\ref{fig:emd_to_future}, we use several sequences as a predictor of the future population. Distance between two sets of sequences, \emph{i.e.} the predictor sequences and the ones of the future population, is defined as the Earth Mover's Distance (EMD). Here, we show that for a population evolving under the same hypotheses as in section~\ref{sub:consensus_sequence_as_a_predictor}, the best \emph{multiple} sequence long term predictor is again the consensus sequence with weight $1$. \\
	Let the predictor be a set of weighted sequences $\{(s^\alpha, q_\alpha)\}$. We again use the fact that in the long term, a unique sequence $\sigma^{n_0}$ from the present will be the ancestor of the entire population. We want to compute the EMD from the predictor to $\sigma^{n_0}$, that is the EMD between the sets $\mathcal{X}=\{(s^\alpha, q_\alpha)\}$ and $\mathcal{Y} = \{\sigma^{n_0},1\}$. Applying the definition of the Methods section, it follows that the weights $\mathbf{w}$ are in this case equal to the $q_\alpha$s. By averaging over all values of $n_0$, we now obtain
	$$ \left\langle \,\text{EMD}\left(\{(s^\alpha, q_\alpha)\}\right) \right\rangle_{n_0} = \sum_{n=1}^N\sum_{\alpha} H(s^\alpha, \sigma^n)\cdot q_\alpha.$$
	By the same calculation procedure as in the previous section, this expression simplifies to 
	$$ \left\langle \,\text{EMD}\left(\{(s^\alpha, q_\alpha)\}\right)\right\rangle_{n_0} = \sum_{i=1}^L\left( 1 - \sum_{a=1}^q p_i(a)q_i(a) \right), $$
	where the profile of the present population $p_i(a)$ has already been defined, and $q_i(a)$ stands for the profile of the predictor, that is 
	$$ q_i(a) = \sum_{\alpha} \delta_{s^\alpha_i, a}q_\alpha.$$
	To minimize this distance, we find a profile $q_i(a)$ that maximizes the quantity $\sum_{\alpha} \delta_{s^\alpha_i, a}q_\alpha$ for each position $i$. It is clear that this is done by assigning a value $q_i(a)=1$ if $a$ maximizes $p_i(a)$, and $q_i(a)=0$ otherwise. Thus, the profile of the predictor must be that of the consensus sequence, which is only possible if the predictor becomes $\{\sigma^{cons},1\}$.
% subsection predictor_based_on_the_local_lbi_maxima (end)

\subsection{Biases in frequency estimations} % (fold)
\label{sub:biases_in_frequency_estimations}

	The frequency of mutations in a given time-bin is simply performed by computing their frequency in sequences sampled in that time bin. 
	This leads to potential biases in estimating frequencies, that arise for two reasons:
	\begin{enumerate}[label=(\roman*)]
		\item A mutation present at frequency $p$ in the population might be observed at another frequency $f\neq p$ if $f$ is estimated using a sub-sample of the population. 
		\item For a neutrally evolving population, the distribution of frequencies of alleles is of the form $P(p)\propto 1/p$. This means that the amount of alleles at frequency $p$ is lower when $p$ is higher. 
	\end{enumerate}
	To illustrate \emph{(i)}, let us compute the probability that a mutation present at ``real'' frequency $p$ in the population is found to be in a given frequency bin $[f_1,f_2]$ when $p$ is estimated from a sample of size $n$. 
	The sample consists of $n$ observations $\{x_i\}$ with $1\leq i\leq n$, with $x_i=1$ if sequence sequence $i$ of the sample bears the mutation, and $x_i=0$ if not. 
	If $n$ is small with regard to the total population size, we can consider the $x_i$ as random variables with a binomial distribution, meaning that $P(x_i=1)=p$ and $P(x_i=0) = 1-p$.  
	The empirical frequency $f$ is then estimated by taking the average of the $x_i$ variables, that is $f = (x_1 + \ldots + x_n)/n$. 
	If those are independently sampled and $n$ is large enough, the probability of measuring value $f$ is given by the Central Limit Theorem:
	\begin{equation}
		P_{n,p}(f) \propto e^{(f - p)^2 / 2\sigma^2},\;\; \text{where } \sigma^2 = \frac{p(1-p)}{n}.
		\label{eq:ctl}
	\end{equation}
	To compute the probability that this mutation is found in a given frequency bin $[f_1,f_2]$, we integrate this distribution: 
	\begin{equation}
		P_{f_1,f_2}(p,n) = \int_{f_1}^{f_2}\text{d}x P_{n,p}(x). % = \int_{f1}^{f_2}\text{d}x\left(2\pi\sigma^2\right)^{-1/2}e^{(x - p)^2 / 2\sigma^2}
		\label{eq:pinterval}
	\end{equation}
	\begin{figure}
		\centering
		\includegraphics[scale=0.5]{SM_figures/frequency_biases.png}
		\caption{\textbf{Left}: For a mutation present at frequency $p$ in the population, probability of being observed in the frequency bin $[0.1, 0.2]$ as a function of $p$ and for different sample sizes $n$. The dashed black line sketches the (non-normalized) background distribution $P_b(p)$. \textbf{Right}: Expected ``real'' average frequency of mutations found in frequency bin $[f_1,f_2]$ as a function of the centre of the bin $(f_1+f_2)/2$, for different sample sizes.}
		\label{fig:freq_biases}
	\end{figure}

	Function $P_{f_1,f_2}(p,n)$ is shown as a function of $p$ for a fixed interval and for different values of $n$ in the first panel of figure \sref{fig:freq_biases}.
	Note the asymmetry of it: the variance of a binomial distribution of parameter $p$ is small when $p$ is close to $0$ or $1$, and goes through a maximum at $p=0.5$. 
	For this reason, mutations present at frequency $p$ close to $0.5$ have a higher probability of being observed in other frequency bins. 
	On the contrary, this is unlikely for very rare or very frequent mutations.  \\
	We now try to estimate biases in frequency estimation due this phenomenon. 
	Given a set of mutations that have been measured in frequency bin $[f_1, f_2]$, what is the average \emph{real} frequency of these mutations? 
	To compute this, we need to sum $P_{f_1,f_2}(p,n)$ over all possible real frequencies $p$, giving us the amount of mutations that are observed in interval $[f_1,f_2]$, and weigh this sum by the frequency value $p$ as well as by the background distribution of frequencies $P_b(p)\propto 1/p$. 
	This last quantity represents the expected amount of mutations that are present at frequency $p$ in the population. 
	Note that there is no divergence problem as the smallest non zero frequency is $1/N$, where $N$ is the population size. 
	This leads us to the following expression for the average of ``real'' frequencies:
	\begin{equation}
		\begin{split}
			\langle p \rangle (f_1,f_2,n) &= \int_{1/N}^{1-1/N} \text{d}p\, P_{f_1,f_2}(p,n)P_b(p)\,p\\
			&= \int_{1/N}^{1-1/N} \text{d}p\, P_{f_1,f_2}(p,n).
		\end{split}
	\end{equation}
	We have not made normalization explicit in these equations.
	It is simply achieved by dividing the above expression by $\int \text{d}p\, P_{f_1,f_2}(p,n)P_b(p)$.\\
	In the second panel of figure~\sref{fig:freq_biases}, $\langle p \rangle (f_1,f_2,n)$ is plotted as a function of the centre of the interval $[f_1,f_2]$ and for different values of $n$. 
	For sample sizes $n>100$, the biases due to this effect are almost non existent. 
	For smaller samples, for instance $n=10$, they are small but non negligible. 
	However, we argue that this is not a significant problem with respect to the main results presented in this article. 
	First, figure~\sref{fig:nseq_per_month} shows that sample sizes of the order of $n=10$ are only the case for a few months in the period going from year 2000 to 2018. 
	From 2010 and onwards, more than a hundred sequences are available per month for most months.  
	Secondly, even if most samples were in the $n=10$ case, deviations shown in figure~\sref{fig:freq_biases} are small enough that results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} would be \emph{qualitatively} unchanged. \\
	Note that using the centre of the interval as a reference in figure~\sref{fig:freq_biases}, \emph{i.e.} $(f_1 + f_2)/2$, would be correct in the case of a very large $n$ and a flat background distribution $P_b(p)$. 
	For figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} of the main text however, the average frequency of mutations found in an interval $[f_1, f_2]$ is computed by taking the average of the observed frequencies, and not the centre of the interval. 
	This partially takes into account biases considered here, as the background distribution $P_b(p)$ is then accounted for, even though it is equivalent to assuming infinite sample sizes.

% subsection biases_in_frequency_estimations (end)
\newpage
\subsection{Cutting off the HA1 159S branch}

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.55]{SM_figures/nextstrain_h3n2_ha_5.png}
	\caption{Tree used for this study, based on a random selection of 100 strains per month from year 2002 to 2018. Nodes and branches are colored according to the amino acid found at position HA1:159. The HA1 159S mutation is visible as a thin but long  light-greened color branch, coalescing with the ``trunk'' around year 2013.}
	\label{fig:tree_159S}
\end{figure}

The analysis of the main text is in a large part based on the probability of fixation of mutations. 
The motivation underlying this choice is the relatively short coalescence time of the H3N2 influenza population, typically around three years. 
This can be seen in figure~\ref{fig:fix_loss_active} of the main text, which shows the typical lifetime of frequency trajectories, ending in fixation or loss after at most 3 years in most cases.  
The tree in figure~\sref{fig:tree_159S} is another illustration of this: for the most part of it, a ``trunk'' is clearly identifiable, and lineages that depart from it have a relatively short lifetime. \\
This is no longer the case since the year $\sim 2013$: two clades have been competing since then, with no definite way to identify a trunk in the tree. 
The clade defined by the HA1$\,$159S mutation, colored in light green on figure~\sref{fig:tree_159S}, is one of these two competing lineages. 
Because of this particular situation, the number of mutations fixating in the population is strongly reduced, as a mutation must appear in both clades to reach a frequency of $1$. 
This is a potential flaw in our analysis, which concentrates on mutations fixating. \\
For this reason, we decided to re-run our analysis after having cut off the HA1$\,$159S clade. 
In other words, we remove from the set of sequences those that carry the HA1$\,$159S mutation. 
Results are shown in figures , equivalent to figures~\ref{fig:fix_loss_active} and~\ref{fig:pfix_stratified} of the main text. 
It is clear that qualitative results are left unchanged when this competing clade is removed. \\
This can be surprising, as almost no complete fixation of an amino acid mutation has occurred since 2013. 
Cutting off the HA1 159S branch should thus result in many new fixations, changing the analysis. 
The reason for the similarity of results can be explained: fixation (resp. loss) of a mutation are defined here as the frequency of this mutation being measured above $95\%$ (resp. $5\%$) frequency for two months in a row. 
As the HA1 159S clade is rather sparsely populated, it reaches frequencies lower than $5\%$ two times (in 2015 and 2017), allowing mutations in the competing clade to ``fix'' as defined here. 
Thus, removing strains carrying HA1 159S does not introduce a significant amount of ``new'' fixation events.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/cutoff_159S_p1.png}
	\caption{Equivalent to figure \ref{fig:fix_loss_active} of the main text, but with strains carrying the HA1 159S mutation removed.}
	\label{fig:cutoff_159S_p1}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/cutoff_159S_p2.png}
	\caption{Equivalent to figure \ref{fig:pfix_stratified} of the main text, but with strains carrying the HA1 159S mutation removed.}
	\label{fig:cutoff_159S_p2}
\end{figure*}

\newpage
\section{H1N1 influenza}

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/panel2_h1n1pdm.png}
	\caption{Equivalent of figure \ref{fig:fix_loss_active} of the main text for H1N1 influenza. \textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency. \textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured. Only new mutations are considered, \emph{i.e.} mutations that were absent in the past. The diagonal dashed line is the expectation from a neutrally evolving population. Colored dashed lines represent synonymous mutations. Colored solid lines represent amino acid mutations. Error bars represent a 95\% confidence interval.}
	\label{fig:fix_loss_active_h1n1}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./SM_figures/panel3_h1n1pdm.png}
	\caption{Equivalent of figure \ref{fig:pfix_stratified} of the main text for the HA gene of H1N1 influenza. Fixation probability $P_{fix}(f)$ as a function of frequency. \textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified_h1n1}
\end{figure*}

\newpage
\subsection{Mutation tables}

% All mutations
\begin{table}[h]
    \centering
    \csvreader[
      tabular=|r*{8}{|c}|,
      table head= \hline Gene & Position & AA & Start date & End date & Shih & Luksza & Koel & Tree counts \\ \hline,
      late after last line=\\\hline,
    ]{Mutation_Tables/all_fixed_mutations.csv}{}%
    {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii & \csvcolviii & \csvcolix}
    \caption{The 30 trajectories that took place between year 2000 and year 2018 and resulted in fixation. Columns \texttt{Shih}, \texttt{Luksza} and \texttt{Koel} respectively indicate whether the position is found in the epitopes lists in (respectively)~\cite{Shih6283}, \cite{luksza_predictive_2014} and \cite{Koel976}. The \texttt{Tree counts} column indicates the number of times the mutation corresponding to the trajectory can be found in the phylogenetic tree. Note that a trajectory is only shown in the table if the sequenced population counts more than $10$ strains at its time of fixation. This explains that only 30 trajectories are displayed, whereas more mutations did fix in this period of time.} %A value $n$ of \texttt{Tree counts} means that there exists at least one time interval of a month during the lifetime of the trajectory where strains found with the corresponding mutation can be traced back to $n$ different occurrences in the tree.
    \label{tab:my_label}
\end{table}
% Shih mutations
\begin{table}[h]
    \centering
    \csvreader[
      tabular=|r*{8}{|c}|,
      table head= \hline Gene & Position & AA & Start date & End date & Fixation & Max. freq. \\ \hline,
      late after last line=\\\hline,
    ]{Mutation_Tables/shih_mutations.csv}{}%
    {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii}
    \caption{Trajectories of mutations at epitope positions in \cite{Shih6283} (\emph{Shih et. al.}) that have been observed at least once above frequency $0.25$. The \texttt{Fixation} column indicates whether the mutation has fixed, disappeared, or is still polymorphic as of October 2018. The \texttt{Max.freq.} column indicates the maximum frequency reached by the trajectory. A maximum frequency of $1$ for mutations that finally disappear is explained by trajectories reaching frequency $1$ for one time bin and going back to lower values for following ones (a frequency above $0.95$ for two time bins in a row defines fixation).} 
    \label{tab:shih}
\end{table}
% Luksza mutations
\begin{table}[h]
    \centering
    \csvreader[
      tabular=|r*{8}{|c}|,
      table head= \hline Gene & Position & AA & Start date & End date & Fixation & Max. freq. \\ \hline,
      late after last line=\\\hline,
    ]{Mutation_Tables/luksza_mutations.csv}{}%
    {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii}
    \caption{Same as table~\sref{tab:shih}, for \cite{luksza_predictive_2014} (\emph{Luksza et. al.}).} 
    \label{tab:luksza}
\end{table}
% Koel mutations
\begin{table}[h]
    \centering
    \csvreader[
      tabular=|r*{8}{|c}|,
      table head= \hline Gene & Position & AA & Start date & End date & Fixation & Max. freq. \\ \hline,
      late after last line=\\\hline,
    ]{Mutation_Tables/koel_mutations.csv}{}%
    {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii}
    \caption{Same as table~\sref{tab:shih}, for \cite{Koel976} (\emph{Koel et. al.}).} 
    \label{tab:koel}
\end{table}
% Wolf mutations
% \begin{table}[h]
%     \centering
%     \csvreader[
%       tabular=|r*{8}{|c}|,
%       table head= \hline Gene & Position & AA & Start date & End date & Fixation & Max. freq. \\ \hline,
%       late after last line=\\\hline,
%     ]{SM_figures/wolf_mutations.csv}{}%
%     {\csvcoli & \csvcolii & \csvcoliii & \csvcoliv & \csvcolv & \csvcolvi & \csvcolvii}
%     \caption{Same as table~\sref{tab:shih}, for \cite{wolf_long_2006} (\emph{Wolf et. al.}).} 
%     \label{tab:wolf}
% \end{table}

\clearpage

\subsection{Supplementary figures} % (fold)
\label{sub:supplementary_figures}

	\begin{figure}[h]
		\centering
		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_year.png}
		\caption{Supplementary figure 1 - Number of HA sequences per year from year 1990. }
		\label{fig:nseq_per_year}
	\end{figure}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.7\textwidth]{Figures/Nseq_per_month.png}
		\caption{Supplementary figure 2 - Number of HA sequences per month from year 2000. }
		\label{fig:nseq_per_month}
	\end{figure}

	\begin{figure}[h]
		\centering
		\includegraphics[width=0.95\textwidth]{Figures/sample_trajectories.png}
		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the HA protein.}
		\label{fig:sample_trajectories}
	\end{figure}

	% \begin{figure}
	% 	\centering
	% 	\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_misc.png}
	% 	\caption{Supplementary figure 4 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for different types of mutations. \textbf{Top panel}: Number of times the mutation appears in the phylogenetic tree. \textbf{Middle panel}: Binary positions, \emph{i.e.} positions at which only 2 amino acids are ever observed at the same time. \textbf{Bottom panel}: Trajectories with a positive derivative when measured at frequency $f$.}
	% 	\label{fig:pfix_v_freq_misc}
	% \end{figure}
	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time.png}
		\caption{Supplementary figure 4 - \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods). \emph{Note}: the words \emph{local} and \emph{global} only reflect the position of the geographic spread of the mutation relative to the median value computed for all mutations found at this frequency. As this median value may change with the considered frequency bin, so does the definition of local and global mutations. \textbf{B}: Mutations whose trajectories are older or more recent, based on the median age of trajectories when reaching the considered frequency $f$.}
		\label{fig:geospread_and_time}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/GeoSpread_v_time_ha.png}
		\caption{Supplementary figure 6 - Geographic spread of mutations as a function of the time for which they have been present in the population above a frequency of 5\%. Points represent individual mutations and for a population in a given time bin. The line is the average of dots for a given value on the $x$-axis.}
		\label{fig:geospread_v_time}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_syn.png}
		\caption{Supplementary figure 5 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for synonymous mutations.}
		\label{fig:pfix_v_freq_syn}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/EMD_to_future.png}
		\caption{Supplementary figure 7 - Earthmover's distance to the future population for different predictors. A present population consists of all HA sequences sampled in a 4 months time window. Quantities are averaged over all possible ``present'' populations from the year 2002. Predictors are: \textbf{Global consensus}: Consensus sequence of the present population. Best long-term predictor for a structure-less neutrally evolving population. \textbf{All present population}: All sequences in the present population. Perfect predictor if the population does not change at all through time. \textbf{Cluster-wise consensus}: Consensus sequence for each cluster in the present population. Clusters are based on maxima of the LBI. Sequences are assigned to a given cluster based on their tree distance to the corresponding local maximum.}
		\label{fig:emd_to_future}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/simulation.png}
		\caption{Fixation probability as a function of frequency for the simulations discussed in the main text. \textbf{Top:} Simulation without antibodies. The three colored curves reflect different rate of change for the fitness landscape. Visual inspection of the frequency trajectories indicates a typical sweep time of $\sim400$ generations. \textbf{Bottom:} Simulation with antibodies. The different colored curves indicate the rate at which antibodies are introduced. }
		\label{fig:simulations}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/simulation_fitness.png}
		\caption{Fixation probability as a function of frequency for the simulations discussed in the main text, with trajectories stratified according to real fitness values. ``High'' and ``low'' fitness classes are defined with respect to the median value. \textbf{Top:} Simulation without antibodies and with changes to the fitness landscape every $dt=10$ generations. \textbf{Bottom:} Simulation with antibodies, with a new antibody every $dt=10$ generations. }
		\label{fig:simulations_fitness}
	\end{figure}

% subsection supplementary_figures (end)

% section supplementary_material (end)

\end{document}

### 2020-07-01 article.tex
Add A/H1N1 and NA

@@
 \newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
+\newcommand{\discuss}[1]{{\color{green}Todiscuss: #1}}
 \newcommand{\sref}[1]{S\ref{#1}}
@@
 
-	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
+	In this work, we use A/A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
 	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
@@
 
-The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
+The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
 \\
@@
 
-In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
+In this work, we use amino-acid sequences of the HA and NA genes of A/A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
 This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year.
 These sequences are binned in non-overlapping intervals of one month.
-Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date.
+Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/A/H3N2 population at a given date.
 The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).
@@
 
-To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.
+To answer this question retrospectively, we use all frequency trajectories extracted from A/A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.
 The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
+	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
 	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories.
 Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$?
-\textbf{B}: All frequency trajectories of amino acid mutations in the HA gene that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point.
+\textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point.
 Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status.
@@
 We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
-Panel \textbf{A} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency.
+Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency.
 Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
-This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
+This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
 We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
-	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency.
-\textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured.
+	\includegraphics[width=0.9\textwidth]{./Figures/Panel2.png}
+	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency for A/H3N2 HA and NA.
+\textbf{B}: Same as \textbf{A} for A/H1N1.
+\textbf{C}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured, for A/H3N2 HA and NA.
 Only new mutations are considered, \emph{i.e.} mutations that were absent in the past.
@@
 Colored solid lines represent amino acid mutations.
-Error bars represent a 95\% confidence interval.}
+Error bars represent a 95\% confidence interval.
+\textbf{D}: Same as \textbf{C} for A/H1N1.}
 	\label{fig:fix_loss_active}
@@
 The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
-Panel \textbf{B} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA. 
-For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
+Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
+\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
 This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
@@
 Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
-Panel \textbf{B} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
+Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/A/H3N2 influenza are in good agreement with the neutral expectation.
+\discuss{Add smth for A/H1N1}
 
-This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
 If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
-In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
+In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}
 
-To investigate further, we try to find categories of mutations that deviate from the neutral expectation of figure~\ref{fig:fix_loss_active}. 
+To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
 We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
@@
 No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
-%However, even in the former case of the positions from \citet{Koel976}, observations are different from what one would expect if the dynamics where dominated by selective sweeps. 
 One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.
@@
 
+\discuss{We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
+Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
+Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case. 
+However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
+Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.}
+
 Since influenza is seasonal in temperate regions, geographic spread and persistence might be predictive of the success of mutations. 
 We quantify geographic spread of a mutation by the entropy of its frequency distribution across regions (see methods) and its persistence by the age of the trajectory by the time it reaches frequency $f$.
-Figure~\sref{fig:geospread_and_time} shows the fixation probabilities as a function of observed frequency for mutations classified according to these scores. 
+Figures~\sref{fig:geospread_and_time_h3n2} and \sref{fig:geospread_and_time_h1n1pdm} show the fixation probabilities as a function of observed frequency for mutations classified according to these scores.
 The two scores also allow a quantitatively moderate distinction between mutations: for a given frequency $f$, mutations found in many regions or those that are older (in the sense that they have taken more time to reach frequency $f$) tend to fix more often than geographically localized mutations or more recent ones, but the effect is small.
 These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\sref{fig:geospread_v_time} of SM.
-% A natural way to explain these observations would be to refer to the seasonal characteristics of A/H3N2. 
-% In temperate regions, the number of A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
+% A natural way to explain these observations would be to refer to the seasonal characteristics of A/A/H3N2. 
+% In temperate regions, the number of A/A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
 % For this reason, one could imagine that some mutations rise in frequency in an ``opportunistic'' way during an epidemic season, but do not survive the following decline in population size. 
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
+	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
 	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency.
@@
 
-The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.
+\discuss{The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.}
 One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}.
@@
 These models are not meant to be accurate models of influenza viruses evolution. 
-But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
-We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.
+\discuss{But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
+We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.}
 
@@
 To quantify the ability of the LBI and other measures to pick good representatives of the future, we construct a large tree of HA sequences with 100 sequences in non-overlapping time bins of 4 months from year 2003 to 2019 (a total of 4402 as some 4 month intervals contain less than 100 sequences).
-Each time bin is considered as a snapshot of the A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
+Each time bin is considered as a snapshot of the A/A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
 From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{./Figures/Panel5.png}
+	\includegraphics[width=0.9\textwidth]{./Figures/Panel4.png}
 	\caption{\textbf{A}: Average Hamming distance of the sequences of different predictors to HA sequences of future influenza populations, themselves averaged over all ``present'' populations from year 2003 to 2019.
@@
 The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
-Under such conditions, it is expected that frequency trajectories will show an persistent behavior which would make them predictable for some time. 
-However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution.
-This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
+Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
+However, we could find only limited evidence for such persistent behavior in the past 19 years of A/A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
+This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
 
@@
 In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
-This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency.
+\discuss{This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency.}
 In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}.
@@
 A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
-Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
+Along these lines, it is possible that the A/A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
 In addition, every successful variant will produce population immunity limiting its future spread. 
@@
 	$\tau$ is the time scale for which the tree is informative of the fitness of a particular node.
-Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA).
+Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA).
 We have observed that given our method to predict the future from present populations corresponding to time bins of 4 months, changing the value of $\tau$ has little effect on the pick of the top LBI strain.
@@
 % subsection measuring_the_geographical_spread_of_a_mutation (end)
+
+\subsection*{Assigning a fitness to trajectories} % (fold)
+\label{sub:assigning_a_fitness_to_trajectories}
+
+% subsection assigning_a_fitness_to_trajectories (end)
 
@@
 The analysis of the main text is in a large part based on the probability of fixation of mutations. 
-The motivation underlying this choice is the relatively short coalescence time of the H3N2 influenza population, typically around three years. 
+The motivation underlying this choice is the relatively short coalescence time of the A/H3N2 influenza population, typically around three years. 
 This can be seen in figure~\ref{fig:fix_loss_active} of the main text, which shows the typical lifetime of frequency trajectories, ending in fixation or loss after at most 3 years in most cases.  
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{./SM_figures/cutoff_159S_p2.png}
+	\includegraphics[width=0.9\textwidth]{./SM_figures/pfix_v_freq_strat_h3n2_cutoff159S.png}
 	\caption{Equivalent to figure \ref{fig:pfix_stratified} of the main text, but with strains carrying the HA1 159S mutation removed.}
@@
 
-\newpage
-\section{H1N1 influenza}
-
-\begin{figure*}
-	\centering
-	\includegraphics[width=0.9\textwidth]{./SM_figures/panel2_h1n1pdm.png}
-	\caption{Equivalent of figure \ref{fig:fix_loss_active} of the main text for H1N1 influenza.
-\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency.
-\textbf{B}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured.
-Only new mutations are considered, \emph{i.e.} mutations that were absent in the past.
-The diagonal dashed line is the expectation from a neutrally evolving population.
-Colored dashed lines represent synonymous mutations.
-Colored solid lines represent amino acid mutations.
-Error bars represent a 95\% confidence interval.}
-	\label{fig:fix_loss_active_h1n1}
-\end{figure*}
-
-\begin{figure*}
-	\centering
-	\includegraphics[width=0.9\textwidth]{./SM_figures/panel3_h1n1pdm.png}
-	\caption{Equivalent of figure \ref{fig:pfix_stratified} of the main text for the HA gene of H1N1 influenza.
-Fixation probability $P_{fix}(f)$ as a function of frequency.
-\textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value.
-\textbf{B}: Different lists of epitope positions in the HA protein.
-The authors and the number of positions is indicated in the legend.
-\textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin.
-\textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
-	\label{fig:pfix_stratified_h1n1}
-\end{figure*}
+% \newpage
+% \section{A/H1N1 influenza}
 
@@
 		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_year.png}
-		\caption{Supplementary figure 1 - Number of HA sequences per year from year 1990. }
+		\caption{Supplementary figure 1 - Number of A/H3N2 HA sequences per year from year 1990. }
 		\label{fig:nseq_per_year}
@@
 		\centering
-		\includegraphics[width=0.7\textwidth]{Figures/Nseq_per_month.png}
-		\caption{Supplementary figure 2 - Number of HA sequences per month from year 2000. }
+		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_month.png}
+		\caption{Supplementary figure 2 - Number of H3N@ HA and NA sequences per month from year 2000. }
 		\label{fig:nseq_per_month}
@@
 		\centering
-		\includegraphics[width=0.95\textwidth]{Figures/sample_trajectories.png}
-		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the HA protein.}
+		\includegraphics[width=0.95\textwidth]{SM_figures/sample_trajectories.png}
+		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the A/H3N2 HA protein.}
 		\label{fig:sample_trajectories}
@@
 
-	% \begin{figure}
-	% 	\centering
-	% 	\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_misc.png}
-	% 	\caption{Supplementary figure 4 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for different types of mutations.
-\textbf{Top panel}: Number of times the mutation appears in the phylogenetic tree.
-\textbf{Middle panel}: Binary positions, \emph{i.e.} positions at which only 2 amino acids are ever observed at the same time.
-\textbf{Bottom panel}: Trajectories with a positive derivative when measured at frequency $f$.}
-	% 	\label{fig:pfix_v_freq_misc}
-	% \end{figure}
+	\begin{figure*}
+		\centering
+		\includegraphics[width=0.9\textwidth]{./SM_figures/pfix_v_freq_strat_h1n1pdm.png}
+		\caption{Equivalent of figure \ref{fig:pfix_stratified} of the main text for the HA gene of A/H1N1 influenza.
+Fixation probability $P_{fix}(f)$ as a function of frequency.
+\textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value.
+\textbf{B}: Different lists of epitope positions in the HA protein.
+The authors and the number of positions is indicated in the legend.
+\textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin.
+\textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
+		\label{fig:pfix_stratified_h1n1}
+	\end{figure*}
+
 	\begin{figure}
 		\centering
-		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time.png}
-		\caption{Supplementary figure 4 - \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
+		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time_h3n2.png}
+		\caption{Supplementary figure 4 - Based on A/H3N2 HA and NA.
+\textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
 \emph{Note}: the words \emph{local} and \emph{global} only reflect the position of the geographic spread of the mutation relative to the median value computed for all mutations found at this frequency.
@@
 \textbf{B}: Mutations whose trajectories are older or more recent, based on the median age of trajectories when reaching the considered frequency $f$.}
-		\label{fig:geospread_and_time}
+		\label{fig:geospread_and_time_h3n2}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time_h1n1pdm.png}
+		\caption{Supplementary figure 4 - Based on A/H1N1 HA and NA.
+\textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
+\emph{Note}: the words \emph{local} and \emph{global} only reflect the position of the geographic spread of the mutation relative to the median value computed for all mutations found at this frequency.
+As this median value may change with the considered frequency bin, so does the definition of local and global mutations.
+\textbf{B}: Mutations whose trajectories are older or more recent, based on the median age of trajectories when reaching the considered frequency $f$.}
+		\label{fig:geospread_and_time_h1n1pdm}
 	\end{figure}
@@
 Points represent individual mutations and for a population in a given time bin.
-The line is the average of dots for a given value on the $x$-axis.}
+The line is the average of dots for a given value on the $x$-axis.
+Based on data for A/H3N2 HA.}
 		\label{fig:geospread_v_time}
-	\end{figure}
-
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.9\textwidth]{SM_figures/Pfix_v_freq_syn.png}
-		\caption{Supplementary figure 5 - Probability of a mutation to fix as a function of the frequency at which it is measured in the population, for the HA protein and for synonymous mutations.}
-		\label{fig:pfix_v_freq_syn}
 	\end{figure}
@@
 		\caption{Supplementary figure 7 - Earthmover's distance to the future population for different predictors.
-A present population consists of all HA sequences sampled in a 4 months time window.
+A present population consists of all A/H3N2 HA sequences sampled in a 4 months time window.
 Quantities are averaged over all possible ``present'' populations from the year 2002.
@@
 \textbf{Cluster-wise consensus}: Consensus sequence for each cluster in the present population.
-Clusters are based on maxima of the LBI.
-Sequences are assigned to a given cluster based on their tree distance to the corresponding local maximum.}
+Clusters are based on local maxima of the LBI.
+Sequences are assigned to a given cluster based on their tree branch-length distance to the corresponding local maximum.}
 		\label{fig:emd_to_future}

### 2020-07-06 article.tex
A/A/H3N2 -> A/H3N2

@@
 
-	In this work, we use A/A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
+	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
 	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
@@
 
-The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
+The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
 \\
@@
 
-In this work, we use amino-acid sequences of the HA and NA genes of A/A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
+In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
 This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year.
 These sequences are binned in non-overlapping intervals of one month.
-Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/A/H3N2 population at a given date.
+Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date.
 The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).
@@
 
-To answer this question retrospectively, we use all frequency trajectories extracted from A/A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.
+To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.
 The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise.
@@
 Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
-This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
+This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
 We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.
@@
 Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
-Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/A/H3N2 influenza are in good agreement with the neutral expectation.
+Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
 \discuss{Add smth for A/H1N1}
@@
 These two scores are in fact correlated, with older trajectories representing mutations that are more geographically spread, as can be seen in figure~\sref{fig:geospread_v_time} of SM.
-% A natural way to explain these observations would be to refer to the seasonal characteristics of A/A/H3N2. 
-% In temperate regions, the number of A/A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
+% A natural way to explain these observations would be to refer to the seasonal characteristics of A/H3N2. 
+% In temperate regions, the number of A/H3N2 infections drastically increases during winter, leading to an at first rapidly growing and then declining population of viruses. 
 % For this reason, one could imagine that some mutations rise in frequency in an ``opportunistic'' way during an epidemic season, but do not survive the following decline in population size. 
@@
 To quantify the ability of the LBI and other measures to pick good representatives of the future, we construct a large tree of HA sequences with 100 sequences in non-overlapping time bins of 4 months from year 2003 to 2019 (a total of 4402 as some 4 month intervals contain less than 100 sequences).
-Each time bin is considered as a snapshot of the A/A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
+Each time bin is considered as a snapshot of the A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
 From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.
@@
 Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
-However, we could find only limited evidence for such persistent behavior in the past 19 years of A/A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
+However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
 This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
@@
 
-This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+\discuss{This failure to meaningfully predict mutation frequency trajectories} is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
 These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
 A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
-Along these lines, it is possible that the A/A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
+Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
 In addition, every successful variant will produce population immunity limiting its future spread. 
@@
 The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work.
+\pierre{I think ``low power'' is okay here?
+Not an overstatement.}
 Picking representatives of the future and predicting frequency dynamics are distinct objectives and that success at the former (as compared to random picks) is consistent with a lack of predictable dynamics.
@@
 	$\tau$ is the time scale for which the tree is informative of the fitness of a particular node.
-Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA).
+Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA).
 We have observed that given our method to predict the future from present populations corresponding to time bins of 4 months, changing the value of $\tau$ has little effect on the pick of the top LBI strain.

### 2020-11-17 article.tex
Submitted

@@
-\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,pre]{revtex4-1}
+\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,rmp]{revtex4-1}
 
@@
 \usepackage{csvsimple}
-\linenumbers
+% \linenumbers
 
@@
 \affiliation{Biozentrum, Universität Basel, Switzerland}
+\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland}
 %
@@
 %
-\author{Richard Neher}
+\author{Richard A.~Neher}
 \email{Correspondence to: Richard Neher, Biozentrum, Klingelbergstrasse 70, 4056, Basel, Switzerland. \bf{richard.neher@unibas.ch}}
 \affiliation{Biozentrum, Universität Basel, Switzerland}
+\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland}
 
@@
 	Here, we studied the predictability of frequency dynamics and fixation of amino acid substitutions.
-	Current frequency was strongest predictor of eventual fixation, as expected in neutral evolution. 
-    Other properties, such as for example falling in characterized epitopes or high/low \emph{Local Branching Index} (LBI) had little predictive power. 
-	Only parallel evolution was found to be predictive of fixation. 
-	While the LBI had little power to predict frequency dynamics, is was still successful at picking strains representative of future populations. 
+	Current frequency was the strongest predictor of eventual fixation, as expected in neutral evolution. 
+    Other properties, such as occurrence in previously characterized epitopes or high \emph{Local Branching Index} (LBI) had little predictive power. 
+	Parallel evolution was found to be moderately predictive of fixation. 
+	While the LBI had little power to predict frequency dynamics, it was still successful at picking strains representative of future populations. 
 	The latter is due to a tendency of the LBI to be high for consensus-like sequences that are closer to the future than the average sequence. 
-	These results are in contrast to simulations of models of adapting populations, where we find the clear signals of predictability. 
+	Simulations of models of adapting populations, in contrast, show clear signals of predictability. 
 	This indicates that the evolution of influenza HA and NA, while driven by strong selection pressure to change, is poorly described by common models of directional selection such as travelling fitness waves.
 \end{abstract}
-
+\date{\today}
 \maketitle
@@
 	While a vaccine targeting a particular strain may be efficient for some time, antigenic drift will sooner or later render it obsolete.
-	World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. 
-	Since developing, manufacturing, and distributing the vaccine takes many months, forecasting the evolution of influenza of essential interest to public health \citep{morris2018predictive,klingen_silico_2018}.
+	The World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. 
+	Since developing, manufacturing, and distributing the vaccine takes many months, forecasting the evolution of influenza is of essential interest to public health \citep{morris2018predictive,klingen_silico_2018}.
 
@@
 	This wealth of data has given rise to an active field of predicting influenza virus evolution ~\cite{morris2018predictive,klingen_silico_2018}. 
-	These models predict the future population of influenza viruses by estimating strain fitness or use proxies of fitness.
+	These models predict the future population of influenza viruses by estimating strain fitness or proxies of fitness.
 	\citet{luksza_predictive_2014}, for example, train a fitness model to capture antigenic drift and protein stability on patterns of epitope and non-epitope mutations. 
-	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} use  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein to predict.
+	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} predict fitness by using  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein.
 	Finally, \citet{neher_predicting_2014} use branching patterns of HA phylogenies as a proxy for fitness. 
@@
 	The surface proteins are under a strong positive selection and change their amino acid sequence much more rapidly than other IAV proteins or than expected under neutral evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-	Epitope positions, i.e., positions targeted by human antibodies, are expected to change particularly often since virus with altered epitopes can evade existing immune responses \citep{Koel976,wolf_long_2006,Shih6283}.
+	Epitope positions, i.e., positions targeted by human antibodies, are expected to change particularly often since viruses with altered epitopes can evade existing immune responses \citep{Koel976,wolf_long_2006,Shih6283}.
     It therefore seems plausible that mutations at these positions have a tendency to increase fitness and a higher probability of fixation \citep{strelkowa_clonal_2012}. 
@@
 
-	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
+	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1pdm influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
 	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
 	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
-	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
+	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
 	This observation holds for many different categories of mutations, including mutations at epitope positions. 
-	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
+	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
 	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
@@
 
-The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
-\\
-There are different ways to make this question more specific.
-First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$.
+The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
+The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments. 
+Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to the mutation.
+
+First, we ask whether we can quantitatively predict the frequency of $X$ at future times $f(t)$.
 In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
@@
 
-In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
-This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year.
+We use amino-acid sequences of the HA and NA genes of A/H3N2 (since the year 2000) and A/H1N1pdm (since the year 2009) influenza available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
+This amounts to $44\,976$ HA and $36\,300$ NA sequences for A/H3N2 and $45\,350$ HA and $40\,412$ NA sequences for A/H1N1, with a minimum of 100 per year.
 These sequences are binned in non-overlapping intervals of one month.
-Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date.
+Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the influenza population at a given date.
 The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).
 
-The central quantities that are derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences.
+The central quantities that we derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences.
 If an amino acid $X_i$ is found at position $i$ at a frequency between 5\% and 95\% in the population of a given time bin $t$, then the population is considered polymorphic at position $i$ and at time $t$.
@@
 In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
-If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.
+The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments. 
+By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
+If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
 
@@
 
-To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.
+To answer this question retrospectively, we use all frequency trajectories extracted from HA and NA sequences that satisfy these conditions for a given $f_0$.
 The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise.
@@
 
-For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. 
+For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$.
+The same analysis was performed for A/H1N1pdm, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
 Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
@@
 
-However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. 
+However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps. 
 The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
-This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$.
-The fact that this average rose $t<0$ has hence no information for $t>0$ and is due to the conditions by which these trajectories were selected.
+This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
+Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
+This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.
 
@@
 The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
-This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with the notion of classic sweeps which take over the population.
+This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with a dynamic dominated by sweeps taking over the population.
 
-While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is uninformative.
+While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is weakly informative.
 
@@
 0).
+Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.
 \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$.
@@
 \subsection*{Prediction of fixation or loss}
-Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
+Instead of predicting future frequency, let's consider the long-term goal of predicting the probability that a mutation fixes in the population. 
 We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
@@
 This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
-We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.
+We also note that the fraction of lost trajectories increases sharply at small times with 40\% of mutations observed above 25\% frequency being lost within one year for A/H3N2, while it takes longer to fix a mutation in the whole population.
 
@@
 The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
-Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
-\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
+Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
+For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
 This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
-Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$.
+% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
+% \begin{equation}
+%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
+%     \begin{cases}
+%     1-e^{-2sfN} & sfN \gg 1 \\
+%     f & |sN| \ll 1 \\
+%     e^{-2s(1-f)N} & -sN \gg 1  
+%     \end{cases}
+% \end{equation}
+% The result for the neutral case is intuitive:
 A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
 Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
-Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
-\discuss{Add smth for A/H1N1}
+Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it.
+In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.
 
-\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
-If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
-In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}
+This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured. 
+In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
 
-To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
+Next, we searched for features of mutations that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
 We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
@@
 Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
-Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.
+Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.
 
-Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
-Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
+Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions. 
+Mutations at these position might mediate immune escape and are therefore likely under strong selection and show sweep-like behavior. 
 We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}.
@@
 No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
-One should also note that many of these positions were determined post-hoc and enriched for positions that experienced rapid substitutions before the publication of the respective studies.
+One should also note that many of these positions were determined post-hoc and might be enriched for positions that experienced rapid substitutions before the publication of the respective studies.
 
 Two ways of categorising mutations, however, suggest some power to predict fixation. 
-In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into binary trajectories where only two amino acid variants co-circulate and non-binary ones with more than two variants. 
+In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants.
 Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation. 
@@
 
-\discuss{We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
+We conduct the same analysis on A/H1N1pdm influenza, with results shown in figure \sref{fig:pfix_stratified_h1n1}. 
 Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
 Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case. 
-However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
-Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.}
+However, this could be due to the shorter time period over which A/H1N1pdm evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
+Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.\\
 
@@
 	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
-	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency.
-\textbf{A}: Mutation with higher or lower LBI values, based on their position with respect to the median LBI value.
+	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency, for A/H3N2 influenza.
+Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1.
+\textbf{A}: HA mutations with higher or lower LBI values, based on their position with respect to the median LBI value.
 \textbf{B}: Different lists of epitope positions in the HA protein.
 The authors and the number of positions is indicated in the legend.
-\textbf{C}: Mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin.
-\textbf{D}: Mutations that appear once or more than once in the tree for a given time bin.}
+\textbf{C}: HA and NA mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin.
+\textbf{D}: HA and NA mutations that appear once or more than once in the tree for a given time bin.}
 	\label{fig:pfix_stratified}
@@
 
-\discuss{The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.}
-One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}.
+The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the idea that seasonal influenza virus evolution is driven by rapid directed positive selection.
+One possible explanation for the weakly predictable behaviour of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
 We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
@@
 In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal. 
-However, even in an extremely fast changing fitness landscape with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ still significantly differs from $f$.
+However, it takes an extremely fast changing fitness landscape to push $P_{fix}$ close to the diagonal: with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ differs from $f$ in a way that is comparable to what is observed in A/H1N1pdm influenza.
 
 These models are not meant to be accurate models of influenza viruses evolution. 
-\discuss{But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
-We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.}
+But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
+We conclude that the pattern in figure~\ref{fig:fix_loss_active} may not be a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.
 
 \subsection*{Why do predictions work?}
-The quasi-neutral statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
+The statistics of frequency trajectories seem to be in conflict with the notion that influenza evolution is predictable.
 Likewise, the LBI, a quantity that correlates with fitness in mathematical models and is used to predict future influenza populations \citep{neher_predicting_2014}, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:pfix_stratified}.
-To resolve this conundrum, we first note that criterion by which predictive power for influenza was measured in \citep{neher_predicting_2014} was the distance between the strain with the highest LBI and the future population, not the ability of the LBI to predict dynamics.
+To resolve this conundrum, we first note that the criterion by which predictive power for influenza was measured in \citep{neher_predicting_2014} was the distance between the strain with the highest LBI and the future population, not the ability of the LBI to predict dynamics.
 The distance was compared to the average distance between the present and future population, as well as the post-hoc optimal representative and the future.
@@
 Each time bin is considered as a snapshot of the A/H3N2 influenza population and we will refer to sequences in time bin $t$ as the population of the \emph{present}. 
-From this present population, we will predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.
+From this present population, we predict \emph{future} populations in time bin $t+\Delta t$, using only sequences in time bin $t$ and before.
 
 To assess the ability of the LBI to pick a close representative of the future, we compute the LBI of each node of one time bin in the tree using only the leaves that belong to that time bin. 
-The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for randomly chosen strain.
+The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for a randomly chosen strain.
 The figure shows the distance averaged over all possible values of $t$ for $\Delta t$ between $0$ and $32$ months, giving us an average efficiency of a predictor over 16 years of influenza evolution.
@@
 We hence reproduce previous results showing that the LBI picks closer than average representatives \citep{neher_predicting_2014}.
-To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different measure: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence).
+To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different predictor: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence).
 The choice is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population in terms of Hamming distance (see SM section \ref{sub:consensus_sequence_as_a_predictor}).
 Figure~\ref{fig:LBI_consensus} shows that the consensus sequence is in fact a equally good or even slightly better representative of the future than the sequence with highest LBI 
-(note that this sequence does \emph{not} necessarily exist in the population).
+(note that the consensus sequence does \emph{not} necessarily exist in the population).
 
-This near equivalence of the consensus and the strain with highest LBI is curious can be explained as follows.
+This near equivalence of the consensus and the strain with highest LBI can be explained as follows:
 The LBI tends to be high for nodes in a tree that are close to the root of a dense and large clade. 
 A typical sample of influenza HA sequences fall into a small number of recognizable clades, and the strains with maximal LBI will often be close to the root of the largest of those clades. 
-This root of the largest clade, however, will therefore often be close to the consensus of the whole population, explaining the similar distance patterns.
+This root of the largest clade will often be close to the consensus of the whole population, explaining the similar distance patterns.
 To test that hypothesis, we measure the hamming distance from the sequence of the top LBI strain to the consensus sequence for populations of all time bins. 
@@
 
-The premises of predicting the trajectory of a mutation are (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
+Predicting the trajectory of a mutation requires (i) significant fitness difference between genomes carrying different variants at the site and (ii) a selection pressure that changes slowly over time.
 Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
-However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
-This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
+However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution. 
+This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
 
-The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
-In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
-\discuss{This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency.}
-In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}.
-Only mutations at the seven ``Koel''-positions fixed more often than expected by chance.
-The only other stratification that had a measurable effect on fixation were was whether or not convergent evolution (at the level of amino acid or position) was observed. 
-On average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.
+The primary focus in this work was the investigation of frequency trajectories of new amino acid mutations.  
+In the short term, we found that on average the direction of trajectory does not persist for longer than a few months. 
+Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''. 
+This suggests that selective sweeps are not representative of typical trajectories.
 
-\discuss{This failure to meaningfully predict mutation frequency trajectories} is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
-A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
-Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
-In addition, every successful variant will produce population immunity limiting its future spread. 
-These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.
+On a longer timescale, we investigated the probability that a novel mutation observed at frequency $f$ fixes.
+In neutral models of evolution this probability equals $f$, while it should be higher or lower than $f$ for mutations with a beneficial or deleterious effect on fitness, respectively.
+However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation. 
+In figure~\ref{fig:pfix_stratified}, we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$. 
+Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.
+Despite the lack of predictability of mutation frequency trajectories, influenza surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
 
-Methods for predicting the future evolution of influenza either construct explicit fitness models \citep{luksza_predictive_2014}, use historical patterns of evolution \citep{luksza_predictive_2014,bush_predicting_1999}, phenotypic assays \citep{neher_prediction_2016,steinbruck_inference_2012}, or dynamic or phylogenetic patterns \citep{neher_predicting_2014,klingen_sweep_2018}.
+% The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
+% In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively, an effect readily seen in simulations of simple models of evolution \citep{kimura_diffusion_1964}.
+% However, in the case of influenza, the probability of fixation differs little from $f$, making the current frequency of a mutation the best predictor for fixation.
+% In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that were previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low LBI \citep{neher_predicting_2014}.
+% Most of these methods did not result in an increased predictability: only mutations at the seven ``Koel''-positions fixed more often than expected by chance. 
+% Other stratification schemes had a measurable effect on fixation, such as whether or not convergent evolution (at the level of amino acid or position) was observed, or whether the position had seen competition between more than two amino acids in the past.
+
+% Concerning the short term prediction of dynamics, we found that on average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.
+% Moreover, the average frequency trajectory in figure \ref{fig:trajectories} shows no ``inertia'' when going from $t<0$ to $t>0$, suggesting that selective sweeps are not representative of typical trajectories.
+% This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+% These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
+
+Methods for predicting the future evolution of influenza either construct explicit fitness models \citep{luksza_predictive_2014,huddleston_integrating_2020}, use historical patterns of evolution \citep{luksza_predictive_2014,bush_predicting_1999}, phenotypic assays \citep{neher_prediction_2016,steinbruck_inference_2012}, or dynamic or phylogenetic patterns \citep{neher_predicting_2014,klingen_sweep_2018}.
 The goal of these methods is to pick strains that are good representatives of future populations and could serve as vaccine candidates \citep{morris2018predictive}.
@@
 The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work. 
-\pierre{I think ``low power'' is okay here?
-Not an overstatement.}
-Picking representatives of the future and predicting frequency dynamics are distinct objectives and that success at the former (as compared to random picks) is consistent with a lack of predictable dynamics.
-The strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one -- consistent with the result in \citep{neher_predicting_2014} -- despite the fact future frequencies are not predicted by the LBI.
-In fact, the very simple method of taking the consensus sequence of all present strains performs slightly but consistently better than picking the strain with the LBI. 
-The consensus sequence is the best possible predictor for a neutrally evolving population, and does not attempt to model fitness in any way. 
-While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, the reason it is predictive for influenza, does not seem to due to an ability to measure fitness of influenza viruses from genealogical structure.
+Picking representatives of the future and predicting frequency dynamics are distinct objectives and success at the former (as compared to random picks) is not necessarily inconsistent with a lack of predictable dynamics.
+In fact, \citep{huddleston_integrating_2020} reports that the rate at which the frequency of a strain changes is often a poor predictor -- consistent with our observations here.
+But despite the fact that future frequencies are not predicted by the LBI, the strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one.
+While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, it does not seem to be predict influenza evolution because it measures fitness from genealogical structure.
 Instead, we believe it picks closer than average strains simply because it has the tendency to be maximal at the base of large and dense clades. 
 These basal genotypes are closer to the future populations than the current tips of the tree and hence a better predictor on average.
+The consensus sequence of all present strains performs slightly but consistently better than picking the strain with the highest LBI. 
+The consensus sequence is the best possible predictor for a neutrally evolving population, and does not attempt to model fitness in any way.
 
@@
 Many processes other than selection, including seasonality and spatio-temporal heterogeneity, can generate effective long tailed offspring distributions even in absence of bona-fide fitness differences, which might explain ladder-like non-Kingman phylogenetic trees.
+
+A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
+Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
+Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population.
+These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.
 
@@
 	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}. 
-	Strain names and accession numbers are given as tables in two supplementary files. \\
+	Strain names and accession numbers are given as tables in two supplementary files. Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/src/config} were removed. \\
 	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}. 
+	% https://github.com/nextstrain/seasonal-flu/tree/081be2f487182252627418491be36cd01d645954/config
 % subsection data (end)
@@
 	In order to avoid estimates of frequencies that are too noisy, we only keep trajectories that are based on a population of at least 10 sequences for \emph{each} time bin.
-As said in the Results section, we also restrict the analysis to trajectories that begin at a $0.$ frequency, in part to avoid double counting.
+As said in the Results section, we also restrict the analysis to trajectories that begin at a $0$ frequency, in part to avoid double counting.
 We find a total of 460 such trajectories.
@@
 	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
-	Region used are the ones defined in the \texttt{nextstrain} tool~\cite{10.1093/bioinformatics/bty407}.
+	Region used are the ones defined in the \texttt{Nextstrain} tool~\cite{10.1093/bioinformatics/bty407}.
 Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
@@
 
-	In order to measure the distance of several predictor sequences to the future population, we rely on the \emph{EarthMover's Distance} (EMD), a metric commonly applied in machinelearning to compare collections of pixels or words \cite{710701,10.5555/3045118.3045221}.
+	In order to measure the distance of several predictor sequences to the future population, we rely on the \emph{Earth Mover's Distance} (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{710701,10.5555/3045118.3045221}.
 Here, we apply it to compute the distance between the sequences of two populations, noted as $\mathcal{X} = \left\{(x^n, p^n)\right\}$ and $\mathcal{Y}=\left\{(y^m,q^m)\right\}$ with $n\in\{1\ldots N\}$ and $m\in\{1\ldots M\}$.
@@
 % section methods (end)
+
+\section{Funding}
+This work was funded in part by NIAID (R01AI127893-01) and the SNF (310030\_188547).
 
@@
 % \newpage
-% \section{A/H1N1 influenza}
+% \section{A/H1N1pdm influenza}
+\newpage
+\subsection{Probability of fixation in single locus model of evolution}
+\label{sub:probability_of_fixation}
+
+In \citep{kimura_diffusion_1964}, Kimura investigates a simple model of evolution with a single locus and a population of size $N$. 
+In this framework, a mutation at this locus with fitness effect $s$ and observed at frequency $f$ has the following probability of fixation: 
+\begin{equation}
+    P_{fix}(f\vert s,N) = \frac{1-e^{-sNf}}{1-e^{-sN}}.
+\end{equation}
+Expanding this formula for $sN\ll 1$, that is in the weak selection regime, yields at the first order
+\begin{equation}
+    P_{fix}(f\vert s,N) = f + f(1-f)\frac{sN}{2}.
+    \label{eq:pfix_first_order}
+\end{equation}
+Equation~\ref{eq:pfix_first_order} tells us two things. First, when the mutation is neutral, that is $s=0$, we have $P_{fix}(f) = f$. 
+This naturally confirms the result obtained for a neutral model of evolution.
+Seconds, when $sN\neq 0$, we can expect deviations from the diagonal in a $P_{fix}$ against $f$ plot. 
+The sign of these deviations is determined by the sign of $s$, with beneficial mutations being found above diagonal while deleterious one are found below. 
+The amplitude of these deviations depends on the strength of selection $sN$, as well as on the frequency through the $f(1-f)$ term, making them larger for $f\sim 0.5$.
 
@@
 		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_year.png}
-		\caption{Supplementary figure 1 - Number of A/H3N2 HA sequences per year from year 1990. }
+		\caption{Number of A/H3N2 HA sequences per year from year 1990. }
 		\label{fig:nseq_per_year}
@@
 		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_month.png}
-		\caption{Supplementary figure 2 - Number of H3N@ HA and NA sequences per month from year 2000. }
+		\caption{Number of H3N2 HA and NA sequences per month from year 2000. }
 		\label{fig:nseq_per_month}
@@
 		\includegraphics[width=0.95\textwidth]{SM_figures/sample_trajectories.png}
-		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the A/H3N2 HA protein.}
+		\caption{Frequency trajectories for the 9 most entropic positions in the A/H3N2 HA protein.}
 		\label{fig:sample_trajectories}
@@
 		\centering
+		\includegraphics[width=0.9\textwidth]{./SM_figures/ShortTerm_h3n2_varf0.png}
+		\caption{Equivalent to panel \textbf{B} of figure \ref{fig:trajectories} of the main text for A/H3N2, with $f_0$ equal 0.5 in \textbf{A} (76 trajectories), and  0.7 in \textbf{B} (63 trajectories).}
+		\label{fig:trajectories_h3n2_varf0}
+	\end{figure*}
+
+	\begin{figure*}
+		\centering
+		\includegraphics[width=0.9\textwidth]{./SM_figures/ShortTerm_h1n1.png}
+		\caption{Equivalent to panels \textbf{B} and \textbf{C} of figure \ref{fig:trajectories} of the main text for A/H1N1pdm influenza.
+$89$ trajectories are shown and participate to the mean (thick black line).
+}
+		\label{fig:trajectories_h1n1}
+	\end{figure*}
+
+	\begin{figure*}
+		\centering
+		\includegraphics[width=0.9\textwidth]{./SM_figures/ShortTerm_h1n1_varf0.png}
+		\caption{Equivalent to panel \textbf{B} of figure \ref{fig:trajectories} of the main text for A/H1N1pdm, with $f_0$ equal 0.5 in \textbf{A} (50 trajectories), and  0.7 in \textbf{B} (41 trajectories).}
+		\label{fig:trajectories_h1n1_varf0}
+	\end{figure*}
+
+	\begin{figure*}
+		\centering
 		\includegraphics[width=0.9\textwidth]{./SM_figures/pfix_v_freq_strat_h1n1pdm.png}
-		\caption{Equivalent of figure \ref{fig:pfix_stratified} of the main text for the HA gene of A/H1N1 influenza.
+		\caption{Equivalent of figure \ref{fig:pfix_stratified} of the main text for the HA gene of A/H1N1pdm influenza.
 Fixation probability $P_{fix}(f)$ as a function of frequency.
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time_h3n2.png}
-		\caption{Supplementary figure 4 - Based on A/H3N2 HA and NA.
+		\caption{Based on A/H3N2 HA and NA.
 \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time_h1n1pdm.png}
-		\caption{Supplementary figure 4 - Based on A/H1N1 HA and NA.
+		\caption{Based on A/H1N1pdm HA and NA.
 \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/GeoSpread_v_time_ha.png}
-		\caption{Supplementary figure 6 - Geographic spread of mutations as a function of the time for which they have been present in the population above a frequency of 5\%.
+		\caption{Geographic spread of mutations as a function of the time for which they have been present in the population above a frequency of 5\%.
 Points represent individual mutations and for a population in a given time bin.
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/EMD_to_future.png}
-		\caption{Supplementary figure 7 - Earthmover's distance to the future population for different predictors.
+		\caption{Earth mover's distance to the future population for different predictors.
 A present population consists of all A/H3N2 HA sequences sampled in a 4 months time window.

### 2020-11-17 article_2907.tex
Submitted

--- 2020-07-06 article.tex
@@
 
-	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
+	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1 influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
 	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
 	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
-	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
+	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
 	This observation holds for many different categories of mutations, including mutations at epitope positions. 
-	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
+	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
 	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
@@
 
-The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
+The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
 \\
@@
 
-In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
-This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year.
+In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 (since the year 2000) and A/H1N1 (since the year 2001) influenza available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors).
+This amounts to $44\,976$ HA and $36\,300$ NA sequences for A/H3N2 and $45\,350$ HA and $40\,412$ NA sequences for A/H1N1, with a minimum of 100 per year.
 These sequences are binned in non-overlapping intervals of one month.
-Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date.
+Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the influenza population at a given date.
 The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).
@@
 
-For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. 
+For $f_0=0.3$, we find $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$.
+The same analysis is performed for A/H1N1, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
 Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
@@
 The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
-This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$.
-The fact that this average rose $t<0$ has hence no information for $t>0$ and is due to the conditions by which these trajectories were selected.
+This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1.
+Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
+This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to reflect in the average trajectory.
 
@@
 The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
-Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
-\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
-This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
+Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
+For both proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, \discuss{that is $P_{fix}(f)\simeq f$.} 
+This result is exactly what is expected in a population evolving in the \emph{absence} of or under very weak selection. 
 Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$.
@@
 Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
-Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
-\discuss{Add smth for A/H1N1}
+Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1 show only small deviations from it.
+In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutations is observed.
 
-\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
-If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
-In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}
+This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured. 
+In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
 
-To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
+To investigate further, we try to find categories of mutations that deviate from the diagonal of panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}. 
 We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
@@
 Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
-Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.
+Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.
 
-Next, we focused on previously reported antigenic sites in the HA protein, referred to as \emph{epitope} positions. 
+Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions. 
 Mutations at these position might mediate immune escape and are therefore likely under strong selection. 
@@
 Two ways of categorising mutations, however, suggest some power to predict fixation. 
-In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into binary trajectories where only two amino acid variants co-circulate and non-binary ones with more than two variants. 
+In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants.
 Novel variants at non-binary positions, \emph{i.e.} ones for which competition between three amino acids or more has occurred at least once, have a higher chance of fixation. 
@@
 
-\discuss{We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
+We conduct the same analysis on A/H1N1 influenza, with results visible in figure \sref{fig:pfix_stratified_h1n1}. 
 Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation. 
@@
 However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
-Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.}
+Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.\\
 
@@
 
-\discuss{The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.}
-One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}.
+The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.
+One possible explanation for the weakly predictable behaviour of mutations (beyond current frequency) might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}.
 We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}. 
@@
 In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal. 
-However, even in an extremely fast changing fitness landscape with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ still significantly differs from $f$.
+However, it takes an extremely fast changing fitness landscape to push $P_{fix}$ close to the diagonal: with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ differs from $f$ in a way that is comparable to what is observed in A/H1N1 influenza.
 
 These models are not meant to be accurate models of influenza viruses evolution. 
-\discuss{But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are not readily reproduced by in models of adapting populations, even if they include strong clonal competition and epistasis.
-We conclude that the pattern in figure~\ref{fig:fix_loss_active} is not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the quasi-neutral yet strongly selected evolutionary dynamics of IAVs.}
+But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
+We conclude that the pattern in figure~\ref{fig:fix_loss_active} may not a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.
 
 \subsection*{Why do predictions work?}
-The quasi-neutral statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
+The statistics of frequency trajectories seems to be in conflict with the notion that influenza evolution is predictable.
 Likewise, the LBI, a quantity that correlates with fitness in mathematical models and is used to predict future influenza populations \citep{neher_predicting_2014}, does not seem to contain any information on whether a specific mutation is going to fix or not, see figure~\ref{fig:pfix_stratified}.
@@
 To assess the ability of the LBI to pick a close representative of the future, we compute the LBI of each node of one time bin in the tree using only the leaves that belong to that time bin. 
-The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for randomly chosen strain.
+The top panel in figure~\ref{fig:LBI_consensus} shows the hamming distance of the strain with the highest LBI to future populations at different $\Delta t$ along with the same distance for a randomly chosen strain.
 The figure shows the distance averaged over all possible values of $t$ for $\Delta t$ between $0$ and $32$ months, giving us an average efficiency of a predictor over 16 years of influenza evolution.
@@
 We hence reproduce previous results showing that the LBI picks closer than average representatives \citep{neher_predicting_2014}.
-To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different measure: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence).
+To investigate whether this apparent success is due to the ability of the LBI to predict fitness or not, we explored a different predictor: the amino acid consensus sequence of the present population (see Methods for a definition of the consensus sequence).
 The choice is motivated by the fact that it can be shown to be the best possible long term predictor for a neutrally evolving population in terms of Hamming distance (see SM section \ref{sub:consensus_sequence_as_a_predictor}).
 Figure~\ref{fig:LBI_consensus} shows that the consensus sequence is in fact a equally good or even slightly better representative of the future than the sequence with highest LBI 
-(note that this sequence does \emph{not} necessarily exist in the population).
+(note that the consensus sequence does \emph{not} necessarily exist in the population).
 
@@
 Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time. 
-However, we could find only limited evidence for such persistent behavior in the past 19 years of A/H3N2 evolution. \pierre{But we do find evidence for A/H1N1, not of a large magnitude.}
-This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
+However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution. 
+This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
 
 The primary quantity we investigated is the probability that a novel mutation fixes, given it is currently observed at frequency $f$.
-In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively \citep{kimura_diffusion_1964}.
-\discuss{This effect is readily seen in simulations of simple models of evolution, but we failed to identify groups of substitutions in which fixation probability deviates substantially from their instantaneous frequency.}
+In neutral models of evolution this probability is equal to $f$, while it should be higher or lower than $f$ for beneficial or deleterious mutations, respectively, an effect readily seen in simulations of simple models of evolution \citep{kimura_diffusion_1964}.
+However, in the case of influenza, the probability of fixation differs little from $f$, making the current frequency of a mutation the best predictor for fixation.
 In figure \ref{fig:pfix_stratified}, we tried to break trajectories into groups that where previously associated with rapid adaptation (e.g.~epitope positions as defined in \citep{Shih6283,Koel976,bush_predicting_1999,luksza_predictive_2014}) or high and low fitness \citep{neher_predicting_2014}.
-Only mutations at the seven ``Koel''-positions fixed more often than expected by chance.
-The only other stratification that had a measurable effect on fixation were was whether or not convergent evolution (at the level of amino acid or position) was observed. 
-On average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.
+Most of these methods did not result in an increased predictability: only mutations at the seven ``Koel''-positions fixed more often than expected by chance. 
+Other stratification schemes had a measurable effect on fixation, such as whether or not convergent evolution (at the level of amino acid or position) was observed, or whether the position had seen competition between more than two amino acids in the past.
 
-\discuss{This failure to meaningfully predict mutation frequency trajectories} is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+Concerning the short term prediction of dynamics, we found that on average, the direction of a frequency trajectory does not persist for longer than a few months and memory of the initial frequency is lost after one year.
+Moreover, the average frequency trajectory in figure \ref{fig:trajectories} shows no ``inertia'' when going from $t<0$ to $t>0$, suggesting that selective sweeps are representative of typical trajectories.
+
+This failure to meaningfully predict mutation frequency trajectories is seemingly at odds with the strong signatures of selection in influenza surface proteins \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
 These signatures clearly indicate that surface proteins change their amino acid sequence to evade human immunity and these pressures can be recapitulated both in vivo and in vitro \citep{lee_mapping_2019,xue_within-host_nodate}.
-A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
+\pierre{Maybe move the rest of this paragraph at the end?} A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
 Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
@@
 The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work. 
-\pierre{I think ``low power'' is okay here?
-Not an overstatement.}
-Picking representatives of the future and predicting frequency dynamics are distinct objectives and that success at the former (as compared to random picks) is consistent with a lack of predictable dynamics.
+Picking representatives of the future and predicting frequency dynamics are distinct objectives and that success at the former (as compared to random picks) is not inconsistent with a lack of predictable dynamics.
 The strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one -- consistent with the result in \citep{neher_predicting_2014} -- despite the fact future frequencies are not predicted by the LBI.
@@
 		\includegraphics[width=0.7\textwidth]{SM_figures/Nseq_per_month.png}
-		\caption{Supplementary figure 2 - Number of H3N@ HA and NA sequences per month from year 2000. }
+		\caption{Number of H3N@ HA and NA sequences per month from year 2000. }
 		\label{fig:nseq_per_month}
@@
 		\includegraphics[width=0.95\textwidth]{SM_figures/sample_trajectories.png}
-		\caption{Supplementary figure 3 - Frequency trajectories for the 9 most entropic positions in the A/H3N2 HA protein.}
+		\caption{Frequency trajectories for the 9 most entropic positions in the A/H3N2 HA protein.}
 		\label{fig:sample_trajectories}
 	\end{figure}
+
+	\begin{figure*}
+		\centering
+		\includegraphics[width=0.9\textwidth]{./SM_figures/ShortTerm_h1n1.png}
+		\caption{Equivalent to panel \textbf{B} of figure \ref{fig:trajectories} of the main text for A/H1N1 influenza. $89$ trajectories are shown and participate to the mean (thick black line). }
+		\label{fig:trajectories_h1n1}
+	\end{figure*}
 
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/geospread_and_time_h1n1pdm.png}
-		\caption{Supplementary figure 4 - Based on A/H1N1 HA and NA.
+		\caption{Based on A/H1N1 HA and NA.
 \textbf{A}: Mutations with a higher or lower geographical spread, based on the median value of the score used (see Methods).
@@
 		\includegraphics[width=0.9\textwidth]{SM_figures/EMD_to_future.png}
-		\caption{Supplementary figure 7 - Earthmover's distance to the future population for different predictors.
+		\caption{Earthmover's distance to the future population for different predictors.
 A present population consists of all A/H3N2 HA sequences sampled in a 4 months time window.

### 2020-11-17 article_old.tex
Submitted

--- 2020-07-06 article.tex
@@
 If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
-In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
+In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e. the rapid rise and fixation. 
 If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.
@@
 
-\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
 If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
-In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}
+In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
 
@@
 However, this could be due to the shorter time period over which A/H1N1 evolved, resulting in a shorter tree and less possibilities of convergent evolution. 
-Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.}
+Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.
 
@@
 
-\discuss{The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.}
+The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the notion that seasonal influenza virus evolution is driven by rapid directed positive selection.
 One possible explanation for the quasi-neutral behaviour of mutations might be tight genetic linkage inside each protein and strong competition between different adaptive mutations \citep{neher_genetic_2011}.

### 2020-11-17 response_to_referees.tex
Template for response ; makedifftex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}

\begin{document}
\section*{Response to reviewers}

\subsection*{Editor}

\response{
Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper. 
}

\editor{The comments and recommendations from two expert reviewers are now available for your manuscript. These reviewers judged the reported discoveries to be of medium significance, and the potential scientific impact of your work to be medium/high. They found that the manuscript needs improvement in text and additional data analysis. Editors generally agree with their concerns and recommendations, which led to a designation of medium/high priority.

Note that many manuscripts receiving medium priority based on reviewer comments are not accepted by the Board of Editors. If appropriate, the board may invite a resubmission following the rejection, which is intended to enable you to improve the manuscript towards receiving a high or top priority. The authors should pay close attention to the detailed review comments and address each comment with significant improvements.}

\response{}

\subsection*{Reviewer: 1}

\refa{The manuscript by Barrat-Charlaix et al. discusses the problem of
predictability of mutations in seasonal influenza viruses. It presents
data analysis of frequency trajectories in H1N1 and H3N2 lineages and
simulations of different selection scenarios.
It is an interesting revisit to predictions in influenza, however I
have some major concerns about the analysis and formulation of the
conclusions that should be addressed:}

\refa{1. The prediction problem is defined for each mutation: based on the
tracked frequency trajectory, can the future (fixation, loss, or
polymorphism) of the mutation be predicted. Such formulation has been
previously proposed by Illingworth and Mustonen, (eg. Genetics 2011,
Plos Pathogens 2012). By averaging over all amino-acid substitutions,
the authors show that the fates of mutations are not determined by the
value of the starting frequency.}

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\refa{3. Going through the previous and cited literature, I think the authors should cite some of these works in a more careful way. Specifically, the very related work by Illingworth and Mustonen is not mentioned. The distribution $P_delta_t(f|f_0)$ has been at the core of
the method of Strelkowa and Lassig, 2012 (which is cited, but at other
parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}

\refa{It would help if the plots with red-green-blue lines were also distinguished by differing markers.}

\subsection*{Reviewer: 2}

\refb{The paper does a retrospective study of amino acid substitutions in seasonal Influenza to determine what properties of these substitutions could help predict their fate in the future. The authors find that future frequency trajectories are surprisingly unpredictable. Even predicting which mutations fix in the population is hard. The authors find that the current frequency of a mutation is the best predictor for the probability of fixation, which would be expected under neutrality but not in a model with selection. 
I appreciate this study, I think it will be of interest to many readers and it is generally well done and fairly easy to follow. 
}

\refb{1. The authors focus on one feature at a time (frequency, epitope status, LBI …). It is interesting to see that each of these is not very predictive of future frequency / prob of fixation. However, I think the obvious next step is to see whether a combination of many features could do a better job of predicting. I am not sure why the authors don’t try to fit a model that takes into account all information they have about sites (say, type of AA change, location in the gene, current frequency etc) and see if a ML model is able to make predictions. }

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
\response{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. }

\refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}

\end{document}

### 2020-11-18 article.tex
SM ClusteringTraj - pfix w John's scores

@@
 
+\subsection{Correcting for nested mutations} % (fold)
+\label{sub:correcting_for_nested_mutations}
+
+The analysis of the main text computes probabilities of fixation assuming that all trajectories are independent. 
+However, it is well-known that mutations in infleunza viruses are nested: they appear on backgrounds that already carry other mutations. 
+Since mutations appearing on the same genomes will jointly fix or disappear, many frequency trajectories are not independent but correlated. 
+In order to compensate for potential biases due to this effect, we attempted to cluster trajectories based on similarity in their strain composition. 
+Our aim is that two trajectories corresponding to mutations appearing mostly on the same genomes will be grouped in the same cluster. 
+We then conduct the same analysis as in the main text on a set of \emph{effectively independent} trajectories constructed by taking one trajectory from each cluster.
+
+In order to perform clustering, we define a distance between trajectories. 
+A frequency trajectory $X$ is characterized by a series of frequency values and a time interval $T$. 
+$f(t)$ for $t\in T$ corresponds to the frequency at which a given mutation $x$ appears in the population at date $t$, and we define $S(t)$ as the strains that carry mutation $x$ at date $t$. 
+With this notation, $f(t)$ is the ratio of the number of elements in $S(t)$ to the total number of strains at date $t$. 
+Let us now consider two frequency trajectories $X_1$ and $X_2$. \\
+We define the distance $d(X_1,X_2)$ between these two trajectories based on the average similarity of the strains $S_1$ and $S_2$ that compose them: 
+$$ d(X_1,X_2) = \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$ 
+where $T_1\cap T_2$ is the time interval where both trajectories are active, and $\vert\cdot\vert$ denotes the number of elements of a set. 
+The quantity summed corresponds to the Jaccard index between strains composing $X_1$ and $X_2$ at a given date. 
+It is 1 if the two trajectories share exactly the same strains for this date, and 0 if they share no strain at all.  
+This leads to the two following properties of $d$: 
+\begin{itemize}
+	\item if $d(X_1,X_2)=0$, then $X_1$ and $X_2$ represent the same frequency trajectory. The mutations $x_1$ and $x_2$ that they correpond to always appear on the same strains and are totally linked.
+	\item if $d(X_1,X_2)=1$, then $X_1$ and $X_2$ can be considered completely independent.
+This can be the case if the two trajectories do not occur at the same dates, *i.e.* $\vert T_1\cap T_2\vert=0$, or if their respective mutations are never present on the same genomes.
+\end{itemize}
+
+we attempt to reduce the potential statistical bias due to the nesting of trajectories by grouping them based on the above defined distance. 
+Given a set of trajectories $\{X\}$, we perform a decomposition of $\{X\}$ into disjoint clusters $C_1(d^\star)\cup\ldots\cup C_n(d^\star)=\{X\}$ where $d^\star$ is an arbitrary threshold distance. 
+Clusters are built in such a way that given two trajectories $X_i$ and $X_j$
+$$ d(X_i,X_j) \leq d^\star \Rightarrow \exists k: X_i,X_j\in C_k $$
+and
+$$ X_i\in C_k \Rightarrow \exists j\in C_k: d(X_i,X_j) \leq d^\star. $$
+These condition imply that the clusters formed are the minimal ones that guarantee that any two trajectories closer than the threshold distance $d^\star$ belong to the same cluster. 
+The number of clusters $n$ depends on the chosen value for $d^\star$.
+
+We compute clusters for different values of $d^\star$ for the case of the HA gene in A/H3N2. 
+The top panel of figure \sref{fig:cluster_examples} shows the number cluster $n$ as a function of $d^\star$. 
+In the $d^\star=0$ case, only trajectories that are exactly identical in terms of strain composition are clustered together. 
+In this case, our clustering amounts to counting mutations that appear on exactly the same strains as one, reducing the number of effective trajectories from 800 to slightly less than 700. 
+For higher values of $d^\star$, the number of clusters steadily goes down until it reaches 1 for $d^\star=1$, which is the maximum value of the distance $d(X_1,X_2)$. 
+The sharp drop in $n$ for $d^\star=0.5$ is explained by the high number of very short (typically one time point) and low frequency trajectories that share one out of two strains.
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{Rev_figures/clusters_examples.png}
+	\caption{\textbf{Top}: Left-axis: number $n$ of clusters as a function of the threshold distance $d^\star$.
+The total number of trajectories is shown as a flat orange line.
+Right-axis: average size of clusters as a function of $d^\star$.
+\textbf{Bottom}: Examples of clusters for four values of $d^\star$.
+The four clusters displayed are the ones to which mutation HA1:33R belongs to.
+}
+	\label{fig:cluster_examples}
+\end{figure}
+
+Since the choice of $d^\star$ is arbitrary and since no particular value can be chosen based on the number of clusters $n(d^\star)$, we decide to test our clustering strategy for five values, namely $d^\star\in\{0, 0.05, 0.1, 0.2, 0.49\}$.
+The bottom panel of figure \sref{fig:cluster_examples} shows examples of a cluster for the four non-zero values of $d^\star$. 
+The cluster displayed in each case is the one containing the mutation HA1:33R. 
+As $d^\star$ increases, more and more unlike trajectories are grouped together. 
+In the case $d^\star=0.49$, the cluster consists of 13 trajectories, 5 of which end up dying while the rest fix. 
+Since such a high value of $d^\star$ results in grouping trajectories that do not have the same fate (fixation or death), we decide to exclude it from the rest of the analysis, resulting in four remaining values $d^\star\in\{0, 0.05, 0.1, 0.2\}$.
+\\
+
+Once clustering is performed, we re-conduct the analysis of the main text on a set of effective trajectories. 
+This set is constructed by taking one trajectory at random from each cluster. 
+Effective trajectories are then considered independent from each other. 
+The left panel of figure \sref{fig:pfix_v_freq_clusters} shows the fixation probability of trajectories as a function of their frequency for different values of $d^\star$, for the HA gene of A/H3N2.
+The result obtained in panel A of figure \ref{fig:fix_loss_active} of the main text is also showed as a reference. 
+For the three lower values of $d^\star$, results do not differ from the one obtained in the main text, even though the number of trajectories in each frequency bin has dropped as can be seen in the right panel of figure \sref{fig:pfix_v_freq_clusters}.
+This indicates that grouping together trajectories that share most of their strains, and are thus very correlated, does not modify the computed fixation probability in any way. 
+For the higher value $d^\star=0.2$, fixation probability drops slightly across all frequency bins, suggesting that fixating trajectories tend to be grouped together more frequently. 
+However, this drop remains of limited amplitude.
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{Rev_figures/Pfix_v_freq_clustering_panel.png}
+	\caption{\textbf{Left}: Fixation probability of trajectories as a function of probability for four values of $d^\star$.
+The reference curve is the same as in panel A of figure \ref{fig:fix_loss_active} for the HA gene.
+For readability, error bars are only displayed for the reference curve.
+\textbf{Bottom}: Number of trajectories in each frequency bin corresponding to the left panel.}
+	\label{fig:pfix_v_freq_clusters}
+\end{figure}
+
+Overall, this analysis leads us to think that even though mutations in influenza may be nested, considering trajectories as independent does not result in strong statistical biases. 
+Indeed, clustering similar trajectories together does significantly modify results presented in the main text.
+
+% subsection correcting_for_nested_mutations (end)
+
 \subsection{Biases in frequency estimations} % (fold)

### 2020-11-19 response_to_referees.tex
Markers - draft for some responses

@@
 \newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
-\newcommand{\refa}[1]{\textbf{R1:} #1}
-\newcommand{\refb}[1]{\textbf{R2:} #1}
-\newcommand{\editor}[1]{\textbf{Editor:} #1}
+\newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
+\newcommand{\refb}[1]{\textbf{R2:} #1\vskip 5mm}
+\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
 %\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
-\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}
+\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
+\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}}
 
@@
 
+\response{This is a good suggestion, and we agree with reviewer 1 that our original manuscript lacked a method to disentangle nested mutations. \\
+We added a new section in the Supplementary Material where we attempt to cluster together trajectories of mutations that partly appear on the same strains. 
+Trajectories corresponding to mutations always or often appearing on the same strains are then counted as one "effective" trajectory. 
+However, this new way of counting mutations does not significantly change our results, and as a consequence we left the figures of main text unchanged.}
+
 \refa{1.2 Despite the more general formulation in the beginning, this
@@
 revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}
+
+\responsedraft{
+	\begin{itemize}
+		\item Illingworth\&Mustonen do not do prediction.
+They fit complete frequency trajectories using two models (no linkage and linkage), and find that only one of the two is able to deliver a good fit.
+The conclusion is that linkage is a dominant effect for shape of trajectories.
+		\item We could indeed make an attempt at modeling the initial shape of a frequency trajectory until the point where a prediction (fixation/loss) is made.
+However, we feel that a strong point of our approach is to be modeling-free.
+		\item Simulations of populations evolving without recombination show that even without modeling, the fixation probability of a rising trajectory is higher than its frequency at the time when the prediction is made.
+Our claim is only that observations in A/H3N2 influenza differ both from what is typically expected the theory of evolving populations and from simulations.
+In particular, we do claim not that the trajectories are intrisically unpredictable.
+	\end{itemize}
+}
 
@@
 \refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
-\response{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. }
+\responsedraft{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. 
+The answer is that yes it's expected that vaccine influence frequency trajectories to some extent, but it is very hard to measure in practice. 
+The above paper attempted to measure this, and did not find clear-cut results.}
 

### 2020-11-19 article.tex
Add markers -- Change Fig 1.

@@
 \usepackage{csvsimple}
-% \linenumbers
+\linenumbers
 
@@
 This suggests that selective sweeps are not representative of typical trajectories.
+\pierre{We could cite the Illingworth paper here.
+They do show that as well, in a different way.
+The fact many trajectories are not sweeps is expected when there is strong linkage.
+What cannot be easily explained by linkage only is the complete lack of inertia on average (at least I think?), and the $P_{fix}(f)=f$.
+}
 

### 2020-11-19 article.tex
Panels with John's scores

@@
 But despite the fact that future frequencies are not predicted by the LBI, the strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one.
-While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, it does not seem to be predict influenza evolution because it measures fitness from genealogical structure.
+While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, it does not seem to be predictive of influenza evolution because it measures fitness from genealogical structure.
 Instead, we believe it picks closer than average strains simply because it has the tendency to be maximal at the base of large and dense clades. 
@@
 
+\subsection{Ability of fitness models to predict fixation} % (fold)
+\label{sub:ability_of_fitness_models_to_predict_fixation}
+
+In a companion article \citep{huddleston_integrating_2020}, authors developed several models to predict the future HA population for A/H3N2 influenza. 
+The premise of these models is to assign a fitness score to the HA gene of each A/H3N2 strain in a given year, and to then apply deterministic evolutionary equations to obtain the future strain composition.
+The fitness scores are derived from several quantities that have been shown to be connected with viral fitness.
+
+Here, we assess the ability of four fitness scores computed in \citep{huddleston_integrating_2020} to predict fixation: mutational load and epitope antigenic novelty introduced in \citep{luksza_predictive_2014}, hemagglutination inhibition (HI) antigenic novelty \citep{neher_prediction_2016} and the previously mentioned Local Branching Index (LBI) \citep{neher_predicting_2014}.
+Figure \sref{fig:pfix_v_freq_indivjohnsmodels} shows fixation probability for HA mutations with fitness scores in the top or bottom half of the fitness distribution for the four fitness measures. 
+\pierre{Maybe use delta-frequency instead of epitope mutations?
+They behave weirdly (high fitness is worse), and delta-frequency uses more than the last time point in the trajectory, which may partly answer one point of Rev1.}
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{Rev_figures/pfix_v_freq_indivJohnsmodels.png}
+	\caption{Ability of fitness scores based on \textbf{A} LBI, \textbf{B} HI titer, \textbf{C} mutational load and \textbf{D} epitope antigenic novelty to predict fixation of mutation frequency trajectories in the HA gene of A/H3N2.
+Similar to figure \ref{fig:pfix_stratified} of the main text.}
+	\label{fig:pfix_v_freq_indivjohnsmodels}
+\end{figure}
+
+Next, we use two fitness scores based on linear combinations of the following quantities: mutational load + LBI and mutational load + HI antigenic novelty. 
+These two combinations were found in \citep{huddleston_integrating_2020} to be the best predictors of future population. 
+The coefficients of the linear combinations were trained each year using the five previous years of data, and here we use their average value over a 25 years training period.
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{Rev_figures/pfix_v_freq_combiJohnsmodels.png}
+	\caption{Ability of fitness scores based on two combinations of the four fitness scores mentioned above: \textbf{A} mutational load + LBI; \textbf{B} mutational load + HI titer.}
+	\label{fig:pfix_v_freq_combijohnsmodels}
+\end{figure}
+
+% subsection ability_of_fitness_models_to_predict_fixation (end)
+
 \subsection{Biases in frequency estimations} % (fold)

### 2020-12-01 response_to_referees.tex
Add markers -- Change Fig 1.
Panels with John's scores
Response draft to Rev1 1.2

@@
 \newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
+\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
 \newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
@@
 \responsedraft{
+	We thank reviewer 1 for pointing out the work of Illingworth\&Mustonen to us, which we were unaware of.
+However, we believe that the method used in their article is not applicable as is in our case.
+To explain our thinking, we first briefly summarize the approach taken in Illingworth\&Mustonen: the authors fit two kind of models to fully known frequency trajectories:
 	\begin{itemize}
-		\item Illingworth\&Mustonen do not do prediction.
-They fit complete frequency trajectories using two models (no linkage and linkage), and find that only one of the two is able to deliver a good fit.
-The conclusion is that linkage is a dominant effect for shape of trajectories.
-		\item We could indeed make an attempt at modeling the initial shape of a frequency trajectory until the point where a prediction (fixation/loss) is made.
-However, we feel that a strong point of our approach is to be modeling-free.
-		\item Simulations of populations evolving without recombination show that even without modeling, the fixation probability of a rising trajectory is higher than its frequency at the time when the prediction is made.
-Our claim is only that observations in A/H3N2 influenza differ both from what is typically expected the theory of evolving populations and from simulations.
-In particular, we do claim not that the trajectories are intrisically unpredictable.
+		\item An "unlinked" model, where the fitness effect of a mutation fully determines its ultimate fixation or loss.
+The only trajectories that the model can produce are upward or downward sweeps, for beneficial or deleterious mutations respectively.
+		\item A "linked" model, in which the fate of a mutation is determined both by its own fitness and by the fitness of other mutations present on the genomes it appears on.
+This model fits actual trajectories much more accurately, which is one of the main findings of the paper.
 	\end{itemize} 
-}
+	Interestingly, both models depend on the same parameters that have to be fitted, namely the fitness effects of individual mutations.
+Linkage between mutations in the second model is introduced by using the measured frequencies of appearance of two mutations on the same genome.
+\\
+	\pierre{What I'm trying to say below
+	\begin{itemize}
+		\item Illingworth\&Mustonen don't do prediction, so there is no clear way to apply their method to our article.
+		\item Fitting the initial part of the trajectory using their models doesn't make much sense. 
+		\item We want to stay model free: it's possible to find signs of selection without fitting (cf. H1N1/simulated data)
+		\item Even if we're doing something very simple, it's surprising to see neutral-like behaviour. 
+	\end{itemize}}
+	We first want to point out that this approach is fundamentally different to the one of our article.
+Since Illingworth\&Mustonen fit models to fully known trajectories, these models cannot meaningfully be used for prediction purposes.
+Indeed, fitness effects of individual mutations are fitted with knowledge of the fixation or loss of the given mutation, and of the frequency of other mutations in the case of the second model.
+This strongly contrasts with our approach which is to find patterns of predictability of the future behaviour of frequency trajectories using past data only: \emph{e.g.} if a frequency trajectory has risen from $0$ to a frequency $f>0$, what can be said of its future shape or of its fixation?
+Thus, methods in Illingworth\&Mustonen are designed to answer a different question than the one we are asking in our work.
+\\
+	A way of reconciling the two approaches would be for us to fit the initial part of a trajectory with either model, and to use this fit to predict its future behaviour.
+However, we feel that this idea is not very relevant to our article for two reasons.
+First, neither model proposed by Illingworth\&Mustonen is really practical in our setting.
+The "unlinked" model only generates sweep-like trajectories, which our results show to be not representative of the "typical" trajectory (in good agreement with Illingworth\&Mustonen).
+The "linked" model can fit more complex trajectories, \emph{e.g.} ones that rise rapidly but ultimately die.
+However, it crucially depends on the frequencies of joint appearance of pairs mutations on genomes to introduce linkage.
+These frequencies vary through time, allowing the \emph{effective} fitness of a mutation to vary, and a trajectory to rise and then fall.
+In our case, these frequencies would only be known for the past, making impossible any prediction of the model.
+\\
+	The second reason is that we believe that our model-free approach has intrinsic benefits.
+[Something saying that we'd prefer not to start fitting models to  data, because it's not the point.
+But I'm not sure how to formulate this.] \\
+	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future.
+However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us.
+The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material).
+This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario.
+Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest.
+\\}
 
@@
 \refa{It would help if the plots with red-green-blue lines were also distinguished by differing markers.}
+\response{We have now added markers in most figures.}
 
@@
 \refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}
+\responsedraft{We have reworked this figure to allow more space for panel B.}
 

### 2020-12-04 article.tex
Refer to sec3 of SI

@@
 
-%Frequency trajectories allow a simple formulation of the above-stated questions.
-The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$.
-On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes.
+Since segments in influenza do not recombine, trajectories defined as above are sometimes heavily correlated with each other. 
+Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior. 
+In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical. 
+This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis. 
+In section \ref{sub:correcting_for_nested_trajectories} of the SM, we attempt to correct for nested trajectories by clustering them based on the strains that compose them. 
+However, this does not significantly change the results presented below.
+
 In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
@@
 	This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.
+\pierre{This result is consistent with\citep{illingworth_components_2012}.
+They show that sweep-like trajectories are very bad fits to actual trajectories.}
 
@@
 Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it.
+\pierre{``limited but clear'' instead of ``small''?}
 In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.
@@
 
-\subsection{Correcting for nested mutations} % (fold)
-\label{sub:correcting_for_nested_mutations}
+\subsection{Correcting for nested trajectories} % (fold)
+\label{sub:correcting_for_nested_trajectories}
 
@@
 \textbf{Bottom}: Examples of clusters for four values of $d^\star$.
-The four clusters displayed are the ones to which mutation HA1:33R belongs to.
+The four clusters displayed are the ones to which the trajectory of mutation HA1:33R belongs.
 }

### 2020-12-06 response_to_referees.tex
add to response

@@
 
+\editor{The comments and recommendations from two expert reviewers are now available for your manuscript.
+These reviewers judged the reported discoveries to be of medium significance, and the potential scientific impact of your work to be medium/high.
+They found that the manuscript needs improvement in text and additional data analysis.
+Editors generally agree with their concerns and recommendations, which led to a designation of medium/high priority.
+}
+
 \response{
@@
 
-\editor{The comments and recommendations from two expert reviewers are now available for your manuscript.
-These reviewers judged the reported discoveries to be of medium significance, and the potential scientific impact of your work to be medium/high.
-They found that the manuscript needs improvement in text and additional data analysis.
-Editors generally agree with their concerns and recommendations, which led to a designation of medium/high priority.
-
-Note that many manuscripts receiving medium priority based on reviewer comments are not accepted by the Board of Editors.
+\editor{Note that many manuscripts receiving medium priority based on reviewer comments are not accepted by the Board of Editors.
 If appropriate, the board may invite a resubmission following the rejection, which is intended to enable you to improve the manuscript towards receiving a high or top priority.
@@
 
-\response{}
+\response{We have used the revision to include additional analysis and improve the presentation.
+We hope that you and the reviewers now feel that our manuscript makes a major contribution towards understanding what aspects of influenza virus evolution are predictable, and which ones are not.
+Given the major public health importance or influenza and the fact that influenza virus evolution is one of the very few systems where adaptive evolution can be observed in real time, we feel that our results are of very high relevance to the community.
+}
 
@@
 
+\response{We address the individual points by the reviewer below.}
+
 \refa{1. The prediction problem is defined for each mutation: based on the
@@
 
+\response{We thank the reviewer for pointing out the work by Illingworth and Mustonen, which is indeed very relevant and complementary to our approach.
+We discus their relation below. }
+
 \refa{1.1. Mutations in influenza are highly nested, with a substantial
@@
 
-\response{This is a good suggestion, and we agree with reviewer 1 that our original manuscript lacked a method to disentangle nested mutations. \\
-We added a new section in the Supplementary Material where we attempt to cluster together trajectories of mutations that partly appear on the same strains. 
-Trajectories corresponding to mutations always or often appearing on the same strains are then counted as one "effective" trajectory. 
-However, this new way of counting mutations does not significantly change our results, and as a consequence we left the figures of main text unchanged.}
+\response{
+The reviewer is correct that our original manuscript did not attempt to account for nested mutations and we agree that this should have been done. \\
+We added a new section in the Supplementary Material where investigate the extent to which nesting affects our results:
+We cluster trajectories of mutations that partly appear on the same strains to various degrees and trajectories corresponding to mutations always or often appearing on the same strains are then counted as one ``effective" trajectory.
+Our analysis shows that nesting of mutations does not change our results significantly.
+We therefore left the figures of main text unchanged and refer to the supplement for a discussion of this issue.
+}
 
@@
 revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}
+
+\response{
+	Illingworth and Mustonen (I\&M) presented a careful analysis to what extent trajectories can be fit using a model assuming a single fitness coefficient, or a model that differentiates different background fitnesses.
+	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does capture the majority of trajectories.
+	Beyond that, our approach and aim differ from I\&M.
+	We investigate a priori features of mutations that have information on whether the mutation fixes, not whether the trajectories can be parameterized by a model.
+	Fitting a model to a partial trajectory to predict fitness is not expected to work: I\&M showed that subsequent interference has a substantial effect and our results indicate that the speed at which the mutation rises does not predict fitness.
+
+	Furthermore, I\&M estimate fixing mutations to have a much higher intrinisic fitness than dying mutation while we show that very little such information is available a priori.
+	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
+	This is why we resort to a model free approach ask what a priori information can help to determine which mutations fix or die.
+}
 
@@
 
+\response{In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
+The LBI was among the highest performing models and we hence used this score in the original manuscript.
+But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston et al.
+These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
+We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.
+}
+
 \refa{3.
@@
 
+\response{We thank the reviewer for this assessment and the constructive suggestions below. }
+
 \refb{1.
@@
 
+\response{
+	A similar point was also brought up by reviewer 1.
+	In addition to the LBI, epitopes, age, and geographic distribution we now also include the fitness scores of the best performing models in a recent machine learning effort to predict the composition of future populations (as opposed to predicting fixation).
+	These models include antigenic, experimental, and sequence or phylogenetic information.
+	Interestingly, all these models do similarly poorly and with the exception of mutational load.
+	These results are now included as supplementary information and discussed.
+}
+
 \refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }
+
+\response{Whether 2A or 2B are perceived as similar or not probably depends on how big a deviation from the diagonal one expected a priori.
+We now discuss now highlight the differences in the text and discuss potential reasons for the different behavior.  }
 
@@
 
+\response{
+	Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
+	Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
+
+}
+
 \refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
+
 \responsedraft{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}.
@@
 The above paper attempted to measure this, and did not find clear-cut results.}
+
+\response{
+	The degree to which the vaccine drives evolution is a very interesting question!
+	In the past, global influenza vaccine uptake was unlikely sufficient to generate much pressure and fairly thorough mixing of variants across countries with very different vaccination rates have been used to argue against a strong effect.
+But this might have changed in recent years and the paper by Wen et al in Viruses, 2018 picks up some signal.
+	Overall, the case for strong effect is still weak to our knowledge.
+}
 

### 2020-12-07 article.tex
small edits in intro

@@
 
+	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
+	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
+	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
+	% Need more context/bridging
 	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1pdm influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations.
@@
 
-Since segments in influenza do not recombine, trajectories defined as above are sometimes heavily correlated with each other. 
-Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior. 
+Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
+%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
 In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
 This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis.
-In section \ref{sub:correcting_for_nested_trajectories} of the SM, we attempt to correct for nested trajectories by clustering them based on the strains that compose them. 
-However, this does not significantly change the results presented below.
+In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
+While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, this does not significantly change the results presented below.
 

### 2020-12-07 article.tex
Updated diff.tex

--- 2020-12-04 article.tex
@@
 These models are not meant to be accurate models of influenza viruses evolution. 
-But figure~\sref{fig:simulations} does show is that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
+But figure~\sref{fig:simulations} does show that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
 We conclude that the pattern in figure~\ref{fig:fix_loss_active} may not be a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.

### 2020-12-11 article.tex
Completed SI

@@
 The premise of these models is to assign a fitness score to the HA gene of each A/H3N2 strain in a given year, and to then apply deterministic evolutionary equations to obtain the future strain composition.
-The fitness scores are derived from several quantities that have been shown to be connected with viral fitness.
-
-Here, we assess the ability of four fitness scores computed in \citep{huddleston_integrating_2020} to predict fixation: mutational load and epitope antigenic novelty introduced in \citep{luksza_predictive_2014}, hemagglutination inhibition (HI) antigenic novelty \citep{neher_prediction_2016} and the previously mentioned Local Branching Index (LBI) \citep{neher_predicting_2014}.
+The fitness values are computed using several quantities that have been shown to be connected with viral fitness.
+Fitness depends either on a single or on two quantities, and coefficient(s) of each fitness model are trained by minimizing the earth movers distance between the observed strain population one year in the future and the estimated population produced by the model.
+The training is performed using HA sequences from 1990 to 2015, and we consider here the models obtained by taking the average value of coefficients obtained for every year.
+
+We first assess the ability of the four most performing fitness models based on individual scores in \citep{huddleston_integrating_2020} to predict fixation: mutational load \citep{luksza_predictive_2014}, hemagglutination inhibition (HI) antigenic novelty \citep{neher_prediction_2016}, a ``delta frequency''score based on the recent increase in frequency of clades, and the previously mentioned Local Branching Index (LBI) \citep{neher_predicting_2014}.
 Figure \sref{fig:pfix_v_freq_indivjohnsmodels} shows fixation probability for HA mutations with fitness scores in the top or bottom half of the fitness distribution for the four fitness measures.
-\pierre{Maybe use delta-frequency instead of epitope mutations?
-They behave weirdly (high fitness is worse), and delta-frequency uses more than the last time point in the trajectory, which may partly answer one point of Rev1.}
+
+The best performing individual score (and second overall) in \citep{huddleston_integrating_2020}, LBI, does not provide any information about fixation (figure \sref{fig:pfix_v_freq_indivjohnsmodels}\textbf{A}).
+This is consistent with the result found in figure \ref{fig:pfix_stratified} of the main text. 
+In panels \textbf{B} and \textbf{D}, we observe the same inability to predict fixation for the fitness models based on HI titer and delta frequency, which were respectively the second and third best individual scores (fifth and sixth overall) in \citep{huddleston_integrating_2020}.
+However, the score based on mutational load (panel \textbf{C}) does show a more significant predicting power, with higher fitness mutations having a $\sim 25$\% higher chance of fixing than lower fitness ones for three frequency bins in a row.
+However, this effect vanishes for higher frequencies, with both fitter and less fit mutations having the same chance to fix.
 
@@
 	\includegraphics[scale=0.5]{Rev_figures/pfix_v_freq_indivJohnsmodels.png}
-	\caption{Ability of fitness scores based on \textbf{A} LBI, \textbf{B} HI titer, \textbf{C} mutational load and \textbf{D} epitope antigenic novelty to predict fixation of mutation frequency trajectories in the HA gene of A/H3N2.
+	\caption{Ability of fitness scores based on \textbf{A} LBI, \textbf{B} HI titer, \textbf{C} mutational load and \textbf{D} delta frequency, to predict fixation of mutation frequency trajectories in the HA gene of A/H3N2.
 Similar to figure \ref{fig:pfix_stratified} of the main text.}
@@
 
-Next, we use two fitness scores based on linear combinations of the following quantities: mutational load + LBI and mutational load + HI antigenic novelty.
-These two combinations were found in \citep{huddleston_integrating_2020} to be the best predictors of future population.
-The coefficients of the linear combinations were trained each year using the five previous years of data, and here we use their average value over a 25 years training period.
+Next, we use two fitness models based on linear combinations of the following quantities: mutational load + LBI and mutational load + HI antigenic novelty.
+These composite models were the first and third best performing ones in \citep{huddleston_integrating_2020}. 
+Figure \sref{fig:pfix_v_freq_combijohnsmodels} again shows fixation probability for HA mutations with fitness scores above or below the median fitness value.
+
+Both models provide some information about fixation, with the mutational load + HI model having more success for frequencies around $0.5$. 
+This observation could be explained by the fact that fitness based on HI titers alone showed some power in this same frequency range. 
+However, both models do not seem to perform significantly better than the one based on mutational load alone. 
+This is not surprising, as LBI and HI titer did not perform well as individual models.
 
@@
 \end{figure}
+
+Combining models in \citep{huddleston_integrating_2020} typically achieved a greater performance than using single models. 
+This was the case for the mutational load + LBI model, combining the second and 7th best predictors overall to obtain the best one, and for the mutational load + HI titer model that combined the 4th and 7th best predictors to obtain the third best one.
+However, our results indicate that the gain in capability of predicting the future strain composition obtained by combining models is not immediately transferable to the exercise of predicting fixation.
+
+It is important to note that the prediction targets in \citep{huddleston_integrating_2020} and in the present article are not identical. 
+It is shown in figure \ref{fig:LBI_consensus} of the main text that even though LBI performs well when predicting the future strain composition, it may not be because it is indicative of fitness.
 

### 2020-12-11 article.tex
edit manuscript and include interference references in places

@@
 	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
-	% Need more context/bridging
-	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1pdm influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations.
+	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
+	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
 	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
 	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
-	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
+	To our surprise, we find that the predictability of these trajectories is very limited: For A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
 	This observation holds for many different categories of mutations, including mutations at epitope positions.
@@
 In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
-While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, this does not significantly change the results presented below.
-
-In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
+While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.
+
+We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
 These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
@@
 Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
-If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
-In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation.
+%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
+In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
 The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
@@
 If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
+Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a consistent tendency to rise in frequency (see below).
 
@@
 The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C.
-Initially, \emph{i.e.} at time $t_0=0$, this distribution is by construction peaked around $f_0$.
+Initially, \emph{i.e.}, at time $t_0=0$, this distribution is by construction peaked around $f_0$.
 If a large fraction of the trajectories keep increasing after  this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.
@@
 Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation.
-Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it.
-\pierre{``limited but clear'' instead of ``small''?}
-In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.
+Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
+For A/H1N1pdm, we find that new mutations fix moderatly but consistently more often then their current frequency.
+In both cases, frequency is a strong predictor of fixation.
 
@@
 
-Two ways of categorising mutations, however, suggest some power to predict fixation.
+In supplementary figure \ref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
+While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.
+
+In addition to mutational load, we find that parallel evolution is predictive of fixation.
 In panel Fig.~\ref{fig:pfix_stratified}\textbf{C}, we split trajectories into those occurring at binary positions where only two amino acid variants co-circulate and non-binary positions with more than two variants.
@@
 One possible explanation for the weakly predictable behaviour of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
+Indeed, \citet{illingworth_components_2012} have shown that background fitness and interference have substantial effects on frequyency trajectories.
+
+Nevertheless, we expect that signatures of intrinsic fitness benefits remain.
 We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}.

### 2020-12-15 response_to_referees.tex
Remove response drafts

@@
 \newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
-\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}}
+\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}
 
@@
 
-\responsedraft{
-	We thank reviewer 1 for pointing out the work of Illingworth\&Mustonen to us, which we were unaware of.
+% \responsedraft{
+% 	We thank reviewer 1 for pointing out the work of Illingworth\&Mustonen to us, which we were unaware of.
 However, we believe that the method used in their article is not applicable as is in our case.
 To explain our thinking, we first briefly summarize the approach taken in Illingworth\&Mustonen: the authors fit two kind of models to fully known frequency trajectories:
-	\begin{itemize}
-		\item An "unlinked" model, where the fitness effect of a mutation fully determines its ultimate fixation or loss.
+% 	\begin{itemize}
+% 		\item An "unlinked" model, where the fitness effect of a mutation fully determines its ultimate fixation or loss.
 The only trajectories that the model can produce are upward or downward sweeps, for beneficial or deleterious mutations respectively.
-		\item A "linked" model, in which the fate of a mutation is determined both by its own fitness and by the fitness of other mutations present on the genomes it appears on.
+% 		\item A "linked" model, in which the fate of a mutation is determined both by its own fitness and by the fitness of other mutations present on the genomes it appears on.
 This model fits actual trajectories much more accurately, which is one of the main findings of the paper.
-	\end{itemize}
-	Interestingly, both models depend on the same parameters that have to be fitted, namely the fitness effects of individual mutations.
+% 	\end{itemize}
+% 	Interestingly, both models depend on the same parameters that have to be fitted, namely the fitness effects of individual mutations.
 Linkage between mutations in the second model is introduced by using the measured frequencies of appearance of two mutations on the same genome.
 \\
-	\pierre{What I'm trying to say below
-	\begin{itemize}
-		\item Illingworth\&Mustonen don't do prediction, so there is no clear way to apply their method to our article.
-		\item Fitting the initial part of the trajectory using their models doesn't make much sense.
-		\item We want to stay model free: it's possible to find signs of selection without fitting (cf. H1N1/simulated data)
-		\item Even if we're doing something very simple, it's surprising to see neutral-like behaviour.
-	\end{itemize}}
-	We first want to point out that this approach is fundamentally different to the one of our article.
+% 	\pierre{What I'm trying to say below
+% 	\begin{itemize}
+% 		\item Illingworth\&Mustonen don't do prediction, so there is no clear way to apply their method to our article.
+% 		\item Fitting the initial part of the trajectory using their models doesn't make much sense.
+% 		\item We want to stay model free: it's possible to find signs of selection without fitting (cf. H1N1/simulated data)
+% 		\item Even if we're doing something very simple, it's surprising to see neutral-like behaviour.
+% 	\end{itemize}}
+% 	We first want to point out that this approach is fundamentally different to the one of our article.
 Since Illingworth\&Mustonen fit models to fully known trajectories, these models cannot meaningfully be used for prediction purposes.
@@
 \\
-	A way of reconciling the two approaches would be for us to fit the initial part of a trajectory with either model, and to use this fit to predict its future behaviour.
+% 	A way of reconciling the two approaches would be for us to fit the initial part of a trajectory with either model, and to use this fit to predict its future behaviour.
 However, we feel that this idea is not very relevant to our article for two reasons.
@@
 \\
-	The second reason is that we believe that our model-free approach has intrinsic benefits.
+% 	The second reason is that we believe that our model-free approach has intrinsic benefits.
 [Something saying that we'd prefer not to start fitting models to  data, because it's not the point.
 But I'm not sure how to formulate this.] \\
-	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future.
+% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future.
 However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us.
@@
 parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
+\responsedraft{Not (fully) addressed yet!
+}
 
@@
 \refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
-
-\responsedraft{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}.
-The answer is that yes it's expected that vaccine influence frequency trajectories to some extent, but it is very hard to measure in practice.
-The above paper attempted to measure this, and did not find clear-cut results.}
-
 \response{
@@
 \refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}
-\responsedraft{We have reworked this figure to allow more space for panel B.}
+\response{We have reworked this figure to allow more space for panel B.}
 

### 2020-12-15 article.tex
Typo

@@
 	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
-	To our surprise, we find that the predictability of these trajectories is very limited: For A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
+	To our surprise, we find that the predictability of these trajectories is very limited: for A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
 	This observation holds for many different categories of mutations, including mutations at epitope positions.
@@
 
-In supplementary figure \ref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
+In supplementary figure \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
 While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.

### 2020-12-16 article.tex
Add ref to Strelkowa & Lassig

@@
 If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
-Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a consistent tendency to rise in frequency (see below).
+Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below).
+\pierre{Is this ``see below'' necessary?
+I think that's what the reader naturally does!
+}
 
@@
 Having observed the frequency trajectory $f(t)$ of a mutation until a given date $t_0$, how much can we say about the future values of $f$ after $t_0$?
-We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f)$ of observing it at a value $f$ at time $t_0 + \Delta t$?
+We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f\vert f_0)$ of observing it at a value $f$ at time $t_0 + \Delta t$?
+This use of the frequency propagator to quantify the strength of selection was also proposed in \citep{strelkowa_clonal_2012}.
 

### 2020-12-16 article.tex
Downtone claims on p8 l47

@@
 However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
-This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) previous methods to predict influenza evolution work primarily because they pick strains that represent the future well, not because they predict future dynamics.
+This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) most methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
+This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
+% work primarily because they pick strains that represent the future well, not because they predict future dynamics.
 

### 2020-12-16 article.tex
More details/reference to new SM section

@@
 If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
-Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect a average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below).
+Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect an average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below).
 \pierre{Is this ``see below'' necessary?
@@
 This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.
-\pierre{This result is consistent with\citep{illingworth_components_2012}.
-They show that sweep-like trajectories are very bad fits to actual trajectories.}
 
@@
 
-This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
 If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured.
@@
 
-Next, we searched for features of mutations that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
+Next, we searched for features of mutations in A/H3N2 that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
 We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node.
@@
 
-In supplementary figure \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for several predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020} including models that use serological data.
-While many of these model predict future population composition, the only score that was predictive of fixation is mutations load.
+In supplementary figures \sref{fig:pfix_v_freq_indivjohnsmodels} and \sref{fig:pfix_v_freq_combijohnsmodels}, we repeat this analysis for the best predictors from our recent comparison of predictive models for influenza \citep{huddleston_integrating_2020}.
+We consider individual predictors based on LBI, mutational load, variation in frequency and serological data, as well as two combinations of these quantities. 
+While many of these models predict future population composition, the only score that was predictive of fixation is mutational load.
 
@@
 However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
-This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) most methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
+\pierre{We could change point (i) below to only mention A/H3N2. }
+This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
 This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
@@
 Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
-This suggests that selective sweeps are not representative of typical trajectories.
-\pierre{We could cite the Illingworth paper here.
-They do show that as well, in a different way.
-The fact many trajectories are not sweeps is expected when there is strong linkage.
-What cannot be easily explained by linkage only is the complete lack of inertia on average (at least I think?), and the $P_{fix}(f)=f$.
-}
+This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way.}
 
@@
 However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
-In figure~\ref{fig:pfix_stratified}, we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
+In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels} , we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
 Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.

### 2020-12-18 response_to_referees.tex
Work on responses

@@
 \response{We thank the reviewer for pointing out the work by Illingworth and Mustonen, which is indeed very relevant and complementary to our approach.
-We discus their relation below. }
+We discuss their relation below. }
 
@@
 The reviewer is correct that our original manuscript did not attempt to account for nested mutations and we agree that this should have been done. \\
-We added a new section in the Supplementary Material where investigate the extent to which nesting affects our results:
-We cluster trajectories of mutations that partly appear on the same strains to various degrees and trajectories corresponding to mutations always or often appearing on the same strains are then counted as one ``effective" trajectory.
+We added a new section in the Supplementary Material (``Correcting for nested trajectories'') where investigate the extent to which nesting affects our results.
+We cluster trajectories of mutations that partly appear on the same strains to various degrees, and trajectories corresponding to mutations always or often appearing on the same strains are then counted as one ``effective" trajectory.
 Our analysis shows that nesting of mutations does not change our results significantly.
@@
 	Illingworth and Mustonen (I\&M) presented a careful analysis to what extent trajectories can be fit using a model assuming a single fitness coefficient, or a model that differentiates different background fitnesses.
-	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does capture the majority of trajectories.
+	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does not capture the majority of trajectories.
 	Beyond that, our approach and aim differ from I\&M.
-	We investigate a priori features of mutations that have information on whether the mutation fixes, not whether the trajectories can be parameterized by a model.
+	We investigate \emph{a priori} features of trajectories that potentially have information on whether the mutation fixes, not whether the trajectories can be parametrized by a model.
 	Fitting a model to a partial trajectory to predict fitness is not expected to work: I\&M showed that subsequent interference has a substantial effect and our results indicate that the speed at which the mutation rises does not predict fitness.
 
-	Furthermore, I\&M estimate fixing mutations to have a much higher intrinisic fitness than dying mutation while we show that very little such information is available a priori.
+	Furthermore, I\&M estimate fixing mutations to have a much higher intrinsic fitness than dying mutations while we show that very little such information is available a priori.
 	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
-	This is why we resort to a model free approach ask what a priori information can help to determine which mutations fix or die.
+	This is why we resort to a model free approach to ask what a priori information can help to determine which mutations fix or die.
 }
-
+\responsedraft{Should we mention all the changes we made in the text? 
+We spend more time explaining that even though there is linkage, and even though we're looking at just the last point, we expect something above diagonal. 
+(Last paragraph before "Results", last paragraph before "Predicting future frequencies"), and first paragraph of "Simulations of models of adaptation"}
 % \responsedraft{
@@
 
-\response{In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
+\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.
+
+In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
 The LBI was among the highest performing models and we hence used this score in the original manuscript.
-But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston et al.
+But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material ("Ability of fitness models to predict fixation").
 These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
 We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.
+
+We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics, but they remain predictive of the composition of the future population.
+In consequence, we have toned down our conclusions about predictability.
+\pierre{I say that because I changed p8 l47.
+I have not touched p9 l48 because the new appendix shows it is a correct statement.}
 }
@@
 parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
-\responsedraft{Not (fully) addressed yet!
+\responsedraft{I cited (Strelkowa and Lassig) in the first paragraph of the section "Predicting future frequencies". 
+Did not cite it close to the simulation part, will do. 
+We cite Illingworth and Mustonen plenty of times now. 
 }
@@
 	These models include antigenic, experimental, and sequence or phylogenetic information.
-	Interestingly, all these models do similarly poorly and with the exception of mutational load.
+	Interestingly, all these models do similarly poorly with the exception of mutational load.
 	These results are now included as supplementary information and discussed.
@@
 \response{Whether 2A or 2B are perceived as similar or not probably depends on how big a deviation from the diagonal one expected a priori.
-We now discuss now highlight the differences in the text and discuss potential reasons for the different behavior.  }
+We now discuss now highlight the differences in the text and discuss potential reasons for the different behaviour.  }
 
@@
 	Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
-
 }
+\responsedraft{We should maybe make clear that this is not directly related to predictions for vaccines (at least it's not the aim). 
+That's John's paper. 
+The main implications are about our understanding of the evolutionary dynamics of influenza. 
+This could also help us to reply to the first comment: doing ML is more useful to predict the future population (for vaccines) than for predicting fixation of a single mutation. }
 

### 2020-12-18 response_to_referees.tex
Work on responses

@@
 }
-\responsedraft{Should we mention all the changes we made in the text? 
-We spend more time explaining that even though there is linkage, and even though we're looking at just the last point, we expect something above diagonal. 
-(Last paragraph before "Results", last paragraph before "Predicting future frequencies"), and first paragraph of "Simulations of models of adaptation"}
+
 % \responsedraft{
@@
 The LBI was among the highest performing models and we hence used this score in the original manuscript.
-But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material ("Ability of fitness models to predict fixation").
+But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston et al.
 These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.

### 2020-12-21 article.tex
small edits

@@
 In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
-This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis.
+This ``nesting'' of frequency trajectories may introduce biases when performing statistical analysis.
 In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
@@
 %If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
-In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
+In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., they rise rapidly and fix.
 The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
@@
 If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
-Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect an average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below).
-\pierre{Is this ``see below'' necessary?
-I think that's what the reader naturally does!
-}
+Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect that they rise in frequency on average.
 

### 2020-12-23 response_to_referees.tex
Small changes to response

--- 2020-12-18 response_to_referees.tex
@@
 parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
-\responsedraft{I cited (Strelkowa and Lassig) in the first paragraph of the section "Predicting future frequencies". 
-Did not cite it close to the simulation part, will do. 
+\responsedraft{I cited (Strelkowa and Lassig) in the first paragraph of the section "Predicting future frequencies" and in the second of ``Simulations''
 We cite Illingworth and Mustonen plenty of times now. 
@@
 	A similar point was also brought up by reviewer 1.
-	In addition to the LBI, epitopes, age, and geographic distribution we now also include the fitness scores of the best performing models in a recent machine learning effort to predict the composition of future populations (as opposed to predicting fixation).
+	In addition to the LBI, epitopes, age, and geographic distribution we now also include the fitness scores of the best performing models in a recent machine learning effort (Huddleston \emph{et.
+al.} 2020) to predict the composition of future populations (as opposed to predicting fixation).
 	These models include antigenic, experimental, and sequence or phylogenetic information.
 	Interestingly, all these models do similarly poorly with the exception of mutational load.
-	These results are now included as supplementary information and discussed.
+	These results are now included in a new section of the supplementary information and discussed.
 }
@@
 \response{Whether 2A or 2B are perceived as similar or not probably depends on how big a deviation from the diagonal one expected a priori.
-We now discuss now highlight the differences in the text and discuss potential reasons for the different behaviour.  }
+We now highlight the differences in the text and discuss potential reasons for the different behaviour.  }
 

### 2020-12-27 response_to_referees.tex
edits, shortening

@@
 We hope that you and the reviewers now feel that our manuscript makes a major contribution towards understanding what aspects of influenza virus evolution are predictable, and which ones are not.
-Given the major public health importance or influenza and the fact that influenza virus evolution is one of the very few systems where adaptive evolution can be observed in real time, we feel that our results are of very high relevance to the community.
+Given the major public health importance of influenza and the fact that influenza viruses are one of the very few systems where adaptive evolution can be observed in real time, we feel that our results are of very high relevance not only to the community studying influenza virus evolution but also those studying rapid adaptive evolution more generally.
 }
@@
 Our analysis shows that nesting of mutations does not change our results significantly.
-We therefore left the figures of main text unchanged and refer to the supplement for a discussion of this issue.
+We therefore left the analysis presented in the main text unchanged and refer to the supplement for a discussion of this issue.
 }
@@
 
-	Furthermore, I\&M estimate fixing mutations to have a much higher intrinsic fitness than dying mutations while we show that very little such information is available a priori.
+	Furthermore, I\&M estimate that fixing mutations have a much higher intrinsic fitness than dying mutations while we show that very little such information is available a priori.
 	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
 	This is why we resort to a model free approach to ask what a priori information can help to determine which mutations fix or die.
+
+    We also would like to point out that we see systematic differences between H3N2 and H1N1pdm.
+    While H1N1pdm shows a fixation behavior that is compatible with interfering selective sweeps, H3N2 does not.
+    H3N2 is typically regarded as the more rapidly adapting influenze lineage with the most "spindly" phylogeny.
+    We therefore do think that our observations are surprising to many people.
 }
@@
 
-We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics, but they remain predictive of the composition of the future population.
+We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics and fixation, but they remain predictive of the composition of the future population.
 In consequence, we have toned down our conclusions about predictability.

### 2020-12-27 article.tex
Add ref to Strelkowa&Laessig in simulations
edits, shortening

@@
 	\citet{luksza_predictive_2014}, for example, train a fitness model to capture antigenic drift and protein stability on patterns of epitope and non-epitope mutations.
-	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} predict fitness by using  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein.
+	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016,huddleston_integrating_2020} predict fitness by using  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein.
 	Finally, \citet{neher_predicting_2014} use branching patterns of HA phylogenies as a proxy for fitness.
@@
 Examples of trajectories can be seen in figure~\sref{fig:sample_trajectories} of the Supplement.
+We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
+These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
 
 Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
-%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
 In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
@@
 
-We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
-These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
-Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
-Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
-%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
-In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., they rise rapidly and fix.
-The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
-By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
-If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
-Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect that they rise in frequency on average.
+%Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
+%Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
+%Since influenza viruses are evading human immunity, we expect novel variants to be enriched for beneficial mutations.
+
+%In classic models of population genetics, strongly advantageous mutations undergo rapid selective \emph{sweeps}, i.e., they rise rapidly and fix.
+%In rapidly adapting population like influenza viruses, however, the sweep of a beneficial mutation is affected by the fitness of the genetic background it arose on and interference by other competing mutations \citep{strelkowa_clonal_2012}.
+%By considering the average behavior of variants that are rising in frequency, we effectively average over backgrounds, and obtain the beneficial on average.
+%If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
+%Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect that they rise in frequency on average.
 
@@
 We consider the idealized case sketched in panel \textbf{A} of figure~\ref{fig:trajectories}: given the trajectory of a \emph{new} mutation, \emph{i.e.} that started at a frequency of 0, and that we observe at frequency $f_0$ at time $t_0$, what is the probability $P_{\Delta t}(f\vert f_0)$ of observing it at a value $f$ at time $t_0 + \Delta t$?
-This use of the frequency propagator to quantify the strength of selection was also proposed in \citep{strelkowa_clonal_2012}.
+Such frequency propagators were originally proposed by \citep{strelkowa_clonal_2012} to quantify the strength of selection in influenza virus evolution.
 
@@
 
-While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is weakly informative.
+In contrast, the same analysis for A/H1N1pdm (Supplementary figure \sref{fig:trajectories_h1n1}) reveals a consistent upward trend of novel frequency trajectories.
+In A/H3N2, however, past dynamics alone is uninformative of future dynamics.
 
@@
 We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen.
-Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency.
+Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency for A/H3N2 and A/H1N1pdm, respectively.
 Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
@@
 	\caption{\textbf{A}: Activity of all rising frequency trajectories seen above 25\% frequency for A/H3N2 HA and NA.
-\textbf{B}: Same as \textbf{A} for A/H1N1.
+\textbf{B}: Same as \textbf{A} for A/H1N1pdm.
 \textbf{C}: Probability of fixation of a mutation (amino acid or synonymous) $P_{fix}(f)$ as a function of the frequency $f$ at which it is measured, for A/H3N2 HA and NA.
@@
 Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.
-For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$.
-This result is exactly what is expected in a population evolving in the \emph{absence} of selection.
+
+In the \emph{absence} of selection, we expect $P_{fix}(f)$ to equal $f$.
 % Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
@@
 Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation.
-For A/H1N1pdm, we find that new mutations fix moderatly but consistently more often then their current frequency.
-In both cases, frequency is a strong predictor of fixation.
+For A/H1N1pdm, we find that new amino acid mutations fix moderately but consistently more often then their current frequency, while synonymous mutations conform with the neutral expectation.
+In all cases, frequency is a strong predictor of fixation.
 
 This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
-If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured.
+If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are observed.
 In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
 
-Next, we searched for features of mutations in A/H3N2 that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}.
+Next, we searched for features of mutations in A/H3N2 that allow prediction of fixation beyond frequency.
 We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node.
@@
 Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution.
-Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.
+Both groups have indistinguishable probabilities of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.
 
 Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions.
-Mutations at these position might mediate immune escape and are therefore likely under strong selection and show sweep-like behavior.
+Mutations at these positions might mediate immune escape and might be expected to be under strong selection and show sweep-like behavior.
 We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}.
 Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes.
-Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance.
+Only mutations at the 7 pivotal sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance.
 No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
@@
 
-We conduct the same analysis on A/H1N1pdm influenza, with results shown in figure \sref{fig:pfix_stratified_h1n1}.
+We conducted the same analysis for A/H1N1pdm, with results shown in figure \sref{fig:pfix_stratified_h1n1}.
 Results are qualitatively similar to those obtained for A/H3N2, with LBI giving little information and mutations at non-binary positions having a higher chance of fixation.
-Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation in the latter case.
-However, this could be due to the shorter time period over which A/H1N1pdm evolved, resulting in a shorter tree and less possibilities of convergent evolution.
-Indeed, error bars for mutations appearing multiple times in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} are relatively large, indicating a lower amount of trajectories.\\
+Panel \textbf{D} differs between figures~\ref{fig:pfix_stratified} and \sref{fig:pfix_stratified_h1n1}, with convergent evolution giving less information on fixation for A/H1N1pdm.
+However, the number of recurring mutations for A/H1N1pdm in \textbf{D} of figure~\sref{fig:pfix_stratified_h1n1} is small and confidence intervals large. \\
 
@@
 However, it is important to note that sampling biases and heterogeneity across time and space (see supplementary figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month}) make answering such specific hypothesis challenging.
-Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non-trivial to measure.
+%Frequency of mutations might thus be amplified through different sampling biases, making the connection between geographic spread, seasonality and mutation frequency non-trivial to measure.
 %It is however interesting to observe that quantities based on the population's geographical and temporal structure can apparently give as much or more information on fixation probability and predictability than genomic or phylogenetic ones.
@@
 The results shown in figures~\ref{fig:fix_loss_active} and \ref{fig:pfix_stratified} are difficult to reconcile with the idea that seasonal influenza virus evolution is driven by rapid directed positive selection.
-One possible explanation for the weakly predictable behaviour of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
-Indeed, \citet{illingworth_components_2012} have shown that background fitness and interference have substantial effects on frequyency trajectories.
-
-Nevertheless, we expect that signatures of intrinsic fitness benefits remain.
-We design a simple model of population evolution based on the \texttt{ffpopsim} simulation software to test this hypothesis~\cite{10.1093/bioinformatics/bts633}.
-The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time.
+One possible explanation for the weakly predictable behavior of mutations (beyond their current frequency) might be tight genetic linkage inside each segment and strong competition between different adaptive mutations \citep{neher_genetic_2011,strelkowa_clonal_2012}.
+Indeed, \citet{illingworth_components_2012} have shown that background fitness and interference have substantial effects on frequency trajectories.
+
+To investigate the degree to which interference can explain the observations, we simulate a simple model of influenza evolution using \texttt{ffpopsim}~\cite{10.1093/bioinformatics/bts633}.
+The model represents a population of binary genomes of length $L=200$ evolving in a fitness landscape that changes through time. A similar setting was used in \citep{strelkowa_clonal_2012}.
 
@@
 In addition, we added a component that mimics immune selection.
-Every $\Delta t$ generation, we now introduce ``antibodies'' that target a specific sub-sequence of length $l=5$, noted $(x^{ab}_{i_1}, \ldots, x^{ab}_{i_l})$.
+Every $\Delta t$ generations, we introduce ``antibodies'' that target a specific sub-sequence of length $l=5$, noted $(x^{ab}_{i_1}, \ldots, x^{ab}_{i_l})$.
 The positions $(i_1\ldots i_l)$ are chosen at random, while the targeted sub-sequence is the dominant state at each position.
-Genomes that include the \emph{exact} sub-sequence targeted by the antibody suffer a strong fitness penalty.
-However, a single mutation away from that sub-sequence removes this penalty completely, resulting in a fitness landscape with very strong epistasis.
+Genomes that include the \emph{exact} sub-sequence targeted by the antibody suffer a strong fitness penalty. However, a single mutation away from that sub-sequence removes this penalty completely.
 This has the effect of triggering a strong competition between adaptive mutations: for a given antibody, $l=5$ possible mutations are now adaptive, but combinations of these mutations do not bring any fitness advantage.
@@
 Figure~\sref{fig:simulations} of the SM shows the $P_{fix}(f)$ as a function of $f$ for the two models and for different values of the inverse rate of change $\Delta t$ of the fitness landscape.
-For all models, this curve deviates significantly from the diagonal.
-This is most evident for the case of a simple additive fitness landscape that changes rarely $\Delta t=1000$: rising mutations almost always fix in the population, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent.
+Increasing interference reduces $P_{fix}(f)$ but even for very strong interference $P_{fix}(f)$ remains significantly larger than $f$.
+
+For a fairly stable simple additive fitness landscape ($\Delta t=1000$) rising mutations almost always fix, with $P_{fix}(f)\simeq 1$ for any $f$ larger than a few percent.
 This is corroborated by visual inspection of the trajectories, which shows that evolution in this regime is driven by regular selective sweeps that take a typical time of $\sim 400$ generations.
-In other regimes, with smaller $\Delta t$ or with strong epistatic competition, $P_{fix}(f)$ is reduced and closer to the diagonal.
+Smaller $\Delta t$ or strong epistatic competition reduce $P_{fix}(f)$.
 However, it takes an extremely fast changing fitness landscape to push $P_{fix}$ close to the diagonal: with $\Delta t=10$, that is about 40 changes to the fitness landscape in the time it would take a selective sweep to go from 0\% to fixation, $P_{fix}(f)$ differs from $f$ in a way that is comparable to what is observed in A/H1N1pdm influenza.
@@
 But figure~\sref{fig:simulations} does show that the patterns observed in influenza virus evolution are only reproduced by models of adapting populations when pushing clonal competition to extreme values.
-We conclude that the pattern in figure~\ref{fig:fix_loss_active} may not be a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.
+We conclude that the quasi-neutral pattern in figure~\ref{fig:fix_loss_active} may not be a straightforward manifestation of genetic linkage and clonal interference, but that some more intricate interplay of epidemiology, seasonality, human immunity and chance gives rise to the weakly predictable yet strongly selected evolutionary dynamics of IAVs.
 

### 2020-12-27 article.tex
edits

@@
 Under such conditions, it is expected that frequency trajectories will show a persistent behavior which would make them predictable for some time.
-However, we could find only limited evidence for such persistent behavior in the past 19 years of IAV evolution.
-\pierre{We could change point (i) below to only mention A/H3N2. }
-This lead us to conclude that (i) influenza virus evolution is qualitatively different from models of rapidly adapting population (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
+However, we could find little evidence for such persistent behavior in the past 19 years of A/H3N2 evolution.
+This lead us to conclude that (i) A/H3N2 influenza virus evolution is qualitatively different from models of rapidly adapting populations (despite clear evidence for frequent positive selection), and (ii) many methods to predict influenza evolution are unable to predict future frequency dynamics even though they work well for predicting the composition of the future population.
 This is illustrated by the case of LBI, which works primarily because it picks strains that represent the future well and not because it is predictive of future dynamics.
-% work primarily because they pick strains that represent the future well, not because they predict future dynamics.
+A/H1N1pdm, in contrast, shows stronger signatures of predictability.
 
@@
 Indeed, the average trajectory in figure \ref{fig:trajectories} takes a sharp turn when going from $t<0$ to $t>0$, instead of showing ``inertia''.
-This suggests that selective sweeps are not representative of typical trajectories. \pierre{We could cite the Illingworth paper here. They do show that as well, in a different way.}
+This suggests that selective sweeps are not representative of typical trajectories \citep{illingworth_components_2012}.
 
@@
 However, in the case of influenza, this probability differs little from $f$, making current frequency the best predictor for fixation.
-In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels} , we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
+In figures~\ref{fig:pfix_stratified} and~\sref{fig:pfix_v_freq_indivjohnsmodels}, we split trajectories into groups for which we expected $P_{fix}$ to deviate from $f$.
 Many of these splits, such as high/low LBI or epitope/non-epitope positions, did not result in an increased predictability, while others gave limited information on fixation.
-Despite the lack of predictability of mutation frequency trajectories, influenza surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
 
@@
 
-At the same time, influenza virus phylogenies show clear deviations from those expected from the neutral Kingman coalescent, similar to those expected under Bolthausen-Sznitman coalescent (BSC) processes that are generated by traveling wave models of rapid evolution \citep{neher_genealogies_2013,desai_genetic_2013}.
+At the same time, influenza virus surface proteins show strong signatures of selection \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+Their phylogenies show clear deviations from those expected from the neutral Kingman coalescent, similar to those expected under Bolthausen-Sznitman coalescent (BSC) processes that are generated by traveling wave models of rapid evolution \citep{neher_genealogies_2013,desai_genetic_2013}.
 The correspondence between the BSC and traveling wave models comes from transient exponential amplification of fit strains before these fitness differences are wiped out by further mutation.

### 2020-12-27 article.tex
edits to M&M and supp

@@
 	For a set of sequences in a given time bin, we compute frequencies of amino acids at each position by simple counting.
-We make the choice of not applying any smoothing method in an attempt to be as close to the data and ``model-less'' as possible.
+We make the choice of not applying any smoothing method in an attempt to be as close to the data and ``model-free'' as possible.
 This is especially important for the short term prediction of frequency trajectories, as estimations of the ``persistence time'' of a trajectory might be biased by a smoothing method.
@@
 \\
-	Note that the fact that we use samples of relatively small sizes -- at least for some time bins -- leads to biases in the estimation of frequencies.
+	Note that the fact that we use samples of relatively small sizes -- at least for some time bins -- can lead to biases in the estimation of frequencies.
 We show in Supplementary Material that these biases are generally small and do not induce any qualitative changes to results presented here.
@@
 
-	LBI was introduced in \cite{neher_predicting_2014} as an approximation of fitness in populations evolving under persistent selective pressure that is fully based on a phylogenetic tree.
+	LBI was introduced in \cite{neher_predicting_2014} as an approximation of fitness in populations evolving under persistent selective pressure.
+The LBI depends only on a phylogenetic tree.
 It relies on the intuition that the tree below high-fitness individuals will show dense branching events, whereas absence of branching is a sign of low-fitness individuals.
@@
 This means that only branches that ultimately lead to a leaf that belongs to the time bin are considered in the integration.\\
-	$\tau$ is the time scale for which the tree is informative of the fitness of a particular node.
+	The time scale $\tau$ parameterizes the distance over which the tree is informative of the fitness of a particular node.
 Here, we use a value of $\tau$ equal to a tenth of $T_C\simeq6$ years, the coalescence time for influenza A/H3N2 strains, converted to units of tree branch length through the average nucleotide substitution rate ($\simeq 4\cdot 10^{-3}$ substitutions per site per year for HA).
@@
 First, since no new mutations are allowed, the population at $t\gg T$ will be clonal, with all individuals having genome $\sigma^{n_0}$.
-Second, since the evolution is neutral and does not favour any genome in particular, the probability that $\sigma^{n_0}$ is equal to a given genome $\sigma$ is $1/N$.
+Second, since the evolution is neutral and does not favor any genome in particular, the probability that $\sigma^{n_0}$ is equal to a given genome $\sigma$ is $1/N$.
 In other words, the probability that a genome at $t=0$ ultimately becomes the ancestor of all the future population is equal to its frequency in the $t=0$ population.
@@
 \label{sub:predictor_based_on_the_local_lbi_maxima}
-	In figure~\ref{fig:emd_to_future}, we use several sequences as a predictor of the future population.
+	In figure~\sref{fig:emd_to_future}, we use several sequences as a predictor of the future population.
 Distance between two sets of sequences, \emph{i.e.} the predictor sequences and the ones of the future population, is defined as the Earth Mover's Distance (EMD).
@@
 The analysis of the main text computes probabilities of fixation assuming that all trajectories are independent.
-However, it is well-known that mutations in infleunza viruses are nested: they appear on backgrounds that already carry other mutations.
+However, it is well-known that mutations are nested: they appear on backgrounds that already carry other mutations.
 Since mutations appearing on the same genomes will jointly fix or disappear, many frequency trajectories are not independent but correlated.
@@
 In order to perform clustering, we define a distance between trajectories.
-A frequency trajectory $X$ is characterized by a series of frequency values and a time interval $T$.
-$f(t)$ for $t\in T$ corresponds to the frequency at which a given mutation $x$ appears in the population at date $t$, and we define $S(t)$ as the strains that carry mutation $x$ at date $t$.
+A frequency trajectory of a mutation $X$ is characterized by a series of frequency values $f(t)$ at time point  $t\in T$.
+We define $S(t)$ as the strains that carry mutation $X$ at date $t$.
 With this notation, $f(t)$ is the ratio of the number of elements in $S(t)$ to the total number of strains at date $t$.
@@
 
-In a companion article \citep{huddleston_integrating_2020}, authors developed several models to predict the future HA population for A/H3N2 influenza.
+\citet{huddleston_integrating_2020} developed several models to predict the future HA population for A/H3N2 influenza.
 The premise of these models is to assign a fitness score to the HA gene of each A/H3N2 strain in a given year, and to then apply deterministic evolutionary equations to obtain the future strain composition.
@@
 However, this effect vanishes for higher frequencies, with both fitter and less fit mutations having the same chance to fix.
+This is consistent with the idea that mutations that rose to very high frequency likely have a minor fitness costs even if classified as deleterious mutation.
 

### 2020-12-28 article.tex
Fix formula for trajectory distance

@@
 We define the distance $d(X_1,X_2)$ between these two trajectories based on the average similarity of the strains $S_1$ and $S_2$ that compose them:
-$$ d(X_1,X_2) = \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$
+$$ d(X_1,X_2) = 1 - \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$
 where $T_1\cap T_2$ is the time interval where both trajectories are active, and $\vert\cdot\vert$ denotes the number of elements of a set.

### 2020-12-28 article.tex
Misc changes

@@
 	This observation holds for many different categories of mutations, including mutations at epitope positions.
-	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
+	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retains clear signatures of predictability.
 	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
@@
 
-For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$.
+For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{C} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$.
 The same analysis was performed for A/H1N1pdm, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
@@
 To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$.
-The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C.
+The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}B.
 Initially, \emph{i.e.}, at time $t_0=0$, this distribution is by construction peaked around $f_0$.
@@
 However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps.
-The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories.
-This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
+The thick black line in Figure~\ref{fig:trajectories}C shows the average frequency of all trajectories.
+This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (figure \sref{fig:trajectories_h1n1}).
 Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
@@
 Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$?
-\textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point.
+\textbf{B}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{C} and for specific values of $\Delta t$.
+\textbf{C}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point.
 Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status.
@@
 0).
-Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.
-\textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$.
-}
+Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.}
 	\label{fig:trajectories}
@@
 
+Next, we cluster trajectories of simulated populations based on the intrinsic fitness effect of corresponding mutations and in the case of a fast changing fitness landscape ($\Delta t=10$). 
+Since competition between adaptive mutations and genetic linkage is very strong in this regime, we do not expect the intrinsic fitness effect to be fully predictive of fixation. 
+However, figure \sref{fig:simulations_fitness} shows that mutations with an above median fitness effect have a significantly larger chance to fix than those with a below median fitness effect. 
+This strongly contrasts with figure \ref{fig:pfix_stratified}, and confirms the intuition that intrinsic fitness effect can be partly predictive of fixation even in a regime of strong clonal interference.
+
 These models are not meant to be accurate models of influenza viruses evolution.
@@
 Panel \textbf{B} of figure~\ref{fig:LBI_consensus} shows these distances, scaled with respect to an average strain (details in caption).
-It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only once is the sequence of the top-LBI strain farther away from the consensus than the average sequence is.
+It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only twice is the sequence of the top-LBI strain farther away from the consensus than the average sequence is.
 Moreover, the sequence of the top-LBI strain \emph{exactly} matches the consensus in 19 cases.
@@
 
-A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
+A recent study proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
 Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
@@
 	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
-	Strain names and accession numbers are given as tables in two supplementary files. Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/src/config} were removed. \\
+	Strain names and accession numbers are given as tables in two supplementary files.
+Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed.
+\\
 	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}.

### 2020-12-28 response_to_referees.tex
Edits

@@
 The LBI was among the highest performing models and we hence used this score in the original manuscript.
-But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material ("Ability of fitness models to predict fixation").
+But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material (``Ability of fitness models to predict fixation'').
 These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
@@
 We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics and fixation, but they remain predictive of the composition of the future population.
-In consequence, we have toned down our conclusions about predictability.
-\pierre{I say that because I changed p8 l47.
-I have not touched p9 l48 because the new appendix shows it is a correct statement.}
+In consequence, we have toned down our conclusions about predictability in the discussion.}
 }

### 2020-12-29 response_to_referees.tex
Edits

@@
 }
-\responsedraft{Should we mention all the changes we made in the text?
-We spend more time explaining that even though there is linkage, and even though we're looking at just the last point, we expect something above diagonal.
-(Last paragraph before "Results", last paragraph before "Predicting future frequencies"), and first paragraph of "Simulations of models of adaptation"}
 % \responsedraft{
@@
 The LBI was among the highest performing models and we hence used this score in the original manuscript.
-But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material (``Ability of fitness models to predict fixation'').
+But we took the suggestion of the reviewer (and reviewer 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material (``Ability of fitness models to predict fixation'').
 These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
-We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.
+We believe this addition has made the manuscript stronger and we thank both reviewers for the suggestion.
 
@@
 In consequence, we have toned down our conclusions about predictability in the discussion.}
-}
 
@@
 Specifically, the very related work by Illingworth and Mustonen is not mentioned.
-The distribution $P_delta_t(f|f_0)$ has been at the core of
+The distribution $P_{\delta t}(f|f_0)$ has been at the core of
 the method of Strelkowa and Lassig, 2012 (which is cited, but at other
 parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
-\responsedraft{I cited (Strelkowa and Lassig) in the first paragraph of the section "Predicting future frequencies" and in the second of ``Simulations''
-We cite Illingworth and Mustonen plenty of times now.
-}
+\response{We now refer to the work by Illingworth and Mustonen in different parts of the text.
+We also added citations to Strelkowa and L\"assig when discussing the future distribution of frequencies and the simulations.}
 
@@
 We now highlight the differences in the text and discuss potential reasons for the different behaviour.  }
+\responsedraft{We don't discuss the potential reasons yet!}
 
@@
 \response{
-	Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
-	Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
+	% Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
+	% Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
+	Methods that are useful for designing vaccines are those that predict which circulating strains will dominate future influenza populations.
+	Such predictions have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
+
+	On the other hand, our model-free approach is more concerned in understanding which aspects of influenza evolution are predictable and which ones are not.
+	The main implications of our results are about the evolutionary dynamics of influenza, and only indirectly about vaccines.  
 }
-\responsedraft{We should maybe make clear that this is not directly related to predictions for vaccines (at least it's not the aim).
-That's John's paper.
-The main implications are about our understanding of the evolutionary dynamics of influenza.
-This could also help us to reply to the first comment: doing ML is more useful to predict the future population (for vaccines) than for predicting fixation of a single mutation. }
 

### 2021-01-05 article.tex
discuss h1n1 h3n2 differences

@@
 Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
-Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population.
+Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population \citep{lee_mapping_2019}.
+Immunity to A/H1N1pdm might be less variable among humans than to A/H3N2 since A/H1N1pdm has been circulating for only 11 years.
+Different age-groups therefore got exposed to more similar A/H1N1pdm variants than their initial exposure to A/H3N2.
 These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

### 2021-01-10 article.tex
Change top/bottom -> left/right in SI

--- 2020-12-28 article.tex
@@
 		\caption{Fixation probability as a function of frequency for the simulations discussed in the main text.
-\textbf{Top:} Simulation without antibodies.
+\textbf{Left:} Simulation without antibodies.
 The three colored curves reflect different rate of change for the fitness landscape.
 Visual inspection of the frequency trajectories indicates a typical sweep time of $\sim400$ generations.
-\textbf{Bottom:} Simulation with antibodies.
+\textbf{Right:} Simulation with antibodies.
 The different colored curves indicate the rate at which antibodies are introduced.
@@
 ``High'' and ``low'' fitness classes are defined with respect to the median value.
-\textbf{Top:} Simulation without antibodies and with changes to the fitness landscape every $dt=10$ generations.
-\textbf{Bottom:} Simulation with antibodies, with a new antibody every $dt=10$ generations.
+\textbf{Left:} Simulation without antibodies and with changes to the fitness landscape every $dt=10$ generations.
+\textbf{Right:} Simulation with antibodies, with a new antibody every $dt=10$ generations.
 }

### 2021-01-13 article.tex

@@
 The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments.
-Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to the mutation.
+Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to novel mutation in this category.
 
@@
 Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
-In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
+In the extreme case were the two mutations always appear in the same strains, their two trajectories will be identical.
 This ``nesting'' of frequency trajectories may introduce biases when performing statistical analysis.
@@
 0).
-Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.}
+Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f_0$.}
 	\label{fig:trajectories}
@@
 \subsection*{Prediction of fixation or loss}
-Instead of predicting future frequency, let's consider the long-term goal of predicting the probability that a mutation fixes in the population.
+Instead of predicting future frequency, we then considered the probability that a mutation fixes in the population.
 We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen.
@@
 For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time.
-The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
+The fixation probability of a mutation at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
 Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.
@@
 
-This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
-If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are observed.
-In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
+%This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
+%If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are observed.
+%In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.
 
@@
 Only mutations at the 7 pivotal sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance.
-No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
+No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} fix less often.
 One should also note that many of these positions were determined post-hoc and might be enriched for positions that experienced rapid substitutions before the publication of the respective studies.
@@
 	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency, for A/H3N2 influenza.
-Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1.
+Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1pdm.
 \textbf{A}: HA mutations with higher or lower LBI values, based on their position with respect to the median LBI value.
@@
 Picking representatives of the future and predicting frequency dynamics are distinct objectives and success at the former (as compared to random picks) is not necessarily inconsistent with a lack of predictable dynamics.
-In fact, \citep{huddleston_integrating_2020} reports that the rate at which the frequency of a strain changes is often a poor predictor -- consistent with our observations here.
+In fact, \citet{huddleston_integrating_2020} reports that the rate at which the frequency of a strain changes is often a poor predictor -- consistent with our observations here.
 But despite the fact that future frequencies are not predicted by the LBI, the strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one.
@@
 	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
-	Strain names and accession numbers are given as tables in two supplementary files.
-Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed.
+	Acknowledgement tables with strain names, accession numbers, and originating and submitting labs are given as two supplementary files.
+Outlier strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed.
 \\
@@
 	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
-	Region used are the ones defined in the \texttt{Nextstrain} tool~\cite{10.1093/bioinformatics/bty407}.
+	Region used are the ones defined in \texttt{Nextstrain/flu}~\cite{10.1093/bioinformatics/bty407}.
 Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
@@
 ``High'' and ``low'' fitness classes are defined with respect to the median value.
-\textbf{Left:} Simulation without antibodies and with changes to the fitness landscape every $dt=10$ generations.
-\textbf{Right:} Simulation with antibodies, with a new antibody every $dt=10$ generations.
+\textbf{Left:} Simulation with a purely additive landscape that changes every $\Delta t=10$ generations.
+\textbf{Right:} Simulation with antibodies, with a new antibody every $\Delta t=10$ generations.
 }

### 2021-01-13 article.tex
Higher dpi figures

@@
 The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments.
-Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to novel mutation in this category.
+Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to novel mutations in this category.
 
